---
title: "Get species data for SDMs"
author: "Brunno F Oliveira & Bioshifts group"
date: "Last compiled on `r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
        code_folding: hide
---

# *Setup*

```{r setup, message = FALSE, warning = FALSE, include = FALSE}

rm(list=ls())
gc()

list.of.packages <- c("dplyr", "here", "tidyr","data.table", "ggplot2", "pbapply","ggvenn")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

if(length(new.packages)) install.packages(new.packages)

sapply(list.of.packages, require, character.only = TRUE)

```

# *Load species list*

```{r}
computer = "personal"

source(here("R/settings.R"))
source(here("R/my_functions.R"))

N_OCC <- read.csv(here("Data/N_OCC.csv"))
splist <- read.csv(here("Data/Bioshifts/splist_v3.csv"), header = T)
bioshifts <- read.csv(here(Bioshifts_dir,Bioshifts_DB_v3))
bioshifts <- bioshifts_fix_columns(bioshifts)

```


# *N species*
N species in bioshifts
```{r}
length(unique(bioshifts$sp_name_std))
```


N species workable for SDMs:
1) Latitude shifts
2) Range shifts within the time-period that matches the environmental data
3) Species in the GBIF taxonomy backbone
4) Species with > 30 occurrence records in GBIF

Let's go:
1) N species with latitude shifts

```{r}
bioshifts_lat <- bioshifts %>%
  filter(Type == "LAT") 

length(unique(bioshifts_lat$sp_name_std))
```
2) N species with range shifts within the time-period that matches the environmental data
```{r}
bioshifts_time <- bioshifts %>%
  filter(Eco == "Ter" & (Start >= (temporal_range_env_data("Ter")[1] + n_yr_bioclimatic)) | # Shifts Marine or Terrestrial + within time period of the environmental data
           (Eco == "Mar" & (Start >= (temporal_range_env_data("Mar")[1] + n_yr_bioclimatic))))

length(unique(bioshifts_time$sp_name_std))
```
3) N species in the GBIF taxonomy backbone
```{r}
bioshifts_gbif <- N_OCC 

length(unique(bioshifts_gbif$scientificName))

```
4) Species with > 30 occurrence records in GBIF
```{r}
bioshifts_gbif <- N_OCC %>%
  filter(N_OCC > 30)

length(unique(bioshifts_gbif$scientificName))

```

Species in GBIF By Eco and group

```{r}
## Terrestrials
N_OCC %>%
  filter(Eco == "Ter") %>%
  group_by(class) %>%
  tally() %>%
  arrange(by = desc(n))

## Marine
N_OCC %>%
  filter(Eco == "Mar") %>%
  group_by(class) %>%
  tally() %>%
  arrange(by = desc(n))

table(N_OCC$Eco)

####################
#### Plot
my_levels <- names(sort(table(N_OCC$class)))
ggplot(N_OCC %>%
         mutate(class = factor(class, levels = my_levels)), 
       aes(y = class))+
  geom_bar()+
  theme_classic()+
  # coord_flip()+
  xlab("N species")+ylab("")+
  # theme(axis.text.x = element_text(angle = 45, hjust=1))+
  facet_wrap(.~Eco, scales = "free")

```

Per range position
```{r fig.height=5, fig.width=7}

ggplot(bioshifts_lat %>%
         filter(sp_name_std %in% bioshifts_gbif$scientificName,
                Eco == "Ter") %>%
         group_by(class, Param) %>%
         summarise(N=length(unique(sp_name_std))), 
       aes(x = N, y = class))+
  ggtitle("Terrestrial")+
  geom_bar(stat = "identity")+
  theme_classic()+
  xlab("N species")+ylab("")+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  facet_grid(.~Param, scales = "free_y")

ggplot(bioshifts_lat %>%
         filter(sp_name_std %in% bioshifts_gbif$scientificName,
                Eco == "Mar") %>%
         group_by(class, Param) %>%
         summarise(N=length(unique(sp_name_std))), 
       aes(x = N, y = class))+
  ggtitle("Marine")+
  geom_bar(stat = "identity")+
  theme_classic()+
  xlab("N species")+ylab("")+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  facet_grid(.~Param, scales = "free_y")

```

# *N shifts*
N shifts in bioshifts
```{r}
nrow(bioshifts)
```
N shifts in bioshifts latitude
```{r}
nrow(bioshifts_lat)
```

N shifts latitude in GBIF
```{r}
nrow(bioshifts_lat %>%
       filter(sp_name_std %in% bioshifts_gbif$scientificName))
```

```{r}
my_levels <- names(sort(table(bioshifts_gbif$class)))

ggplot(bioshifts_gbif %>%
         mutate(class = factor(class, levels = my_levels)) %>%
         filter(Eco == "Ter" | Eco == "Mar"), 
       aes(y = class))+
  geom_bar()+
  theme_classic()+
  # coord_flip()+
  xlab("N shifts")+ylab("")+
  # theme(axis.text.x = element_text(angle = 45, hjust=1))+
  facet_wrap(.~Eco, scales = "free")

```

Per range position
```{r fig.height=5, fig.width=7}

ggplot(bioshifts_lat %>%
         filter(sp_name_std %in% bioshifts_gbif$scientificName,
                Eco == "Ter"), 
       aes(y = class))+
  ggtitle("Terrestrial")+
  geom_bar()+
  theme_classic()+
  # coord_flip()+
  xlab("N shifts")+ylab("")+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  facet_grid(.~Param, scales = "free_y")

ggplot(bioshifts_lat %>%
         filter(sp_name_std %in% bioshifts_gbif$scientificName,
                Eco == "Mar"), 
       aes(y = class))+
  ggtitle("Marine")+
  geom_bar()+
  theme_classic()+
  # coord_flip()+
  xlab("N shifts")+ylab("")+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  facet_grid(.~Param, scales = "free_y")

```

# *Completeness GBIF in Bioshifts*

```{r fig.height=5, fig.width=7}

x <- list(Temp_Range = unique(bioshifts_time$sp_name_std),
          GBIF = unique(bioshifts_gbif$scientificName),
          Latitude = unique(bioshifts_lat$sp_name_std),
          All_spp = unique(bioshifts$sp_name_std))

names(x) = c("Temperature\nrange", "GBIF","Latitude","Bioshifts")

ggvenn(
  x,
  # fill_color = c("#0073C2FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 4
)

```

This includes elevation shifts. If we considering only latitudinal shifts...

```{r fig.height=5, fig.width=5}
x <- list(Temp_Range = unique(bioshifts_time$sp_name_std),
          GBIF = unique(bioshifts_gbif$scientificName),
          Latitude = unique(bioshifts_lat$sp_name_std))

names(x) = c("Temperature range", "GBIF","Latitude")

ggvenn(
  x,
  # fill_color = c("#0073C2FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 4
)
```

## Marine

```{r fig.height=5, fig.width=5}

bioshifts_time_mar <- bioshifts_time %>%
  filter(Eco == "Mar") 

bioshifts_gbif_mar <- bioshifts_gbif %>%
  filter(Eco == "Mar") 

bioshifts_lat_mar <- bioshifts_lat %>%
  filter(Eco == "Mar")

x <- list(Temp_Range = unique(bioshifts_time_mar$sp_name_std),
          GBIF = unique(bioshifts_gbif_mar$scientificName),
          Latitude = unique(bioshifts_lat_mar$sp_name_std))

names(x) = c("Temperature range", "GBIF","Latitude")

ggvenn(
  x,
  # fill_color = c("#0073C2FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 4
)

```

## Terrestrial

```{r fig.height=5, fig.width=5}

bioshifts_time_ter <- bioshifts_time %>%
  filter(Eco == "Ter") 

bioshifts_gbif_ter <- bioshifts_gbif %>%
  filter(Eco == "Ter")

bioshifts_lat_ter <- bioshifts_lat %>%
  filter(Eco == "Ter")

x <- list(Temp_Range = unique(bioshifts_time_ter$sp_name_std),
          GBIF = unique(bioshifts_gbif_ter$scientificName),
          Latitude = unique(bioshifts_lat_ter$sp_name_std))

names(x) = c("Temperature range", "GBIF","Latitude")

ggvenn(
  x,
  # fill_color = c("#0073C2FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 4
)

```

```{r fig.height=10, fig.width=4}
# Freq tables
# bioshifts_spp
bioshifts_spp <- bioshifts %>%
  group_by(class) %>%
  summarise(Bioshifts_Freq = length(unique(sp_name_std)))

# bioshifts_gbif_spp
bioshifts_gbif_spp <- bioshifts_gbif %>%
  group_by(class) %>%
  summarise(GBIF_Freq = length(unique(scientificName)))

# lat data
bioshifts_lat_spp <- bioshifts_lat %>%
  group_by(class) %>%
  summarise(Lat_Freq = length(unique(sp_name_std)))

# temporal data
bioshifts_time_spp <- bioshifts_time %>%
  group_by(class) %>%
  summarise(Time_Freq = length(unique(sp_name_std)))

bio <- merge(bioshifts_spp, bioshifts_gbif_spp, all.x = TRUE)
bio <- merge(bio, bioshifts_lat_spp, all.x = TRUE)
bio <- merge(bio, bioshifts_time_spp, all.x = TRUE)
bio[is.na(bio)] <- 0

bio$GBIF_perc = paste(round((bio$GBIF_Freq/bio$Bioshifts_Freq)*100,2),"%")
# bio$Lat_perc = paste(round((bio$Lat_Freq/bio$Bioshifts_Freq)*100,2),"%")
# bio$Temp_perc = paste(round((bio$Time_Freq/bio$Bioshifts_Freq)*100,2),"%")

bio_freq_melted <- reshape::melt(bio[,c(1:5)], id="class")
bio_freq_melted$variable <- gsub("_Freq","",bio_freq_melted$variable)
bio_freq_melted$Freq <- bio_freq_melted$value

bio_perc_melted <- reshape::melt(bio[,c(1,6)], id="class")
bio_perc_melted$variable <- gsub("_perc","",bio_perc_melted$variable)
bio_perc_melted$Perc <- bio_perc_melted$value

bio_melted <- merge(bio_freq_melted[,-3], bio_perc_melted[,-3],
                    by=c("class","variable"),
                    all.x = TRUE)

# 
# melted <- merge(melted, bio[,c(1,4)])
# melted$percent[which(melted$variable == "GBIF_Freq")] = NA
# melted
# keep <- unique(melted$class[which(melted$value > 20)])
# toplot = melted[which(melted$class %in% keep),]
# levels_plot <- toplot %>%
#     filter(!is.na(percent))
```


```{r fig.height=10, fig.width=7}

bio_melted$class <- factor(bio_melted$class, levels = unique(bio_melted$class[order(bio_melted$Freq)]))

ggplot(bio_melted, 
       aes(x = class, y = Freq, fill = variable))+
  geom_bar(stat="identity", position = "identity", alpha = .5) +
  geom_text(aes(label=Perc),
            stat='identity', size = 3, hjust = "inward")+
  coord_flip()+
  ylab("N species")+
  theme_classic()+
  theme(legend.position="bottom",
        legend.title = element_blank())

```

Filter classes with > 10 species

```{r}
# > 10 species
my_sel <- bio_melted %>%
  filter(variable == "Bioshifts" & Freq > 10)
my_sel <- bio_melted %>%
  filter(class %in% my_sel$class)
nice_table(my_sel)

ggplot(my_sel, aes(x = class, y = Freq, fill = variable))+
  geom_bar(stat="identity", position = "identity", alpha = .5) +
  geom_text(aes(label=Perc),
            stat='identity', size = 3, hjust = "inward")+
  coord_flip()+
  ylab("N species")+
  theme_classic()+
  theme(legend.position="bottom",
        legend.title = element_blank())

```

Filter latitude shifts only

```{r}

bio <- merge(bioshifts_gbif_spp, bioshifts_lat_spp, all.x = TRUE)
bio <- merge(bio, bioshifts_time_spp, all.x = TRUE)
bio[is.na(bio)] <- 0

bio$GBIF_perc = paste(round((bio$GBIF_Freq/bio$Lat_Freq)*100,2),"%")
# bio$Lat_perc = paste(round((bio$Lat_Freq/bio$Bioshifts_Freq)*100,2),"%")
# bio$Temp_perc = paste(round((bio$Time_Freq/bio$Bioshifts_Freq)*100,2),"%")

bio_freq_melted <- reshape::melt(bio[,c(1:4)], id="class")
bio_freq_melted$variable <- gsub("_Freq","",bio_freq_melted$variable)
bio_freq_melted$Freq <- bio_freq_melted$value

bio_perc_melted <- reshape::melt(bio[,c(1,5)], id="class")
bio_perc_melted$variable <- gsub("_perc","",bio_perc_melted$variable)
bio_perc_melted$Perc <- bio_perc_melted$value

bio_melted <- merge(bio_freq_melted[,-3], bio_perc_melted[,-3],
                    by=c("class","variable"),
                    all.x = TRUE)
bio_melted$class <- factor(bio_melted$class, levels = unique(bio_melted$class[order(bio_melted$Freq)]))

# > 10 species
my_sel <- bio_melted %>%
  filter(variable == "Lat" & Freq > 10)
my_sel <- bio_melted %>%
  filter(class %in% my_sel$class)
nice_table(my_sel)

ggplot(my_sel, aes(x = class, y = Freq, fill = variable))+
  geom_bar(stat="identity", position = "identity", alpha = .5) +
  geom_text(aes(label=Perc),
            stat='identity', size = 3, hjust = "inward")+
  coord_flip()+
  ylab("N species")+
  theme_classic()+
  theme(legend.position="bottom",
        legend.title = element_blank())
```

# *Session info*
```{r}
sessionInfo()
```

