---
title: "Exploratory analyses"
author: "Bioshifts"
format: 
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-summary: "Code"
editor: source
---

# *Setup*

```{r warning=FALSE, message=FALSE}

rm(list=ls())
gc()

library(dplyr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(knitr)
library(pbapply)
library(data.table)
library(GGally)
library(here)
library(terra)
library(tidyterra)
library(rcompanion)
library(lme4)
library(cowplot)
library(circular)
library(caret)
library(MuMIn)
library(ggh4x)
# library(plotrix)

computer = "personal"
source(here("R/custom_legend.R"))
source(here("R/settings.R"))
source(here("R/my_functions.R"))
source(here("R/velocity_functions.R"))

```

```{r warning=FALSE, message=FALSE}
myramp1 <- colorRampPalette(colors = c("red","gray","blue","gray","red"))(360)

the_palette_fc <- leaflet::colorNumeric(
    palette = myramp1, 
    domain = c(0,360),
    reverse = FALSE)

myramp_angle <- the_palette_fc(seq(0, 360, length.out = 50))

```

# *Load data*

```{r}

bioshifts <- read.csv(here("Data/Output/Bioshifts_merge_Exposure_all.csv"))

```

```{r warning=FALSE, message=FALSE}

# fixing Locomotion_mode
bioshifts$Locomotion_mode = as.character(str_to_sentence(bioshifts$Locomotion_mode))
unique(bioshifts$Locomotion_mode)
# if multiple locomotion modes, use the first (the main locomotion mode)
# bioshifts$Locomotion_mode <- sapply(bioshifts$Locomotion_mode, function(x) {
#   strsplit(x, split = "/", fixed = TRUE)[[1]][1] 
# })

pos <- grep("Flying",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Fly"
pos <- grep("wimming",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Swim"
pos <- grep("alk",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Run/Walk"
pos <- grep("unning",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Run/Walk"
pos <- grep("lithering",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Slither/Crawl"
pos <- grep("rawling",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Slither/Crawl"
pos <- grep("Burrowing",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Sessile"

unique(bioshifts$Locomotion_mode)

# bioshifts$Locomotion_mode <- factor(
#   bioshifts$Locomotion_mode,
#   levels = c("Flying", "Running", "Swimming", "Walking", "Slithering", "Crawling",  "Planktonic", "Sessile", "Burrowing"))
# 
# # group Slithering/Crawling and Running/Walking
# pos <- grep("Slithering",bioshifts$Locomotion_mode)
# bioshifts$Locomotion_mode[pos] <- "Slithering/Crawling"
# pos <- grep("Crawling",bioshifts$Locomotion_mode)
# bioshifts$Locomotion_mode[pos] <- "Slithering/Crawling"
# 
# pos <- grep("Running",bioshifts$Locomotion_mode)
# bioshifts$Locomotion_mode[pos] <- "Running/Walking"
# pos <- grep("Walking",bioshifts$Locomotion_mode)
# bioshifts$Locomotion_mode[pos] <- "Running/Walking"
# 
bioshifts$Locomotion_mode <- factor(
    bioshifts$Locomotion_mode,
    levels = c("Fly", "Run/Walk", "Swim", "Slither/Crawl", "Planktonic", "Sessile"))

# fixing Category
table(bioshifts$Category)
# remove synthesis because there is a only a single study
bioshifts <- bioshifts |> filter(!Category == "Synthesis")
# combine census-resurvey and survey-resurvey 
pos <- grep("Census-Resurvey",bioshifts$Category)
bioshifts$Category[pos] <- "Survey-Resurvey"
table(bioshifts$Category)
bioshifts$Category <- factor(bioshifts$Category, 
                             levels = c("TimeSeries", "CensusPeriods", "Survey-Resurvey"))

# make numeric as.numeric
pos <- which(sapply(bioshifts, is.numeric))
tmp <- lapply(bioshifts[pos], as.numeric)
tmp <- do.call(cbind,tmp)
bioshifts[pos] <- tmp

# remove one outlier species
bioshifts <- bioshifts |>
    filter(!(Eco == "Ter" & bvel_lat < -15))


# remove swimming terrestrials
bioshifts <- bioshifts |>
    filter(!(Eco == "Ter" & Locomotion_mode == "Swim"))

# select classes with more than X observations per edge
N_obs_class = 5
class_param_select <- bioshifts |>
    group_by(class, Param) |>
    tally() |>
    filter(n >= N_obs_class) |>
    mutate(class_param = paste(class,Param))

bioshifts <- bioshifts |> 
    mutate(class_param = paste(class,Param)) |> 
    filter(class_param %in% class_param_select$class_param) |>
    select(!class_param)


```

```{r}

# N species
length(unique(bioshifts$sp_name_std))

bioshifts |> 
    group_by(Param) |> 
    summarise(N=length(unique(sp_name_std)))

bioshifts |> 
    group_by(Eco) |> 
    summarise(N=length(unique(sp_name_std)))

# N range shifts
length(bioshifts$sp_name_std)

bioshifts |> 
    group_by(Param) |> 
    summarise(N=length(sp_name_std))

bioshifts |> 
    group_by(Eco) |>
    summarise(N=length(sp_name_std))

```

Colors for lifeforms

```{r}

ramps <- list(Endotherm = c("red4", "pink"),
              Ectotherm = c("deepskyblue", "lightblue"),
              Cryptogam = c("darkgreen", "lightgreen"),
              Phanerogam = c("darkorange", "yellow"))

bio_cols <- bioshifts |>
    group_by(lifeform, class) |>
    tally() |>
    select(lifeform, class)
bio_cols$hex <- NA

for(i in 1:length(ramps)){
    pos <- which(bio_cols$lifeform == names(ramps)[i])
    ramp_i <- ramps[[i]]
    cols_i <- bio_cols[pos,]
    bio_cols$hex[pos] <- colorRampPalette(ramp_i)(length(unique(cols_i$class)))
}

# add colors
bioshifts <- merge(bioshifts, bio_cols[,2:3], by="class")

GP_legend <- custom_legend(mdf = bioshifts |>
                               mutate(Top_lifeform = lifeform,
                                      Top_class = class), 
                           cdf = bio_cols |>
                               mutate(Top_lifeform = lifeform,
                                      Top_class = class), 
                           group_level = "lifeform", 
                           subgroup_level = "class", 
                           x = "bvel_lat", y = "ShiftRate",
                           legend_orientation = "horizontal")

```

# *SDM evaluation results*

```{r fig.width=8, fig.height=4*2}

N_OCC <- read_csv("Data/sdms_occ.csv")
N_OCC$sp_name_std <- gsub(" ","_",N_OCC$Species)

sdms_CV <- bioshifts |>
    select("sensitivity", "calibration","validation","Boyce",
           "Eco","sp_name_std","class") |>
    group_by(sp_name_std,Eco,class) |>
    summarise(sensitivity = mean(sensitivity,na.rm=TRUE),
              calibration = mean(calibration,na.rm=TRUE),
              validation = mean(validation,na.rm=TRUE),
              Boyce = mean(Boyce,na.rm=TRUE)) |>
    mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                           Eco == "Ter" ~ "Terrestrial"))

sdms_CV <- merge(sdms_CV, N_OCC[,c("sp_name_std","N_occ")], 
                 by = "sp_name_std",
                 all.x = TRUE)

```


```{r}

summary(sdms_CV$N_occ)

# how many species with < 100 occurrences?
length(which(sdms_CV$N_occ < 100))
# in %
length(which(sdms_CV$N_occ < 100))/nrow(sdms_CV)

plot(log(sdms_CV$N_occ),
     sdms_CV$Boyce)

plot(log(sdms_CV$N_occ),
     sdms_CV$calibration)

plot(log(sdms_CV$N_occ),
     sdms_CV$validation)
```


```{r fig.width=8, fig.height=4*2}
N_sps_class <- sdms_CV |>
    group_by(class) |>
    summarise(N = length(class))

sdms_CV <- merge(sdms_CV, N_sps_class)
sdms_CV$class <- paste0(sdms_CV$class, " (",sdms_CV$N,")")

sdms_CV_mean <- sdms_CV |>
    group_by(Eco,class) |>
    summarise(sensitivity = mean(sensitivity,na.rm=TRUE),
              calibration = mean(calibration,na.rm=TRUE),
              validation = mean(validation,na.rm=TRUE),
              Boyce = mean(Boyce,na.rm=TRUE))
```

```{r fig.width=8, fig.height=4*2}
p1 <-
    ggplot(sdms_CV, aes(x = validation, 
                        y = factor(class, level = rev(sort(unique(sdms_CV$class))))))+
    geom_jitter(alpha=.3, height = .1, color = "gray30")+
    geom_boxplot(outlier.alpha = 0, alpha = .5)+
    geom_point(data = sdms_CV_mean,
               aes(x = validation, 
                   y = factor(class, level = rev(sort(unique(sdms_CV$class))))),
               size = 4,shape = 21, 
               fill = "white", color = "black", alpha = .5)+
    facet_wrap(.~Eco, scales = "free")+
    labs(x="TSS validation",y="")+
    scale_x_continuous(limits=c(0,1))+
    # intercept 0
    geom_vline(xintercept = 0, color='gray', linetype = "dashed")+
    # grand mean
    geom_vline(xintercept = mean(sdms_CV$validation,na.rm=TRUE),
               color='black', linewidth = 1, alpha = .5)+
    # grand mean terrestrial
    geom_vline(data=filter(sdms_CV, Eco=="Terrestrial"),
               aes(xintercept = mean(validation, na.rm = TRUE)),
               color='red', linewidth = 1, alpha = .5)+
    # grand mean marine
    geom_vline(data=filter(sdms_CV, Eco=="Marine"),
               aes(xintercept = mean(validation, na.rm = TRUE)),
               color='red', linewidth = 1, alpha = .5)+
    theme_classic()

p2 <- ggplot(sdms_CV, aes(x = Boyce, 
                          y = factor(class, level = rev(sort(unique(sdms_CV$class))))))+
    geom_jitter(alpha=.3, height = .1, color = "gray30")+
    geom_boxplot(outlier.alpha = 0,
                 alpha = .5)+
    geom_point(data = sdms_CV_mean,
               aes(x = Boyce, 
                   y = factor(class, level = rev(sort(unique(sdms_CV$class))))),
               size = 4,shape = 21, 
               fill = "white", color = "black", alpha = .5)+
    facet_wrap(.~Eco, scales = "free")+
    labs(x="Boyce",y="")+
    # intercept 0
    geom_vline(xintercept = 0, color='gray', linetype = "dashed")+
    # grand mean
    geom_vline(xintercept = mean(sdms_CV$Boyce,na.rm=TRUE),
               color='black', linewidth = 1, alpha = .5)+
    # grand mean terrestrial
    geom_vline(data=filter(sdms_CV, Eco=="Terrestrial"),
               aes(xintercept = mean(Boyce, na.rm = TRUE)),
               color='red', linewidth = 1, alpha = .5)+
    # grand mean marine
    geom_vline(data=filter(sdms_CV, Eco=="Marine"),
               aes(xintercept = mean(Boyce, na.rm = TRUE)),
               color='red', linewidth = 1, alpha = .5)+
    theme_classic()
```

```{r fig.width=8, fig.height=4*2}

cowplot::plot_grid(p1,p2,
                   labels = "AUTO",
                   ncol = 1)

ggsave(filename = here("Figures/SDM_evaluation_all.png"), 
       device = "png", units = "in", dpi = 300,
       width = 8, height = 8)


```

## TSS calibration

```{r fig.width=8, fig.height=4*3}
sdms_CV |>
    ungroup() |>
    summarise(Mean=mean(calibration,na.rm=TRUE),
              SD=sd(calibration,na.rm=TRUE),
              Range=paste(range(calibration,na.rm=TRUE),collapse = " - "))

sdms_CV |>
    group_by(Eco) |>
    summarise(Mean=mean(calibration,na.rm=TRUE),
              SD=sd(calibration,na.rm=TRUE),
              Range=paste(range(calibration,na.rm=TRUE),collapse = " - "))

maov <- aov(calibration~Eco, sdms_CV)
summary(maov)
TukeyHSD(maov)

maov <- aov(calibration~class, sdms_CV)
summary(maov)
TukeyHSD(maov)
```

## TSS validation

```{r fig.width=8, fig.height=4*3}
sdms_CV |>
    ungroup() |>
    summarise(Mean=mean(validation,na.rm=TRUE),
              SD=sd(validation,na.rm=TRUE),
              Range=paste(range(round(validation,2),na.rm=TRUE),collapse = " - "))

sdms_CV |>
    group_by(Eco) |>
    summarise(Mean=mean(validation,na.rm=TRUE),
              SD=sd(validation,na.rm=TRUE),
              Range=paste(range(validation,na.rm=TRUE),collapse = " - "))

maov <- aov(validation~Eco, sdms_CV)
summary(maov)
TukeyHSD(maov)

# top highest values
sdms_CV |>
    group_by(class,Eco) |>
    summarise(Mean=mean(validation,na.rm=TRUE),
              N = length(class)) |>
    arrange(desc(Mean))

```

## Sensitivity

```{r fig.width=8, fig.height=4*3}
sdms_CV |>
    ungroup() |>
    summarise(Mean=mean(sensitivity,na.rm=TRUE),
              SD=sd(sensitivity,na.rm=TRUE),
              Range=paste(range(sensitivity,na.rm=TRUE),collapse = " - "))

sdms_CV |>
    group_by(Eco) |>
    summarise(Mean=mean(sensitivity,na.rm=TRUE),
              SD=sd(sensitivity,na.rm=TRUE),
              Range=paste(range(sensitivity,na.rm=TRUE),collapse = " - "))

maov <- aov(sensitivity~Eco, sdms_CV)
summary(maov)
TukeyHSD(maov)
```

## Boyce

```{r fig.width=8, fig.height=4*3}
sdms_CV |>
    ungroup() |>
    summarise(Mean=mean(Boyce,na.rm=TRUE),
              SD=sd(Boyce,na.rm=TRUE),
              Range=paste(range(Boyce,na.rm=TRUE),collapse = " - "))

sdms_CV |>
    group_by(Eco) |>
    summarise(Mean=mean(Boyce,na.rm=TRUE),
              SD=sd(Boyce,na.rm=TRUE),
              Range=paste(range(Boyce,na.rm=TRUE),collapse = " - "))


maov <- aov(Boyce~Eco, sdms_CV)
summary(maov)
TukeyHSD(maov)

sdms_CV |>
    group_by(class) |>
    summarise(Mean=mean(Boyce,na.rm=TRUE),
              SD=sd(Boyce,na.rm=TRUE),
              Range=paste(range(Boyce,na.rm=TRUE),collapse = " - "))

maov <- aov(Boyce~class, sdms_CV)
summary(maov)
TukeyHSD(maov)

# top highest values
sdms_CV |>
    group_by(class,Eco) |>
    summarise(Mean=mean(Boyce,na.rm=TRUE),
              N = length(class)) |>
    arrange(desc(Mean))

bioshifts |>
    filter(class == "Elasmobranchii") |>
    select(class,family) |>
    group_by(class,family) |>
    tally()
```

## Correlation between boyce and TSS

```{r}

plot(validation~Boyce,data=sdms_CV)

cor.test(sdms_CV$Boyce, sdms_CV$validation)

```

# *Explore relationships*

## *N shifts per parameter*

```{r}

ggplot(bioshifts |> 
           group_by(Param) |>
           tally(), aes(x = n, y = Param))+
    geom_col()+
    theme_classic()+
    geom_text(aes(label = n), vjust = -0.5)+
    labs(x="N shifts",y="N obs")

ggplot(bioshifts |> 
           group_by(Param, Eco) |>
           tally(), aes(x = n, y = Param))+
    geom_col()+
    theme_classic()+
    geom_text(aes(label = n), vjust = -0.5)+
    labs(x="N shifts",y="N obs")+
    facet_wrap(Eco~., scales = "free")

bioshifts |> 
    group_by(Param, class) |>
    tally()

```

# *How related are observed shifts and climate exposure metrics?*

```{r, fig.width=10, fig.height=8,message=FALSE,warning=FALSE}

mydata <- bioshifts |>
    mutate(`Bioclimatic vel.` = bvel_lat,
           `Biotic vel.` = ShiftRate) |>
    select(Eco, Param, `Bioclimatic vel.`, `Biotic vel.`) |>
    pivot_longer(cols = -c(Eco, Param), names_to = "Metric", values_to = "Velocity") |>
    na.omit()

mydata$Metric <- factor(mydata$Metric, levels = rev(c("Biotic vel.",  "Bioclimatic vel.")))

plot_cat <- expand.grid(Param=c("LE","O","TE"),
                        realms=c("Ter","Mar"))

plotlist <- lapply(1:nrow(plot_cat), function(i){
    toplot <- mydata|>
        filter(Eco == plot_cat$realms[i],
               Param == plot_cat$Param[i])
    ggplot(toplot, 
           aes(x = Velocity, fill = Metric, color = Metric))+
        ggtitle(paste(plot_cat$realm[i],plot_cat$Param[i],sep = " - "))+
        geom_histogram(alpha = .5)+
        theme_classic()+
        # scale_x_continuous(limits = quantile(toplot$Velocity, c(0.1, 0.9), na.rm = TRUE))+
        theme(aspect.ratio=1,
              plot.title = element_text(hjust = 0.5))
})

ggpubr::ggarrange(
    plotlist = plotlist, 
    align = "v", 
    ncol=3, nrow=2,common.legend = TRUE)



```

```{r}

ggplot(bioshifts, 
       aes(x = log(signoise), fill = Param, color = Param))+
    geom_histogram(alpha = .5)+
    theme_classic()+
    geom_hline(yintercept = 0)+
    # scale_x_continuous(limits = quantile(toplot$Velocity, c(0.1, 0.9), na.rm = TRUE))+
    theme(aspect.ratio=1,
          plot.title = element_text(hjust = 0.5))


quantile(bioshifts$signoise, na.rm = TRUE)

```

# *Modeled latitudinal shift*

## Angle

### Frequency

#### Eco

```{r}

# Define the breaks
breaks <- seq(0, 360, length.out = 25)

# Calculate bin width
bin_width <- diff(breaks)[1]

# Bin the data and calculate midpoints and counts
hist_data <- bioshifts |>
    group_by(Eco) |>
    mutate(bin = cut(bvel_angle, breaks = breaks, include.lowest = TRUE, right = FALSE, labels = FALSE)) |>
    group_by(bin,Eco) |>
    summarise(counts = n()) |>
    mutate(midpoints = (breaks[bin] + breaks[bin + 1]) / 2)

# Select and arrange the relevant columns
hist_data <- hist_data |>
    select(midpoints, counts, Eco) |>
    arrange(midpoints)

ggplot(hist_data, aes(x = midpoints, y = counts))+
    ggtitle("Frequence of bioclimatic direction")+
    geom_bar(stat = "identity", aes(fill=Eco)) +
    scale_x_continuous(breaks = seq(from=0, to=359, by=45), limits=c(0, 360)) +
    coord_polar() +
    theme_minimal()+
    labs(x="",y="")+
    theme(legend.position = "none",
          panel.grid.major.x = element_blank()
    )+
    facet_wrap(.~Eco, scales = "free")

bioshifts |>
    summarise(Polewards = length(which((Hemisphere == "North hemisphere" & bvel_lat>0) | 
                                           Hemisphere == "South hemisphere" & bvel_lat<0)),
              Equatorwards = length(which((Hemisphere == "North hemisphere" & bvel_lat<0) | 
                                              Hemisphere == "South hemisphere" & bvel_lat>0)))

bioshifts |>
    group_by(Eco) |>
    summarise(Polewards = length(which((Hemisphere == "North hemisphere" & bvel_lat>0) | 
                                           Hemisphere == "South hemisphere" & bvel_lat<0)),
              Equatorwards = length(which((Hemisphere == "North hemisphere" & bvel_lat<0) | 
                                              Hemisphere == "South hemisphere" & bvel_lat>0)))

```

#### Hemisphere

```{r}

# Define the breaks
breaks <- seq(0, 360, length.out = 25)

# Calculate bin width
bin_width <- diff(breaks)[1]

# Bin the data and calculate midpoints and counts
hist_data <- bioshifts |>
    group_by(Hemisphere) |>
    mutate(bin = cut(bvel_angle, breaks = breaks, include.lowest = TRUE, right = FALSE, labels = FALSE)) |>
    group_by(bin,Hemisphere) |>
    summarise(counts = n()) |>
    mutate(midpoints = (breaks[bin] + breaks[bin + 1]) / 2)

# Select and arrange the relevant columns
hist_data <- hist_data |>
    select(midpoints, counts, Hemisphere) |>
    arrange(midpoints)

ggplot(hist_data, aes(x = midpoints, y = counts))+
    ggtitle("Frequence of bioclimatic direction")+
    geom_bar(stat = "identity", aes(fill=Hemisphere)) +
    scale_x_continuous(breaks = seq(from=0, to=359, by=45), limits=c(0, 360)) +
    coord_polar() +
    theme_minimal()+
    labs(x="",y="")+
    theme(legend.position = "none",
          panel.grid.major.x = element_blank()
    )+
    facet_wrap(.~Hemisphere, scales = "free")

```

#### Hemisphere vs Eco

```{r, fig.height=6, fig.width=6}

# Bin the data and calculate midpoints and counts
hist_data <- bioshifts |>
    group_by(Eco, Hemisphere) |>
    mutate(bin = cut(bvel_angle, breaks = breaks, include.lowest = TRUE, right = FALSE, labels = FALSE)) |>
    group_by(Eco,bin, Hemisphere) |>
    summarise(counts = n()) |>
    mutate(midpoints = (breaks[bin] + breaks[bin + 1]) / 2,
           Eco = case_when(Eco == "Ter"~"Terrestrial",
                           Eco == "Mar"~"Marine"))

# Select and arrange the relevant columns
hist_data <- hist_data |>
    select(Eco,midpoints, counts, Hemisphere) |>
    arrange(midpoints)

ggplot(hist_data, aes(x = midpoints, y = counts,
                      fill = Eco))+
    # ggtitle("Frequence of bioclimatic direction")+
    geom_bar(stat = "identity") +
    scale_x_continuous(breaks = seq(from=0, to=359, by=45), limits=c(0, 360)) +
    coord_polar() +
    scale_fill_manual(values = c("dodgerblue","orange4"))+
    theme_minimal()+
    labs(x="",y="")+
    theme(legend.position = "none",
          panel.grid.major.x = element_blank()
    )+
    facet_wrap(Hemisphere~Eco, scales = "free")


ggsave(filename = here("Figures/SDM_angle_all.png"), 
       device = "png", units = "in", dpi = 300,
       width = 6, height = 6)

```

<!-- #### Hemisphere vs Eco vs Param -->

```{r fig.width=10,fig.height=15}

# Bin the data and calculate midpoints and counts
hist_data <- bioshifts |>
    group_by(Eco,Param,Hemisphere) |>
    mutate(bin = cut(bvel_angle, breaks = breaks, include.lowest = TRUE, right = FALSE, labels = FALSE)) |>
    group_by(Eco,Param,bin,Hemisphere) |>
    summarise(counts = n()) |>
    mutate(midpoints = (breaks[bin] + breaks[bin + 1]) / 2)

# Select and arrange the relevant columns
hist_data <- hist_data |>
    select(Eco,midpoints, counts,Hemisphere) |>
    arrange(midpoints)

ggplot(hist_data, aes(x = midpoints, y = counts,
                      fill = Eco))+
    ggtitle("Frequence of bioclimatic direction")+
    geom_bar(stat = "identity", width=12) +
    scale_x_continuous(breaks = seq(from=0, to=359, by=45), limits=c(0, 360)) +
    coord_polar() +
    scale_fill_manual(values = c("dodgerblue","orange4"))+
    theme_minimal()+
    labs(x="",y="")+
    theme(legend.position = "none",
          panel.grid.major.x = element_blank()
    )+
    facet_wrap(Hemisphere+Eco~Param, drop = FALSE, ncol = 3)



```

### Magnitude

```{r}

# difference in magnitude
test <- 
    bioshifts |> 
    group_by(Eco) |> 
    filter(!ShiftRate==0,
           !bvel_lat==0) |> 
    mutate(Frac = (ShiftRate - bvel_lat)/ShiftRate)

pos <- which(test$ShiftRate < 0)
test$Frac[pos] <- test$Frac[pos] * -1

test |> 
    group_by(Eco) |>
    summarise(Frac = mean(Frac))

```

#### Hemisphere

```{r}
ggplot(bioshifts, aes(x = bvel_angle, 
                      y = abs(bvel_lat), color = Hemisphere))+
    ggtitle("Absolute magnitude of bioclimatic direction (km/year)")+
    geom_point(alpha=.5) +
    geom_linerange(aes(ymin = 0, ymax = abs(bvel_lat)), alpha=.5) +
    scale_x_continuous(breaks = seq(from=0, to=359, by=45), limits=c(0, 360)) +
    coord_polar() +
    theme_minimal()+
    labs(x="",y="")+
    theme(legend.position = "none",
          panel.grid.major.x = element_blank()
    )+
    facet_wrap(.~Hemisphere)
```

#### Hemisphere vs Eco

```{r, fig.height=8, fig.width=8}
ggplot(bioshifts, aes(x = bvel_angle, 
                      y = abs(bvel_lat),
                      color = Eco))+
    ggtitle("Absolute magnitude of bioclimatic direction (km/year)")+
    geom_point(alpha=.5) +
    geom_linerange(aes(ymin = 0, ymax = abs(bvel_lat)), alpha=.5) +
    scale_color_manual(values = c("dodgerblue","orange4"))+
    scale_x_continuous(breaks = seq(from=0, to=359, by=45), limits=c(0, 360)) +
    coord_polar() +
    theme_minimal()+
    labs(x="",y="")+
    theme(legend.position = "none",
          panel.grid.major.x = element_blank()
    )+
    facet_wrap(Hemisphere~Eco)
```

#### Hemisphere vs Eco vs Param

```{r fig.width=10,fig.height=15}

ggplot(bioshifts, aes(x = bvel_angle, 
                      y = abs(bvel_lat),
                      color = Eco))+
    ggtitle("Absolute magnitude of bioclimatic direction (km/year)")+
    geom_point(alpha=.5) +
    geom_linerange(aes(ymin = 0, ymax = abs(bvel_lat)), alpha=.5) +
    scale_color_manual(values = c("dodgerblue","orange4"))+
    scale_x_continuous(breaks = seq(from=0, to=359, by=45), limits=c(0, 360)) +
    coord_polar() +
    theme_minimal()+
    labs(x="",y="")+
    theme(legend.position = "none",
          panel.grid.major.x = element_blank()
    )+
    facet_wrap(Hemisphere+Eco~Param, scales = "free", drop = FALSE, ncol = 3)

```

### How much of the angle variability is poleward?

```{r}
# Calculate the proportion of the non-projected shift captured by latitude

bioshifts |>
    group_by(Eco) |>
    mutate(proportions = abs(bvel_lat) / bvel) |>
    select(proportions,Eco) |>
    na.omit() |>
    summarise(proportions_mean = mean(proportions),
              proportions_sd = sd(proportions))

# variance explained
data_mar <- 
    plot(bvel ~ abs(bvel_lat),
         data = bioshifts |>
             filter(Eco == "Mar"))
mod_mar <- glm(bvel ~ abs(bvel_lat),
               data = bioshifts |>
                   filter(Eco == "Mar"))

r.squaredLR(mod_mar)

# Concordance Correlation Coefficient (CCC)
tmp <- DescTools::CCC(mod_mar$model$bvel,mod_mar$model$`abs(bvel_lat)`)
tmp$rho.c
# Spearman’s Rho (rank-based correlation)
spearmanRho(bvel ~ abs(bvel_lat),
            data = bioshifts |>
                filter(Eco == "Mar"))

# Q² (Predictive R²) via cross-validation
# 10-fold cross-validation
cv_result <- boot::cv.glm(bioshifts |>
                              filter(Eco == "Mar"), mod_mar, K = 10)

# Extract cross-validated prediction error
PRESS <- cv_result$delta[1]

# Total sum of squares
TSS <- sum((mod_mar$model$bvel - mean(mod_mar$model$bvel))^2)

# Q²
Q2 <- 1 - (PRESS * nrow(mtcars)) / TSS
Q2


###########
plot(bvel ~ abs(bvel_lat),
     data = bioshifts |>
         filter(Eco == "Ter"))
mod_ter <- glm(bvel ~ abs(bvel_lat),
               data = bioshifts |>
                   filter(Eco == "Ter"))

r.squaredLR(mod_ter)
# Concordance Correlation Coefficient (CCC)
tmp <- DescTools::CCC(mod_ter$model$bvel,mod_ter$model$`abs(bvel_lat)`)
tmp$rho.c
# Spearman’s Rho (rank-based correlation)
spearmanRho(bvel ~ abs(bvel_lat),
            data = bioshifts |>
                filter(Eco == "Ter"))

# Q² (Predictive R²) via cross-validation
# 10-fold cross-validation
cv_result <- boot::cv.glm(bioshifts |>
                              filter(Eco == "Ter"), mod_ter, K = 10)

# Extract cross-validated prediction error
PRESS <- cv_result$delta[1]

# Total sum of squares
TSS <- sum((mod_ter$model$bvel - mean(mod_ter$model$bvel))^2)

# Q²
Q2 <- 1 - (PRESS * nrow(mtcars)) / TSS
Q2


```

### Mean angle

```{r}
# Devide North and South hemisphere

# Terrestrial
bioshifts |>
    group_by(Hemisphere,Eco) |>
    count()

tmp <- bioshifts |>
    group_by(Eco, Hemisphere) |>
    summarise(mean_angle = mean(circular(na.omit(bvel_angle), 
                                         units = "degrees", 
                                         template = "geographics",
                                         modulo = "2pi")),
              sd_angle = sd(circular(na.omit(bvel_angle), 
                                     units = "degrees", 
                                     template = "geographics",
                                     modulo = "2pi")),
              var_angle = var(circular(na.omit(bvel_angle), 
                                       units = "degrees", 
                                       template = "geographics",
                                       modulo = "2pi")))
tmp

ggplot(tmp, aes(x = mean_angle, y = 1))+
    geom_bar(stat = "identity", aes(fill=Eco)) +
    scale_x_continuous(breaks = seq(from=0, to=359, by=45), limits=c(0, 360)) +
    coord_polar() +
    theme_minimal()+
    labs(x="",y="")+
    theme(legend.position = "none",
          panel.grid.major.x = element_blank()
    )+
    facet_wrap(Hemisphere~Eco, scales = "free")

```

## Direction match

```{r fig.height=4.2, fig.width=4.2}

d1 <- ggplot(bioshifts |>
                 filter(ShiftRate!=0) |>
                 group_by(match_direction_class) |>
                 count() |>
                 na.omit() |> 
                 data.frame() |>
                 mutate(n = (n/sum(n))*100,
                        bioclimatic = c("Equatorward","Equatorward","Poleward","Poleward"),
                        biotic = c("Equatorward","Poleward","Equatorward","Poleward"),
                        match_direction_class = match_direction_class), 
             aes(x = bioclimatic, y = biotic)) + 
    geom_point(aes(size = n, fill = as.factor(match_direction_class)), 
               alpha = 0.75, shape = 22) + 
    scale_fill_manual(values = c("negneg" = "#56B4E9", "pospos" = "#56B4E9",
                                 "posneg" = "#CC79A7", "negpos" = "#CC79A7"))+
    geom_text(aes(label = paste(round(n,1),"%")), alpha = 0.75, shape = 22) + 
    scale_size_continuous(limits = c(1, 100), range = c(1,100), breaks = c(1,10,50,75,100)) +
    labs( x= "Bioclimatic direction", y = "Biotic direction", 
          size = "Relative frequency (%)", fill = "")  + 
    theme(legend.key=element_blank(), 
          axis.text.y = element_text(angle = 90, vjust = 0.5, hjust=.5),
          panel.background = element_blank(), 
          panel.border = element_rect(colour = "black", fill = NA, size = 1), 
          legend.position = "none") 

d1

```

```{r fig.width=4.5, fig.height=3}

d2 <- ggplot(bioshifts |>
                 filter(ShiftRate!=0) |>
                 group_by(Eco, match_direction_class) |>
                 count() |>
                 na.omit() |> 
                 data.frame() |>
                 group_by(Eco) |>
                 mutate(n = (n/sum(n))*100) |>
                 ungroup() |>
                 mutate(bioclimatic = rep(c("Equatorward","Equatorward","Poleward","Poleward"),2),
                        biotic = rep(c("Equatorward","Poleward","Equatorward","Poleward"),2),
                        match_direction_class = match_direction_class,
                        Eco = case_when(Eco == "Mar" ~ "Marine",
                                        Eco == "Ter" ~ "Terrestrial")), 
             aes(x = bioclimatic, y = biotic)) + 
    geom_point(aes(size = n, fill = as.factor(match_direction_class)), 
               alpha = 0.5, shape = 22, color = "white") + 
    scale_fill_manual(values = c("negneg" = "gray16", "pospos" = "gray16",
                                 "posneg" = "gray81", "negpos" = "gray81"))+
    geom_text(aes(label = paste(round(n,1),"%")), alpha = 0.75) + 
    scale_size_continuous(limits = c(1, 40), range = c(1,40), breaks = seq(1,40,length.out=4)) +
    labs( x= "Bioclimatic direction", y = "Biotic direction", 
          size = "Relative frequency (%)", fill = "")  + 
    theme(axis.text.y = element_text(angle = 90, vjust = 0.5, hjust=.5),
          panel.background = element_blank(), 
          panel.border = element_blank(), 
          legend.position = "none",
          aspect.ratio=1)+
    facet_wrap(.~Eco)

d2

ggsave(filename = here("Figures/directionPlot_Mar_Ter_all.svg"), device = "svg",
       width = 4.5, height = 3, units = "in")

bioshifts |>
    select(match_direction,Eco) |>
    na.omit() |>
    group_by(Eco) |>
    summarise(match = length(which(match_direction=='match'))/length(match_direction),
              mismatch = length(which(match_direction=='mismatch'))/length(match_direction))

```

## Scatter plots

### Shift vs NS Shift

```{r fig.width=12, fig.height=6, warning=FALSE, message=FALSE, message=FALSE}


ggplot(bioshifts |>
           plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) |>
           mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                                  Eco == "Ter" ~ "Terrestrial")))+
    geom_point(aes(x = bvel_lat, y = bvel_lat2, color = bvel_angle), size = 2, alpha = .5)+
    labs(x="Modeled latitudinal shift velocity (km/yr)\nLatitude/time", 
         y="Modeled latitudinal shift velocity (km/yr)\nAngle transformation")+
    theme_classic() +
    theme(aspect.ratio=1,
          plot.title = element_text(hjust = 0.5)) +
    scale_color_gradientn(colors = myramp_angle)+
    geom_abline(intercept = 0, slope = 1, linetype = "dashed")+
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30")+
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30")+
    facet_wrap(Eco~., scales = "free")


ggplot(bioshifts |>
           plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) |>
           mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                                  Eco == "Ter" ~ "Terrestrial")))+
    geom_point(aes(x = abs(bvel_lat), y = bvel, color = bvel_angle), size = 2, alpha = .5)+
    labs(x="Modeled absolute latitudinal shift velocity (km/yr)\nLatitude/time", 
         y="Modeled range shift velocity (km/yr)")+
    theme_classic() +
    theme(aspect.ratio=1,
          plot.title = element_text(hjust = 0.5)) +
    scale_color_gradientn(colors = myramp_angle)+
    geom_abline(intercept = 0, slope = 1, linetype = "dashed")+
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30")+
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30")+
    facet_wrap(Eco~., scales = "free")

ggplot(bioshifts |>
           plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) |>
           mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                                  Eco == "Ter" ~ "Terrestrial")))+
    geom_point(aes(x = abs(bvel_lat2), y = bvel, color = bvel_angle), size = 2, alpha = .5)+
    labs(x="Modeled absolute latitudinal shift velocity (km/yr)\nAngle transformation", 
         y="Modeled range shift velocity (km/yr)")+
    theme_classic() +
    theme(aspect.ratio=1,
          plot.title = element_text(hjust = 0.5)) +
    scale_color_gradientn(colors = myramp_angle)+
    geom_abline(intercept = 0, slope = 1, linetype = "dashed")+
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30")+
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30")+
    facet_wrap(Eco~., scales = "free")

ggplot(bioshifts |>
           plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) |>
           mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                                  Eco == "Ter" ~ "Terrestrial")))+
    geom_point(aes(x = bvel_lat, y = ShiftRate, color = bvel_angle), size = 2, alpha = .5)+
    labs(x="Modeled latitudinal shift velocity (km/yr)\nLatitude/time", 
         y="Documented latitudinal shift velocity (km/yr)")+
    theme_classic() +
    theme(aspect.ratio=1,
          plot.title = element_text(hjust = 0.5)) +
    scale_color_gradientn(colors = myramp_angle)+
    geom_abline(intercept = 0, slope = 1, linetype = "dashed")+
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30")+
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30")+
    facet_wrap(Eco~., scales = "free")

ggplot(bioshifts |>
           plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) |>
           mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                                  Eco == "Ter" ~ "Terrestrial")))+
    geom_point(aes(x = bvel_lat2, y = ShiftRate, color = bvel_angle), size = 2, alpha = .5)+
    labs(x="Modeled latitudinal shift velocity (km/yr)\nAngle transformation", 
         y="Documented latitudinal shift velocity (km/yr)")+
    theme_classic() +
    theme(aspect.ratio=1,
          plot.title = element_text(hjust = 0.5)) +
    scale_color_gradientn(colors = myramp_angle)+
    geom_abline(intercept = 0, slope = 1, linetype = "dashed")+
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30")+
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30")+
    facet_wrap(Eco~., scales = "free")

plotrix::polar.plot(
    start = 90,
    lengths = c(rnorm(360,mean = 1, sd = 0.001)),
    polar.pos = seq(0,360,by=1),
    clockwise = TRUE,
    cex=.1,
    show.grid.labels=0,
    line.col=myramp_angle)

```

### Eco

```{r fig.width=12, fig.height=6, warning=FALSE, message=FALSE, message=FALSE}

new_data <- as_tibble(bioshifts) |>
    select(ID, sp_name_std, class, Param, bvel_lat, bvel95_lat, ShiftRate, Eco)|>
    group_by(ID, class, Param, Eco) |>
    mutate(Nsps = length(unique(sp_name_std)),
           Eco = case_when(Eco == "Mar" ~ "Marine",
                           Eco == "Ter" ~ "Terrestrial")) |>
    filter(Nsps>3) |>
    summarise(Nsps = mean(Nsps),
              bvel_lat_sa = mean(bvel_lat),
              bvel95_lat_sa = mean(bvel95_lat),
              shift_sa  = mean(ShiftRate),
              bvel_lat_sa_sd = sd(bvel_lat),
              bvel_lat_sa_cv = (bvel_lat_sa_sd/bvel_lat_sa)*100,
              bvel95_lat_sa_sd = sd(bvel95_lat),
              bvel95_lat_sa_cv = (bvel95_lat_sa_sd/bvel95_lat_sa)*100,
              shift_sa_sd = sd(ShiftRate),
              shift_sa_cv = (shift_sa_sd/shift_sa)*100) |>
    plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) 

my_colors <- bioshifts |>
    select(class, hex) |>
    unique()

my_colors_names <- my_colors$class
my_colors <- my_colors$hex
names(my_colors) <- my_colors_names


p1 <- ggplot(bioshifts |>
                 plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) |>
                 mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                                        Eco == "Ter" ~ "Terrestrial"),
                        Eco = factor(Eco, levels = c("Terrestrial", "Marine")))) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30")+
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30")+
    geom_abline(slope=1, linetype = "dashed", color="gray30") + # add 1:1 line  
    geom_point(aes(x = bvel_lat, y = ShiftRate, fill = Class), 
               shape = 21)+
    scale_fill_manual(values = my_colors)+
    labs(x="Modeled latitudinal range shift (km/yr)", 
         y="Documented latitudinal range shift (km/yr)")+
    theme_classic() +
    theme(aspect.ratio=1,
          plot.title = element_text(hjust = 0.5),
          strip.background = element_rect(colour = NA),
          strip.text = element_text(size = 14)) +
    scale_color_manual(values = my_colors)+
    # scale_y_continuous(limits = range(-30,50))+
    facet_wrap(Eco~., scales = "free")

p1

p1_95 <- ggplot(bioshifts |>
                    plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) |>
                    mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                                           Eco == "Ter" ~ "Terrestrial"),
                           Eco = factor(Eco, levels = c("Terrestrial", "Marine")))) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30")+
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30")+
    geom_abline(slope=1, linetype = "dashed", color="gray30") + # add 1:1 line  
    geom_point(aes(x = bvel95_lat, y = ShiftRate, fill = Class), 
               shape = 21)+
    scale_fill_manual(values = my_colors)+
    labs(x="Modeled latitudinal range shift (km/yr)", 
         y="Documented latitudinal range shift (km/yr)")+
    theme_classic() +
    theme(aspect.ratio=1,
          plot.title = element_text(hjust = 0.5),
          strip.background = element_rect(colour = NA),
          strip.text = element_text(size = 14)) +
    scale_color_manual(values = my_colors)+
    # scale_y_continuous(limits = range(-30,50))+
    facet_wrap(Eco~., scales = "free")

p1_95

```



```{r fig.width=12, fig.height=6, warning=FALSE, message=FALSE, message=FALSE}

p2 <- p1 + 
    geom_point(data = new_data,
               aes(x=bvel_lat_sa,y=shift_sa, 
                   size = Nsps, color = Class, fill = Class), 
               shape = 21, 
               alpha=.3) +
    scale_size_continuous(range = c(5,10),
                          breaks = seq(4,100,length.out=3))+
    geom_linerange(data = new_data,
                   aes(x=bvel_lat_sa,y=shift_sa,
                       xmin = bvel_lat_sa-bvel_lat_sa_sd,
                       xmax = bvel_lat_sa+bvel_lat_sa_sd), 
                   alpha = .3) +
    geom_linerange(data = new_data,
                   aes(x=bvel_lat_sa,y=shift_sa,
                       ymin = shift_sa-shift_sa_sd,
                       ymax = shift_sa+shift_sa_sd), 
                   alpha = .3)

p2

p3 <- p1 + 
    geom_point(data = new_data,
               aes(x=bvel_lat_sa,y=shift_sa, size = Nsps), 
               shape = 21, 
               alpha=.3) +
    scale_size_continuous(range = c(5,10),
                          breaks = seq(4,100,length.out=3))+
    geom_linerange(data = new_data,
                   aes(x=bvel_lat_sa,y=shift_sa,
                       xmin = bvel_lat_sa-bvel_lat_sa_sd,
                       xmax = bvel_lat_sa+bvel_lat_sa_sd), 
                   alpha = .3) +
    geom_linerange(data = new_data,
                   aes(x=bvel_lat_sa,y=shift_sa,
                       ymin = shift_sa-shift_sa_sd,
                       ymax = shift_sa+shift_sa_sd), 
                   alpha = .3)

p3

p4 <- ggplot()+
    geom_point(data = new_data,
               aes(x=bvel_lat_sa,y=shift_sa, size = Nsps, color = Class, fill = Class), 
               shape = 21, 
               alpha=.3) +
    scale_size_continuous(range = c(5,10),
                          breaks = seq(4,100,length.out=3))+
    geom_linerange(data = new_data,
                   aes(x=bvel_lat_sa,y=shift_sa,
                       xmin = bvel_lat_sa-bvel_lat_sa_sd,
                       xmax = bvel_lat_sa+bvel_lat_sa_sd), 
                   alpha = .3) +
    geom_linerange(data = new_data,
                   aes(x=bvel_lat_sa,y=shift_sa,
                       ymin = shift_sa-shift_sa_sd,
                       ymax = shift_sa+shift_sa_sd), 
                   alpha = .3)+
    labs(x="Modeled latitudinal range shift (km/yr)", 
         y="Documented latitudinal range shift (km/yr)")+
    theme_classic() +
    theme(aspect.ratio=1,
          plot.title = element_text(hjust = 0.5)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30")+
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30")+
    facet_wrap(Eco~., scales = "free_x")

p4

```


### Load model results

```{r}

sm_SA_glmm_res <- read.csv(here("Data/Output/res_sm_SA_glmm_all.csv"))

# remember that here a significant p-value means the slope significantly diverges from the 1:1 relationship. therefore, we code P-value < 0.5 as 0.
sm_SA_glmm_res$sig <- ifelse(sm_SA_glmm_res$P.value < .05, 0, 1)

sm_SA_glmm_res <- sm_SA_glmm_res |>
    mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                           Eco == "Ter" ~ "Terrestrial")) |>
    select(Param, Estimate, Eco, sig)

sm_SA_glmm_res <- sm_SA_glmm_res |>
    pivot_wider(names_from = Param, values_from = c(Estimate,sig))
names(sm_SA_glmm_res) <- c("Eco","intercept","slope","sig_Intercept", "sig_slope")

sm_SA_glmm_res$Eco <- factor(sm_SA_glmm_res$Eco, levels = c("Marine","Terrestrial"))


sm_glmm_res <- read.csv(here("Data/Output/res_glmm_sm_all.csv"))


# remember that here a significant p-value means the slope significantly diverges from the 1:1 relationship. therefore, we code P-value < 0.5 as 0.
sm_glmm_res$sig <- ifelse(sm_glmm_res$P.value < .05, 0, 1)

sm_glmm_res <- sm_glmm_res |>
    mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                           Eco == "Ter" ~ "Terrestrial"))|>
    select(Param, Estimate, Eco, sig)

sm_glmm_res <- sm_glmm_res |>
    pivot_wider(names_from = Param, values_from = c(Estimate,sig))
names(sm_glmm_res) <- c("Eco","intercept","slope","sig_Intercept", "sig_slope")

sm_glmm_res$Eco <- factor(sm_glmm_res$Eco, levels = c("Marine","Terrestrial"))

```

#### Inset plot (match)

```{r fig.width=8, fig.height=6, warning=FALSE, message=FALSE, message=FALSE}

plot_inset_dir_theme <- list(
    scale_fill_manual(values = c("negneg" = "#404788FF", "pospos" = "#404788FF",
                                 "posneg" = "#FDE725FF", "negpos" = "#FDE725FF")),
    geom_text(aes(label = paste(round(n,1),"%")), alpha = 0.75, size = 3), 
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30"),
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30"),
    scale_size_continuous(limits = c(1, 50), range = c(1,20), breaks = seq(1,50,length.out=4)),
    scale_x_continuous(limits = c(-2,2)),
    scale_y_continuous(limits = c(-2,2)),
    labs( x= "", y = ""),
    theme_void(),
    theme(panel.background = element_blank(), 
          panel.border = element_blank(), 
          legend.position = "none",
          aspect.ratio=1))

plot_inset_SA_theme <- list(
    scale_size_continuous(range = c(1,5),
                          breaks = seq(1,50,length.out=10)),
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30"),
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30"),
    labs(x="", y=""),
    scale_linetype_manual(values= c("1"="solid","0"="dotted")),
    theme_bw(),
    theme(legend.position="none", aspect.ratio=1, 
          axis.text = element_text(size = 8),
          panel.grid = element_blank(),
          panel.background = element_rect(fill='white',
                                          color='white'),
          plot.background = element_rect(fill='transparent',
                                         color='transparent')))

plot_inset_dir_mar <- ggplot(bioshifts |>
                                 filter(Eco == "Mar") |> 
                                 group_by(match_direction_class) |>
                                 count() |>
                                 na.omit() |> 
                                 data.frame() |>
                                 mutate(n = (n/sum(n))*100) |>
                                 ungroup() |>
                                 mutate(
                                     bioclimatic = c(-1,-1,1,1),
                                     biotic = c(-1,1,-1,1),
                                     match_direction_class = match_direction_class), 
                             aes(x = bioclimatic, y = biotic)) + 
    geom_point(aes(size = n, fill = as.factor(match_direction_class)), 
               alpha = 0.5, shape = 22, color = "white") + 
    plot_inset_dir_theme

plot_inset_dir_ter <- ggplot(bioshifts |>
                                 filter(Eco == "Ter") |> 
                                 group_by(match_direction_class) |>
                                 count() |>
                                 na.omit() |> 
                                 data.frame() |>
                                 mutate(n = (n/sum(n))*100) |>
                                 ungroup() |>
                                 mutate(
                                     bioclimatic = c(-1,-1,1,1),
                                     biotic = c(-1,1,-1,1),
                                     match_direction_class = match_direction_class), 
                             aes(x = bioclimatic, y = biotic)) + 
    geom_point(aes(size = n, fill = as.factor(match_direction_class)), 
               alpha = 0.5, shape = 22, color = "white") +  
    plot_inset_dir_theme


```


```{r fig.width=9, fig.height=6, warning=FALSE, message=FALSE, message=FALSE}
pp <- p1 + 
    geom_abline(
        data = sm_glmm_res,
        aes(intercept = intercept, slope = slope, 
            linetype = as.factor(sig_slope)), 
        color = ifelse(sm_glmm_res$sig_slope==1,"red",NA)) +
    scale_linetype_manual(values= c("1"="solid","0"="dotted"))+
    # coord_cartesian(xlim = c(-20,25),
    #                 ylim = c(-30,42))+
    facetted_pos_scales(
        x = list(
            scale_x_continuous(limits = c(-20,25)),
            NULL),
        y = list(
            scale_y_continuous(limits = c(-30,42)),
            NULL)) +
    theme(legend.position="none",
          strip.text.x = element_text(vjust = 2),
          panel.spacing = unit(2, "lines")) # Panel 2

pg <- plot_grid(pp, 
                GP_legend,  rel_heights = c(1, .4), nrow = 2)

plot_inset_SA_mar <-
    ggplot(data = new_data |>
               filter(Eco == "Marine")) +
    plot_inset_SA_theme +
    coord_cartesian(xlim = c(-2,3),
                    ylim = c(-5,12)) + 
    scale_fill_manual(values = my_colors)+
    scale_color_manual(values = my_colors) +
    geom_linerange(aes(x=bvel_lat_sa,y=shift_sa,
                       ymin = shift_sa-shift_sa_sd,
                       ymax = shift_sa+shift_sa_sd, 
                       color = Class), 
                   alpha = .3)+
    geom_linerange(aes(x=bvel_lat_sa,y=shift_sa,
                       xmin = bvel_lat_sa-bvel_lat_sa_sd,
                       xmax = bvel_lat_sa+bvel_lat_sa_sd, 
                       color = Class), 
                   alpha = .3)+
    geom_point(aes(x=bvel_lat_sa,y=shift_sa, size = Nsps, fill = Class), shape = 21)+
    geom_abline(slope=1, linetype = "dashed", color="gray30") + # add 1:1 line+
    geom_abline(
        data = sm_SA_glmm_res |>
            # data = sm_SA_deming_res |>
            filter(Eco == "Marine"),
        aes(intercept = intercept, slope = slope, linetype = as.factor(sig_slope)), 
        color = ifelse(sm_SA_glmm_res$sig_slope[which(sm_SA_glmm_res$Eco=="Marine")]==1,"red",NA)) 

plot_inset_SA_ter <-
    ggplot(data = new_data |>
               filter(Eco == "Terrestrial")) +
    plot_inset_SA_theme +
    coord_cartesian(xlim = c(-1.5,2.5),
                    ylim = c(-15,15))+
    scale_fill_manual(values = my_colors)+
    scale_color_manual(values = my_colors) +
    geom_linerange(aes(x=bvel_lat_sa,y=shift_sa,
                       ymin = shift_sa-shift_sa_sd,
                       ymax = shift_sa+shift_sa_sd, 
                       color = Class), 
                   alpha = .3)+
    geom_linerange(aes(x=bvel_lat_sa,y=shift_sa,
                       xmin = bvel_lat_sa-bvel_lat_sa_sd,
                       xmax = bvel_lat_sa+bvel_lat_sa_sd, 
                       color = Class), 
                   alpha = .3)+
    geom_point(aes(x=bvel_lat_sa,y=shift_sa, size = Nsps, fill = Class), shape = 21)+
    geom_abline(slope=1, linetype = "dashed", color="gray30") + # add 1:1 line+
    geom_abline(data = sm_SA_glmm_res |>
                    # data = sm_SA_deming_res |>
                    filter(Eco == "Terrestrial"),
                aes(intercept = intercept, 
                    slope = slope, 
                    linetype = as.factor(sig_slope)), 
                color = ifelse(sm_SA_glmm_res$sig_slope[which(sm_SA_glmm_res$Eco=="Terrestrial")]==1,"red",NA))

plot_w_inset <-
    ggdraw() +
    draw_plot(pg) +
    draw_plot(plot_inset_dir_mar, x = .537, y = .74, width = .2, height = .2) +
    draw_plot(plot_inset_dir_ter, x = 0.078, y = .74, width = .2, height = .2) +
    draw_plot(plot_inset_SA_mar, x = .72, y = .65, width = .33, height = .33) +
    draw_plot(plot_inset_SA_ter, x = .26, y = .65, width = .33, height = .33)

plot_w_inset
```


```{r fig.width=9, fig.height=6, warning=FALSE, message=FALSE, message=FALSE}

ggsave(filename = here("Figures/Scatter_Mar_Ter_inset.svg"), device = "svg",
       width = 9, height = 6, units = "in")


```

### Load model results 95

```{r}

sm_SA_glmm_res <- read.csv(here("Data/Output/res_sm_SA_glmm95_all.csv"))

# remember that here a significant p-value means the slope significantly diverges from the 1:1 relationship. therefore, we code P-value < 0.5 as 0.
sm_SA_glmm_res$sig <- ifelse(sm_SA_glmm_res$P.value < .05, 0, 1)

sm_SA_glmm_res <- sm_SA_glmm_res |>
    mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                           Eco == "Ter" ~ "Terrestrial")) |>
    select(Param, Estimate, Eco, sig)

sm_SA_glmm_res <- sm_SA_glmm_res |>
    pivot_wider(names_from = Param, values_from = c(Estimate,sig))
names(sm_SA_glmm_res) <- c("Eco","intercept","slope","sig_Intercept", "sig_slope")

sm_SA_glmm_res$Eco <- factor(sm_SA_glmm_res$Eco, levels = c("Marine","Terrestrial"))


sm_glmm_res <- read.csv(here("Data/Output/res_glmm_sm95_all.csv"))


# remember that here a significant p-value means the slope significantly diverges from the 1:1 relationship. therefore, we code P-value < 0.5 as 0.
sm_glmm_res$sig <- ifelse(sm_glmm_res$P.value < .05, 0, 1)

sm_glmm_res <- sm_glmm_res |>
    mutate(Eco = case_when(Eco == "Mar" ~ "Marine",
                           Eco == "Ter" ~ "Terrestrial"))|>
    select(Param, Estimate, Eco, sig)

sm_glmm_res <- sm_glmm_res |>
    pivot_wider(names_from = Param, values_from = c(Estimate,sig))
names(sm_glmm_res) <- c("Eco","intercept","slope","sig_Intercept", "sig_slope")

sm_glmm_res$Eco <- factor(sm_glmm_res$Eco, levels = c("Marine","Terrestrial"))


```

#### Inset plot 95 (match + SA)

```{r fig.width=8, fig.height=6, warning=FALSE, message=FALSE, message=FALSE}

plot_inset_dir_theme <- list(
    scale_fill_manual(values = c("negneg" = "#404788FF", "pospos" = "#404788FF",
                                 "posneg" = "#FDE725FF", "negpos" = "#FDE725FF")),
    geom_text(aes(label = paste(round(n,1),"%")), alpha = 0.75, size = 3), 
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30"),
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30"),
    scale_size_continuous(limits = c(1, 50), range = c(1,20), breaks = seq(1,50,length.out=4)),
    scale_x_continuous(limits = c(-2,2)),
    scale_y_continuous(limits = c(-2,2)),
    labs( x= "", y = ""),
    theme_void(),
    theme(panel.background = element_blank(), 
          panel.border = element_blank(), 
          legend.position = "none",
          aspect.ratio=1))

plot_inset_SA_theme <- list(
    scale_size_continuous(range = c(1,5),
                          breaks = seq(1,50,length.out=10)),
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray30"),
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray30"),
    labs(x="", y=""),
    scale_linetype_manual(values= c("1"="solid","0"="dotted")),
    theme_bw(),
    theme(legend.position="none", aspect.ratio=1, 
          axis.text = element_text(size = 8),
          panel.grid = element_blank(),
          panel.background = element_rect(fill='white',
                                          color='white'),
          plot.background = element_rect(fill='transparent',
                                         color='transparent')))

plot_inset_dir_mar <- ggplot(bioshifts |>
                                 filter(Eco == "Mar") |> 
                                 group_by(match_direction95_class) |>
                                 count() |>
                                 na.omit() |> 
                                 data.frame() |>
                                 mutate(n = (n/sum(n))*100) |>
                                 ungroup() |>
                                 mutate(
                                     bioclimatic = c(-1,-1,1,1),
                                     biotic = c(-1,1,-1,1),
                                     match_direction_class = match_direction95_class), 
                             aes(x = bioclimatic, y = biotic)) + 
    geom_point(aes(size = n, fill = as.factor(match_direction_class)), 
               alpha = 0.5, shape = 22, color = "white") + 
    plot_inset_dir_theme

plot_inset_dir_ter <- ggplot(bioshifts |>
                                 filter(Eco == "Ter") |> 
                                 group_by(match_direction95_class) |>
                                 count() |>
                                 na.omit() |> 
                                 data.frame() |>
                                 mutate(n = (n/sum(n))*100) |>
                                 ungroup() |>
                                 mutate(
                                     bioclimatic = c(-1,-1,1,1),
                                     biotic = c(-1,1,-1,1),
                                     match_direction_class = match_direction95_class), 
                             aes(x = bioclimatic, y = biotic)) + 
    geom_point(aes(size = n, fill = as.factor(match_direction_class)), 
               alpha = 0.5, shape = 22, color = "white") +  
    plot_inset_dir_theme


```


```{r fig.width=9, fig.height=6, warning=FALSE, message=FALSE, message=FALSE}
pp <- p1_95 + 
    geom_abline(
        data = sm_glmm_res,
        aes(intercept = intercept, slope = slope, 
            linetype = as.factor(sig_slope)), 
        color = ifelse(sm_glmm_res$sig_slope==1,"red",NA)) +
    scale_linetype_manual(values= c("1"="solid","0"="dotted"))+
    # coord_cartesian(xlim = c(-20,25),
    #                 ylim = c(-30,42))+
    facetted_pos_scales(
        x = list(
            scale_x_continuous(limits = c(-20,25)),
            NULL),
        y = list(
            scale_y_continuous(limits = c(-30,42)),
            NULL)) +
    theme(legend.position="none",
          strip.text.x = element_text(vjust = 2),
          panel.spacing = unit(2, "lines")) # Panel 2

pg <- plot_grid(pp, 
                GP_legend,  rel_heights = c(1, .4), nrow = 2)

plot_inset_SA_mar <-
    ggplot(data = new_data |>
               filter(Eco == "Marine")) +
    plot_inset_SA_theme +
    coord_cartesian(xlim = c(-2,3),
                    ylim = c(-5,12)) + 
    scale_fill_manual(values = my_colors)+
    scale_color_manual(values = my_colors) +
    geom_linerange(aes(x=bvel95_lat_sa,y=shift_sa,
                       ymin = shift_sa-shift_sa_sd,
                       ymax = shift_sa+shift_sa_sd, 
                       color = Class), 
                   alpha = .3)+
    geom_linerange(aes(x=bvel95_lat_sa,y=shift_sa,
                       xmin = bvel95_lat_sa-bvel95_lat_sa_sd,
                       xmax = bvel95_lat_sa+bvel95_lat_sa_sd, 
                       color = Class), 
                   alpha = .3)+
    geom_point(aes(x=bvel95_lat_sa,y=shift_sa, size = Nsps, fill = Class), shape = 21)+
    geom_abline(slope=1, linetype = "dashed", color="gray30") + # add 1:1 line+
    geom_abline(
        data = sm_SA_glmm_res |>
            # data = sm_SA_deming_res |>
            filter(Eco == "Marine"),
        aes(intercept = intercept, slope = slope, linetype = as.factor(sig_slope)), 
        color = ifelse(sm_SA_glmm_res$sig_slope[which(sm_SA_glmm_res$Eco=="Marine")]==1,"red",NA)) 

plot_inset_SA_ter <-
    ggplot(data = new_data |>
               filter(Eco == "Terrestrial")) +
    plot_inset_SA_theme +
    coord_cartesian(xlim = c(-1.5,2.5),
                    ylim = c(-15,15))+
    scale_fill_manual(values = my_colors)+
    scale_color_manual(values = my_colors) +
    geom_linerange(aes(x=bvel95_lat_sa,y=shift_sa,
                       ymin = shift_sa-shift_sa_sd,
                       ymax = shift_sa+shift_sa_sd, 
                       color = Class), 
                   alpha = .3)+
    geom_linerange(aes(x=bvel95_lat_sa,y=shift_sa,
                       xmin = bvel95_lat_sa-bvel95_lat_sa_sd,
                       xmax = bvel95_lat_sa+bvel95_lat_sa_sd, 
                       color = Class), 
                   alpha = .3)+
    geom_point(aes(x=bvel95_lat_sa,y=shift_sa, size = Nsps, fill = Class), shape = 21)+
    geom_abline(slope=1, linetype = "dashed", color="gray30") + # add 1:1 line+
    geom_abline(data = sm_SA_glmm_res |>
                    # data = sm_SA_deming_res |>
                    filter(Eco == "Terrestrial"),
                aes(intercept = intercept, 
                    slope = slope, 
                    linetype = as.factor(sig_slope)), 
                color = ifelse(sm_SA_glmm_res$sig_slope[which(sm_SA_glmm_res$Eco=="Terrestrial")]==1,"red",NA))

plot_w_inset <-
    ggdraw() +
    draw_plot(pg) +
    draw_plot(plot_inset_dir_mar, x = .537, y = .74, width = .2, height = .2) +
    draw_plot(plot_inset_dir_ter, x = 0.078, y = .74, width = .2, height = .2) +
    draw_plot(plot_inset_SA_mar, x = .72, y = .65, width = .33, height = .33) +
    draw_plot(plot_inset_SA_ter, x = .26, y = .65, width = .33, height = .33)

plot_w_inset
```

```{r fig.width=9, fig.height=6, warning=FALSE, message=FALSE, message=FALSE}

ggsave(filename = here("Figures/Scatter_Mar_Ter_inset95.svg"), device = "svg",
       width = 9, height = 6, units = "in")

```
