---
params:
    shiftmodtype:
    models_included:
title: "SDMs vs Range shifts: Models"
subtitle: "Shift Mod: `r params$shiftmodtype`. Models included: `r params$models_included`"
author: "Brunno Oliveira & Bioshifts working group"
output:
  html_document
---



```{r include=FALSE, eval=FALSE}

rm(list=ls())
gc()

# Params
shiftmodtype_to_go <- "75" # select 95 or 75
models_included_to_go <- "all" # select "best" or "all" only the best models (TSS > .5)

```



```{r include=FALSE, eval=FALSE}

# render script
# run this at the console

library(rmarkdown)

output_file <- here::here(paste0("R/10_analyses/3_SDM_models_report","_shift-",shiftmodtype_to_go,"_models-",models_included_to_go,".html"))

rmarkdown::render(
    params = list(shiftmodtype = shiftmodtype_to_go,
                  models_included = models_included_to_go),
    input = here::here("R/10_analyses/3_SDM_models.qmd"),
    output_file = output_file,
    output_format = html_document(toc = TRUE,
                                  toc_depth = 2,
                                  toc_float = TRUE,
                                  number_sections = TRUE,
                                  code_folding = "hide",
                                  code_download = TRUE,
                                  code_summary = "Code"))

```



```{r include=FALSE}

library(rmarkdown)

# rm(list=ls())
# gc()

if(is.null(params$shiftmodtype)){
    shiftmodtype = shiftmodtype_to_go
} else {
    shiftmodtype = params$shiftmodtype
}

if(is.null(params$models_included)){
    models_included = models_included_to_go
} else {
    models_included = params$models_included
}
```

# *Setup*

```{r warning=FALSE, message=FALSE}

list.of.packages <- c("dplyr", "tidyverse", "tidyr", "ggplot2", "knitr", "pbapply", "data.table", "GGally", "here", "terra", "tidyterra", "rcompanion", "lme4", "glmmTMB", "MuMIn", "emmeans", "parallel", "sjPlot", "jtools", "gt", "gtsummary", "deming", "cowplot", "DescTools", "gtools", "ggpp", "gridExtra", "ggeffects", "svglite","RSQLite","hexbin", "ggpubr","rnaturalearth","parallel", "effects","rnaturalearthdata","gpkg","scales")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
sapply(list.of.packages, require, character.only = TRUE)

computer = "personal"
source(here("R/settings.R"))
source(here("R/my_functions.R"))
source(here("R/emm_extra_functions.R"))

options(na.action = "na.fail")

```


# *Load data*

```{r}

# select vars based on shift mode

bioshifts <- read.csv(here("Data/Output/Bioshifts_merge_Exposure_all_new.csv"))

if(shiftmodtype == "95"){
    bioshifts <- bioshifts |>
        mutate(bvel_lat = bvel95_lat,
               lag_overshoot = lag_overshoot95,
               mismatch = mismatch95,
               mismatch_percent = mismatch95_percent,
               mismatch_trend = mismatch95_trend,
               match_direction = match_direction95,
               match_direction_class = match_direction95_class)
}

bioshifts$obs_ID <- 1:nrow(bioshifts)

if(models_included == "best"){
    bioshifts <- bioshifts |>
        filter(TSS > .5)
}

```



Extra fixes for variables
```{r}

# fixing Locomotion_mode
bioshifts$Locomotion_mode = as.character(str_to_sentence(bioshifts$Locomotion_mode))
unique(bioshifts$Locomotion_mode)
# if multiple locomotion modes, use the first (the main locomotion mode)
# bioshifts$Locomotion_mode <- sapply(bioshifts$Locomotion_mode, function(x) {
#   strsplit(x, split = "/", fixed = TRUE)[[1]][1] 
# })

pos <- grep("Flying",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Fly"
pos <- grep("wimming",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Swim"
pos <- grep("alk",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Run/Walk"
pos <- grep("unning",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Run/Walk"
pos <- grep("lithering",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Slither/Crawl"
pos <- grep("rawling",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Slither/Crawl"
pos <- grep("Burrowing",bioshifts$Locomotion_mode)
bioshifts$Locomotion_mode[pos] <- "Sessile"

unique(bioshifts$Locomotion_mode)

# bioshifts$Locomotion_mode <- factor(
#   bioshifts$Locomotion_mode,
#   levels = c("Flying", "Running", "Swimming", "Walking", "Slithering", "Crawling",  "Planktonic", "Sessile", "Burrowing"))
# 
# # group Slithering/Crawling and Running/Walking
# pos <- grep("Slithering",bioshifts$Locomotion_mode)
# bioshifts$Locomotion_mode[pos] <- "Slithering/Crawling"
# pos <- grep("Crawling",bioshifts$Locomotion_mode)
# bioshifts$Locomotion_mode[pos] <- "Slithering/Crawling"
# 
# pos <- grep("Running",bioshifts$Locomotion_mode)
# bioshifts$Locomotion_mode[pos] <- "Running/Walking"
# pos <- grep("Walking",bioshifts$Locomotion_mode)
# bioshifts$Locomotion_mode[pos] <- "Running/Walking"
# 
bioshifts$Locomotion_mode <- factor(
    bioshifts$Locomotion_mode,
    levels = c("Fly", "Run/Walk", "Swim", "Slither/Crawl", "Planktonic", "Sessile"))

# fixing Category
table(bioshifts$Category)
# remove synthesis because there is a only a single study
bioshifts <- bioshifts |> filter(!Category == "Synthesis")
# combine census-resurvey and survey-resurvey 
pos <- grep("Census-Resurvey",bioshifts$Category)
bioshifts$Category[pos] <- "Survey-Resurvey"
table(bioshifts$Category)
bioshifts$Category <- factor(bioshifts$Category, 
                             levels = c("TimeSeries", "CensusPeriods", "Survey-Resurvey"))

# make numeric as.numeric
pos <- which(sapply(bioshifts, is.numeric))
tmp <- lapply(bioshifts[pos], as.numeric)
tmp <- do.call(cbind,tmp)
bioshifts[pos] <- tmp

# remove one outlier species
bioshifts <- bioshifts |>
    filter(!(Eco == "Ter" & bvel_lat < -15))


# remove swimming terrestrials
bioshifts <- bioshifts |>
    filter(!(Eco == "Ter" & Locomotion_mode == "Swim"))

# select classes with more than X observations per edge
N_obs_class = 5
class_param_select <- bioshifts |>
    group_by(class, Param) |>
    tally() |>
    filter(n >= N_obs_class) |>
    mutate(class_param = paste(class,Param))

bioshifts <- bioshifts |> 
    mutate(class_param = paste(class,Param)) |> 
    filter(class_param %in% class_param_select$class_param) |>
    select(!class_param)

```

# Make dataset for modeling
```{r}

mod_data <- bioshifts |>
    mutate(ShiftRate_raw = ShiftRate,
           ShiftMod_raw = bvel_lat,
           ShiftRate = scale(ShiftRate),
           ShiftMod = scale(bvel_lat),
           Lat_extent = scale(log(LatExtentk)),
           Study_duration = scale(Duration),
           Boyce = scale(Boyce),
           Interannual_var = scale(log1p(bvel_lat_error)),
           N_periods = scale(N_periodes),
           Connectivity = scale(connectivity),
           mismatch = abs(mismatch),
           Mobility = as.character(str_to_sentence(Mobility)),
           Locomotion_mode = as.character(str_to_sentence(Locomotion_mode)),
           Life_form = as.character(str_to_sentence(lifeform)),
           Sampling_method = Category,
           Random_ID_Param_Class = paste(ID,Param,class),
           Random_ID_Class = paste(ID,class),
           Random_Param_Class = paste(Param,class),
           
           match_direction = 
               case_when(match_direction=="match" ~ "Alignment",
                         match_direction=="mismatch" ~ "Misalignment"),
           match_direction = relevel(factor(match_direction), ref = "Misalignment"),
           
           mismatch_binary = 
               case_when(match_direction=="Misalignment" ~ 0,
                         match_direction=="Alignment" ~ 1),
           
           Shift_alignment = 
               case_when(match_direction_class=="pospos" ~ "Poleward",
                         match_direction_class=="negneg" ~ "Equatorward",
                         match_direction == "Misalignment" ~ "Misalignment"),
           Shift_alignment = relevel(factor(Shift_alignment), ref = "Misalignment"),
           
           lag_overshoot = 
               case_when(lag_overshoot=="lag" ~ "Lag",
                         lag_overshoot=="overshoot" ~ "Overshoot",
                         lag_overshoot == "misalignment" ~ "Misalignment"),
           lag_overshoot = relevel(factor(lag_overshoot), ref = "Misalignment")) |>
    select(ShiftRate_raw, ShiftMod_raw, ShiftRate, ShiftMod,
           mismatch, match_direction, mismatch_binary, Shift_alignment,
           mismatch_trend, lag_overshoot, lag_overshoot95,
           mismatch, mismatch_percent,
           Lat_extent, Sampling_method, Boyce, Interannual_var, N_periods, Study_duration, Hemisphere,
           Life_form, Mobility, Locomotion_mode, Connectivity, 
           ID, Param, class, Eco, sp_name_std, 
           Random_ID_Param_Class, Random_Param_Class, Random_ID_Class, obs_ID)


```

# Stats

N species
```{r summarizing}
length(unique(mod_data$sp_name_std))

mod_data |> 
    group_by(Param) |> 
    summarise(N=length(unique(sp_name_std)))

mod_data |> 
    group_by(Eco) |> 
    summarise(N=length(unique(sp_name_std)))

mod_data |> 
    group_by(Eco,Param) |> 
    summarise(N=length(unique(sp_name_std)))

mod_data |> 
    group_by(Eco,Hemisphere) |> 
    summarise(N=length(unique(sp_name_std)))

DT::datatable(mod_data |>
                  group_by(Eco,class) |>
                  tally())

```

N range shifts
```{r summarizing1}
nrow(mod_data)

mod_data |> 
    group_by(Param) |> 
    summarise(N=length(sp_name_std))

mod_data |> 
    group_by(Eco) |>
    summarise(N=length(sp_name_std))

mod_data |> 
    group_by(Eco,Param) |>
    summarise(N=length(sp_name_std))
```

N classes 
```{r summarizing3}
mod_data |> 
    group_by(Eco) |>
    summarise(N = length(unique(class)))
```

N studies 
```{r summarizing4}
mod_data |> 
    group_by(Eco) |>
    summarise(N = length(unique(ID)))
```

How faster are documented to modeled shifts?
```{r testfaster}
tmp <- mod_data |>
    select(Eco, ShiftRate_raw, ShiftMod_raw) |>
    pivot_longer(!Eco, names_to = "Shift")

ggplot(tmp, 
       aes(x = value, color = Shift)) +
    geom_density() +
    scale_x_continuous(limits = c(quantile(tmp$value, c(.05, .95))))+
    geom_vline(data = tmp |>
                   group_by(Eco,Shift) |>
                   summarise(M = median(value)),
               aes(xintercept = M, color = Shift))+
    facet_wrap(.~Eco)

tmp <- mod_data |>
    mutate(frac_shift = ShiftRate_raw/ShiftMod_raw) |>
    filter(!is.infinite(mismatch_percent),
           !is.infinite(frac_shift),
           match_direction == "Alignment") |> 
    select(Eco, frac_shift) |>
    pivot_longer(!Eco, names_to = "Shift")

ggplot(tmp, 
       aes(x = value, color = Eco)) +
    geom_density() +
    scale_x_continuous(limits = c(quantile(tmp$value, c(.05, .95))))+
    geom_vline(data = tmp |>
                   group_by(Eco,Shift) |>
                   summarise(M = median(value)),
               aes(xintercept = M, color = Eco))


```


```{r testfaster2}
# how faster are documented to modeled shifts when they match in direction?
table_data <- mod_data |>
    group_by(Eco) |>
    mutate(frac_shift = ShiftRate_raw / ShiftMod_raw) |>
    filter(!is.infinite(mismatch_percent),
           !is.infinite(frac_shift),
           match_direction == "Alignment") |> 
    summarise(
        # Documented_Mean = mean(ShiftRate_raw),
        Documented_Median = median(ShiftRate_raw),
        Documented_SD = sd(ShiftRate_raw),
        
        # Modeled_Mean = mean(ShiftMod_raw),
        Modeled_Median = median(ShiftMod_raw),
        Modeled_SD = sd(ShiftMod_raw),
        
        # Mismatch_Mean = mean(mismatch),
        Mismatch_Median = median(mismatch),
        Mismatch_SD = sd(mismatch),
        
        # `Mismatch-Percent_Mean` = mean(mismatch_percent, na.rm = TRUE),
        `Mismatch-Percent_Median` = median(mismatch_percent, na.rm = TRUE),
        `Mismatch-Percent_SD` = sd(mismatch_percent, na.rm = TRUE),
        
        # `Mismatch-Fold_Mean` = mean(frac_shift, na.rm = TRUE),
        `Mismatch-Fold_Median` = median(frac_shift, na.rm = TRUE),
        `Mismatch-Fold_SD` = sd(frac_shift, na.rm = TRUE)
    ) |>
    pivot_longer(cols = -Eco, names_to = "Metric", values_to = "Value") |>
    separate(Metric, into = c("Metric", "Stat"), sep = "_", extra = "merge") |>
    pivot_wider(names_from = c(Eco, Stat), values_from = Value) |>
    mutate(across(where(is.numeric), ~ round(.x, 2))) |>
    rename(
        # `Terrestrial Mean` = Ter_Mean,
        `Terrestrial Median` = Ter_Median,
        `Terrestrial SD` = Ter_SD,
        # `Marine Mean` = Mar_Mean,
        `Marine Median` = Mar_Median,
        `Marine SD` = Mar_SD
    )

# Reorder metrics to match your image
table_data$Metric <- gsub("-"," ",table_data$Metric)
metric_order <- c("Documented", "Modeled", "Mismatch", "Mismatch Percent", "Mismatch Fold")
table_data$Metric <- factor(table_data$Metric, levels = metric_order)

table_data_final <- table_data %>%
    select(Metric,
           `Terrestrial Median`, `Terrestrial SD`,
           `Marine Median`, `Marine SD`) %>%  # Reorder columns
    gt() %>%
    tab_spanner(
        label = "Marine",
        columns = c("Marine Median", "Marine SD")
    ) %>%
    tab_spanner(
        label = "Terrestrial",
        columns = c("Terrestrial Median", "Terrestrial SD")
    ) %>%
    cols_label(
        Metric = "",
        `Marine Median` = "Median",
        `Marine SD` = "SD",
        `Terrestrial Median` = "Median",
        `Terrestrial SD` = "SD"
    ) %>%
    fmt(
        columns = c(`Terrestrial SD`, `Marine SD`),
        fns = function(x) scientific(x, digits = 2)
    )

table_data_final
```


```{r testfaster3}
gt::gtsave(table_data_final,
           filename = here("Tables/Shift_stats.html"))
```


N lags, overshoots and misalignments
```{r summarizing5}

mod_data |> 
    group_by(lag_overshoot) |>
    summarise(Perc=length(lag_overshoot)/nrow(mod_data)*100) |>
    mutate(across(where(is.numeric), round, 1))

# N alignment and misalignments
mod_data |> 
    group_by(Eco, match_direction) |>
    summarise(Count = n(),
              Mean = mean(mismatch)) |> 
    mutate(Perc = (Count / sum(Count)) * 100) |> 
    mutate(across(where(is.numeric), round, 2))

# N alignment and lag_overshoot
mod_data |> 
    group_by(Eco, lag_overshoot) |>
    summarise(Count = n(),
              Mean = mean(mismatch)) |> 
    mutate(Perc = (Count / sum(Count)) * 100) |> 
    mutate(across(where(is.numeric), round, 2))

# N alignment and misalignments for the aggregated data
agg_data <- as_tibble(mod_data) |>
    select(ID, sp_name_std, class, Param, ShiftMod_raw, ShiftRate_raw, Eco, mismatch_trend)|>
    group_by(ID, class, Param, Eco) |>
    mutate(Nsps = length(unique(sp_name_std)),
           Eco = case_when(Eco == "Mar" ~ "Marine",
                           Eco == "Ter" ~ "Terrestrial")) |>
    filter(Nsps>3) |>
    summarise(Nsps = mean(Nsps),
              ShiftMod_sa = mean(ShiftMod_raw),
              ShiftRate_sa  = mean(ShiftRate_raw),
              ShiftMod_sa_sd = sd(ShiftMod_raw),
              ShiftRate_sa_sd = sd(ShiftRate_raw),
              mismatch_trend = mean(mismatch_trend)) |>
    plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) |> 
    mutate(match_direction = case_when(
        ShiftRate_sa >= 0 & ShiftMod_sa  >= 0 ~ "match",
        ShiftRate_sa <= 0 & ShiftMod_sa  <= 0 ~ "match",
        ShiftRate_sa <= 0 & ShiftMod_sa  >= 0 ~ "mismatch",
        ShiftRate_sa >= 0 & ShiftMod_sa  <= 0 ~ "mismatch",
        TRUE ~ "mismatch")) |>
    # tolerance for direction match (2x resolution of the environmental data)
    mutate(match_direction = case_when(
        Eco == "Terrestrial" & abs(mismatch_trend) < 2 ~ "match",
        Eco == "Marine" & abs(mismatch_trend) < 50 ~ "match",
        TRUE ~ match_direction))

agg_data  |>
    group_by(Eco, match_direction) |>
    summarise(Count = n()) |> 
    mutate(Perc = (Count / sum(Count)) * 100) |> 
    mutate(across(where(is.numeric), round, 2))

# what is the proportion of lags/overshoots after removing misalignments?
mod_data |> 
    filter(!lag_overshoot == "Misalignment") |>
    group_by(lag_overshoot) |>
    summarise(Count = n()) |> 
    mutate(Perc = (Count / sum(Count)) * 100) |> 
    mutate(across(where(is.numeric), round, 1))

# mean mistmach value by category
bioshifts |>
    group_by(Eco,lag_overshoot) |>
    filter(!is.infinite(mismatch_percent)) |>
    summarise(mean_mismatch = mean(mismatch)) |>
    mutate(across(where(is.numeric), round, 2))
```




# Check for correlations
## Scatters

```{r fig.width=10,fig.height=5}
# is Duration correlated with range shift mismatch?
ggplot(bioshifts, aes(Duration,mismatch))+
    geom_point()+
    theme_classic()+
    geom_smooth()+
    facet_wrap(.~Eco)

# is Duration correlated with env variability?
ggplot(bioshifts, aes(bvel_lat_error,Duration))+
    geom_point()+
    theme_classic()+
    geom_smooth()+
    facet_wrap(.~Eco)

# is Duration correlated Sampling method?
ggplot(bioshifts, aes(Category,Duration))+
    geom_boxplot()+
    theme_classic()+
    geom_smooth()+
    facet_wrap(.~Eco)

TukeyHSD(aov(Duration~Category, bioshifts |> filter(Eco == "Ter")))
TukeyHSD(aov(Duration~Category, bioshifts |> filter(Eco == "Mar")))

# is NPeriods correlated with range shift mismatch?
ggplot(bioshifts, aes((N_periodes),mismatch))+
    geom_point()+
    theme_classic()+
    geom_smooth()+
    facet_wrap(.~Eco)

# is env variability correlated with range shift mismatch?
ggplot(bioshifts, aes(bvel_lat_error,mismatch))+
    geom_point()+
    theme_classic()+
    geom_smooth()+
    facet_wrap(.~Eco)

# is NPeriods correlated with Duration?
ggplot(bioshifts, aes((N_periodes),Duration))+
    geom_point()+
    theme_classic()+
    geom_smooth()+
    facet_wrap(.~Eco)

# is NPeriods correlated with Category?
ggplot(bioshifts, aes(Category,(N_periodes)))+
    geom_point()+
    theme_classic()+
    geom_smooth()+
    facet_wrap(.~Eco)

# is Duration correlated with Category?
ggplot(bioshifts, aes(Category,Duration))+
    geom_boxplot()+
    theme_classic()+
    geom_smooth()+
    facet_wrap(.~Eco)

TukeyHSD(aov(Duration~Category, bioshifts |> filter(Eco == "Ter")))
TukeyHSD(aov(Duration~Category, bioshifts |> filter(Eco == "Mar")))

# is climate variability correlated with Category?
ggplot(bioshifts, aes(Category,log(bvel_lat_error)))+
    geom_boxplot(outliers = FALSE)+
    theme_classic()+
    geom_smooth()+
    facet_wrap(.~Eco)

TukeyHSD(aov(bvel_lat_error~Category, bioshifts |> filter(Eco == "Ter")))
TukeyHSD(aov(bvel_lat_error~Category, bioshifts |> filter(Eco == "Mar")))




```



Removed N_periods as it is highly correlated with Sampling_method

# VIF
## Terrestrial

```{r}
subdata_ter <- mod_data |>
    filter(Eco=="Ter") |>
    select("mismatch", "Boyce", "Study_duration", "Interannual_var", "Lat_extent", "Connectivity", "Sampling_method", "Life_form", "Locomotion_mode") |>
    na.omit()

mm_formula_ter <- "mismatch ~ Boyce + Study_duration + Interannual_var + Lat_extent + Connectivity + Sampling_method + Life_form + Locomotion_mode"

m1 <- glm(mm_formula_ter, 
          data = subdata_ter,
          family = Gamma(link = "log"))
summary(m1)
# locomotion mode is problematic >> remove

mm_formula_ter <- "mismatch ~ Boyce + Study_duration + Interannual_var + Lat_extent + Connectivity + Sampling_method + Life_form"

m1 <- glm(mm_formula_ter, 
          data = subdata_ter,
          family = Gamma(link = "log"))
summary(m1)
VIF(m1)
# high VIF value for life form
# try keeping locomotion mode and removing life form

mm_formula_ter <- "mismatch ~ Boyce + Study_duration + Interannual_var + Lat_extent + Connectivity + Sampling_method + Locomotion_mode"
m1 <- glm(mm_formula_ter, 
          data = subdata_ter,
          family = Gamma(link = "log"))
summary(m1)
VIF(m1) # Ok

```

## Marine

```{r}

subdata_mar <- mod_data |>
    filter(Eco=="Mar") |>
    select("mismatch", "Boyce", "Study_duration", "N_periods", "Interannual_var", "Connectivity", "Lat_extent", "Sampling_method", "Life_form", "Locomotion_mode") |>
    na.omit()

mm_formula_mar <- "mismatch ~ Boyce + Study_duration + Interannual_var + Lat_extent + Connectivity + Sampling_method + Life_form + Locomotion_mode"

m1 <- glm(mm_formula_mar, 
          data = subdata_mar,
          family = Gamma(link = "log"))
summary(m1)
VIF(m1) 
# high VIF value for locomotion mode
# try keeping locomotion mode and removing life form

mm_formula_mar <- "mismatch ~ Boyce + Study_duration + Interannual_var + Lat_extent + Connectivity + Sampling_method + Locomotion_mode"

m1 <- glm(mm_formula_mar,
          data = subdata_mar,
          family = Gamma(link = "log"))
summary(m1)
VIF(m1) # Ok

```

# Models

Set random and fixed effects
```{r}
# Random effects

## Nested 
RE_nest <- "(1 | Param/class/ID)"
## custom 
RE_cust <- "(1 | Random_ID_Param_Class)"

########################

# fixed effects
cont_vars <- c(
    "Boyce",
    "Study_duration",
    "Interannual_var",
    "Lat_extent",
    "Connectivity"
)
cat_vars <- c(
    "Life_form",
    "Sampling_method",
    "Locomotion_mode"
)

cont_vars_original <- c(
    "Boyce",
    "Duration",
    "bvel_lat_error", # log+1 transformed
    "LatExtentk", # log transformed
    "connectivity"
)
cat_vars_original <- c(
    "lifeform",
    "Category",
    "Locomotion_mode"
)

cont_vars_trans <- c(
    NA,
    NA,
    "log1p", # log+1 transformed
    "log", # log transformed
    NA
)

cont_vars_names <- c(
    "Model performance",
    "Study duration",
    "Variability in suitability",
    "Latitudinal extent",
    "Landscape connectivity"
)
cat_vars_names <- c(
    "Life form",
    "Sampling method",
    "Locomotion mode"
)
cont_vars_names_with_units <- c(
    "Model performance (Boyce index)",
    "Study duration (Years)",
    "Variability in suitability (°C)",
    "Latitudinal extent (Km)",
    "Landscape connectivity"
)
all_vars <- c(cont_vars, cat_vars, "ShiftRate")
all_vars_names <- c(cont_vars_names, cat_vars_names, "Documented Shift")
all_vars_original <- c(cont_vars_original, cat_vars_original, "ShiftRate")
all_vars_data <- data.frame(all_vars,all_vars_names,all_vars_original)

```


```{r}

if(is.null(params$shiftmodtype)){
    shiftmodtype = shiftmodtype_to_go
} else {
    shiftmodtype = params$shiftmodtype
}

# simple model
sm_fixed_eff <- "ShiftRate ~  ShiftMod"
sm_formula <- paste(sm_fixed_eff, RE_nest, sep = " + ")
sm_fixed_eff_raw <- "ShiftRate_raw ~  ShiftMod_raw"
sm_formula_raw <- paste(sm_fixed_eff_raw, RE_nest, sep = " + ")

# simple model * direction
sm_fixed_eff_dir <- "ShiftRate ~  ShiftMod * Shift_alignment"
sm_formula_dir <- paste(sm_fixed_eff_dir, RE_nest, sep = " + ")

# mismatch model
# set above during the VIF process

# mismatch model lag vs overshoot
mm_lagover_formula_ter <- mm_formula_ter
mm_lagover_formula_ter <- strsplit(mm_lagover_formula_ter, "~")[[1]]
response <- mm_lagover_formula_ter[1]
response <- gsub(" ","",response)
predictors <- mm_lagover_formula_ter[2]
predictors <- strsplit(predictors,"+",fixed = TRUE)[[1]]
predictors <- gsub(" ","",predictors)
predictors <- paste0(predictors,"*lag_overshoot")
predictors <- paste(predictors,collapse = " + ")

mm_lagover_formula_ter <- paste(response, predictors, sep = " ~ ")
mm_lagover_formula_ter <- paste(mm_lagover_formula_ter, "+", RE_nest)

mm_lagover_formula_mar <- mm_formula_mar
mm_lagover_formula_mar <- strsplit(mm_lagover_formula_mar, "~")[[1]]
response <- mm_lagover_formula_mar[1]
response <- gsub(" ","",response)
predictors <- mm_lagover_formula_mar[2]
predictors <- strsplit(predictors,"+",fixed = TRUE)[[1]]
predictors <- gsub(" ","",predictors)
predictors <- paste0(predictors,"*lag_overshoot")
predictors <- paste(predictors,collapse = " + ")

mm_lagover_formula_mar <- paste(response, predictors, sep = " ~ ")
mm_lagover_formula_mar <- paste(mm_lagover_formula_mar, "+", RE_nest)

```


## Simple model - species-level (Shift ~ ShiftMod)


### Terrestrial
```{r fig.width=4, fig.height=4}

subdata_ter <- mod_data |>
    filter(Eco=="Ter")

# length(unique(mod_data$Random_ID_Param_Class))

sm_ter <- glmmTMB(as.formula(sm_formula_raw),
                  data = subdata_ter)

r.squaredGLMM(sm_ter)

# summary(sm_ter)

# tab_model(sm_ter)
# 
# plot(effects::allEffects(sm_ter))

# R-squared
r.squaredGLMM(sm_ter)

# RMSE 
# Get model predictions
predicted <- predict(sm_ter)

# Get the observed values
observed <- subdata_ter$ShiftMod_raw

# Compute RMSE
rmse_val <- sqrt(mean((observed - predicted)^2))
rmse_val

# Normalized Metrics (Optional)
# You can compute NRMSE (normalized RMSE) = RMSE / mean(y), which allows comparisons across datasets.
# For proportion data (e.g., binomial with response in 0–1), RMSE below 0.1–0.2 is often considered good, but again it depends on your context.
rmse_val / mean(observed)
rmse_val / sd(observed)

# 1:1 test
fit.emt <- emtrends(sm_ter, ~1, var="ShiftMod_raw") 
fit.emt

fit_11_ter <- test(fit.emt, null=1)
fit_11_ter
```


### Marine

```{r fig.width=4, fig.height=4}

subdata_mar <- mod_data |>
    filter(Eco=="Mar")

# length(unique(subdata_mar$Random_ID_Param_Class))

sm_mar <- glmmTMB(as.formula(sm_formula_raw),
                  data = subdata_mar)

r.squaredGLMM(sm_mar)

# tab_model(sm_mar)
# 
# plot(effects::allEffects(sm_mar))

# R-squared
r.squaredGLMM(sm_mar)

# RMSE 
# Get model predictions
predicted <- predict(sm_mar)

# Get the observed values
observed <- subdata_mar$ShiftMod_raw

# Compute RMSE
rmse_val <- sqrt(mean((observed - predicted)^2))
rmse_val

# Normalized Metrics (Optional)
# You can compute NRMSE (normalized RMSE) = RMSE / mean(y), which allows comparisons across datasets.
# For proportion data (e.g., binomial with response in 0–1), RMSE below 0.1–0.2 is often considered good, but again it depends on your context.
rmse_val / mean(observed)
rmse_val / sd(observed)

# 1:1 test
fit.emt <- emtrends(sm_mar, ~1, var="ShiftMod_raw") 
fit.emt

fit_11_mar <- test(fit.emt, null=1)
fit_11_mar
```



### Save model results
```{r}

# Make table for the 1:1 test
rbind(fit_11_ter,
      fit_11_mar)  |> 
    select(-null) |>
    mutate(Eco = c("Ter","Mar"),
           p.value = case_when(
               p.value < 0.001 ~ "<0.001",
               p.value < 0.01  ~ "<0.01",
               p.value < 0.05  ~ "<0.05",
               TRUE            ~ sprintf("%.2f", p.value)  # Format to 3 decimal places
           ))|> 
    mutate(across(where(is.numeric), round, 2)) |> 
    gt()

# save regular table
mar_sm_res <- summary(sm_mar)
mar_sm_res <- data.frame(mar_sm_res$coefficients$cond)
mar_sm_res$Param <- rownames(mar_sm_res)
names(mar_sm_res) <- c("Estimate", "Std_Error", "z-value", "P-value", "Param")
# use p-value of the 1:1 test
mar_sm_res$`P-value`[2] <- fit_11_mar$p.value

ter_sm_res <- summary(sm_ter)
ter_sm_res <- data.frame(ter_sm_res$coefficients$cond)
ter_sm_res$Param <- rownames(ter_sm_res)
names(ter_sm_res) <- c("Estimate", "Std_Error", "z-value", "P-value", "Param")
# use p-value of the 1:1 test
ter_sm_res$`P-value`[2] <- fit_11_ter$p.value

res_glmm_simple_model <- rbind(
    data.frame(Eco = "Ter",
               ter_sm_res),
    data.frame(Eco = "Mar",
               mar_sm_res)
)


if(shiftmodtype=="95"){
    write.csv(res_glmm_simple_model, 
              here("Data/Output/res_glmm_sm95_all.csv"), 
              row.names = FALSE)
} else {
    write.csv(res_glmm_simple_model, 
              here("Data/Output/res_glmm_sm_all.csv"), 
              row.names = FALSE)
}

```


### Combined model output

```{r}
tab_model(sm_ter,sm_mar, 
          dv.labels = c("Terrestrial","Marine"),  
          show.icc = FALSE, 
          show.r2 = FALSE,
          show.ngroups = FALSE,
          show.re.var = FALSE)
```


## Simple model - Study-level (Shift ~ ShiftMod)

### Terrestrial
```{r fig.width=4, fig.height=4}

new_data <- as_tibble(bioshifts) |>
    filter(Eco=="Ter") |>
    select(ID, sp_name_std, class, Param, bvel_lat, bvel95_lat, ShiftRate, Eco)|>
    group_by(ID, class, Param, Eco) |>
    mutate(Nsps = length(unique(sp_name_std)),
           ShiftMod_raw = bvel_lat,
           ShiftMod_raw95 = bvel95_lat,
           ShiftRate_raw = ShiftRate) |>
    filter(Nsps>3) |>
    summarise(Nsps = mean(Nsps),
              ShiftMod = mean(ShiftMod_raw),
              ShiftMod95 = mean(ShiftMod_raw95),
              ShiftRate  = mean(ShiftRate_raw),
              ShiftMod95_sd = sd(ShiftMod_raw95),
              ShiftMod_sd = sd(ShiftMod_raw),
              ShiftRate_sd = sd(ShiftRate_raw)) |>
    plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) 

# Fit a deming model
sm_SA_deming_ter <- deming(
    formula = as.formula(sm_fixed_eff),
    xstd = ShiftMod_sd,
    ystd = ShiftRate_sd,
    weights = Nsps,
    data = new_data)

sm_SA_deming_ter

# Fit a GLMM (more appropriate because contains the random effects)

sm_SA_glmm_ter <- glmmTMB(
    as.formula(paste(sm_fixed_eff, "+ (1 | Param/Class)")),
    weights = Nsps, 
    new_data)

# tab_model(sm_SA_glmm_ter)
tmp <- summary(sm_SA_glmm_ter)

{
    plot(ShiftRate~ShiftMod, new_data)
    abline(h = 0, v = 0, lty = "dashed")
    abline(sm_SA_deming_ter, col = "blue")
    abline(b = tmp$coefficients$cond[2,1], a = tmp$coefficients$cond[1,1], col = "red")
}

# R-squared
r.squaredGLMM(sm_SA_glmm_ter)

# 1:1 test
fit.emt <- emtrends(sm_SA_glmm_ter, ~1, var="ShiftMod") 
fit.emt

fit_11_ter_SA <- test(fit.emt, null=1)
fit_11_ter_SA
```

No association between modeled shift velocity and observed shift velocity.

### Marine

```{r fig.width=4, fig.height=4}

new_data <- as_tibble(bioshifts) |>
    filter(Eco=="Mar") |>
    select(ID, sp_name_std, class, Param, bvel_lat, bvel95_lat, ShiftRate, Eco)|>
    group_by(ID, class, Param, Eco) |>
    mutate(Nsps = length(unique(sp_name_std)),
           ShiftMod_raw = bvel_lat,
           ShiftMod_raw95 = bvel95_lat,
           ShiftRate_raw = ShiftRate) |>
    filter(Nsps>3) |>
    summarise(Nsps = mean(Nsps),
              ShiftMod = mean(ShiftMod_raw),
              ShiftMod95 = mean(ShiftMod_raw95),
              ShiftRate  = mean(ShiftRate_raw),
              ShiftMod95_sd = sd(ShiftMod_raw95),
              ShiftMod_sd = sd(ShiftMod_raw),
              ShiftRate_sd = sd(ShiftRate_raw)) |>
    plyr::rename(replace = c(class = "Class"), warn_missing = FALSE) 

# new_data$ShiftMod_sd[which(new_data$ShiftMod_sd==0)] <- 0.0001
# new_data$ShiftRate_sd[which(new_data$ShiftMod_sd==0)] <- 0.0001
# new_data[is.na(new_data)] <- 0.0001

# Fit a deming model


# sm_SA_deming_mar <- deming(
#     formula = as.formula(sm_fixed_eff),
#     xstd = ShiftMod_sd,
#     ystd = ShiftRate_sd,
#     weights = Nsps,
#     data = new_data)
# 
# sm_SA_deming_mar

# Fit a GLMM (more appropriate because contains the random effects)

sm_SA_glmm_mar <- glmmTMB(
    as.formula(paste(sm_fixed_eff, "+ (1 | Param/Class)")),
    weights = Nsps, 
    new_data)

# tab_model(sm_SA_glmm_mar)
tmp <- summary(sm_SA_glmm_mar)

{
    plot(ShiftRate~ShiftMod, new_data)
    abline(h = 0, v = 0, lty = "dashed")
    # abline(sm_SA_deming_mar, col = "blue")
    abline(b = tmp$coefficients$cond[2,1], 
           a = tmp$coefficients$cond[1,1], col = "red")
}

# R-squared
r.squaredGLMM(sm_SA_glmm_mar)

# 1:1 test
fit.emt <- emtrends(sm_SA_glmm_mar, ~1, var="ShiftMod") 
fit.emt

fit_11_mar_SA <- test(fit.emt, null=1)
fit_11_mar_SA

```

Same here...


### Save model results
```{r}

# Make table for the 1:1 test
rbind(fit_11_ter_SA,
      fit_11_mar_SA)  |> 
    select(-null) |>
    mutate(Eco = c("Ter","Mar"),
           p.value = case_when(
               p.value < 0.001 ~ "<0.001",
               p.value < 0.01  ~ "<0.01",
               p.value < 0.05  ~ "<0.05",
               TRUE            ~ sprintf("%.2f", p.value)  # Format to 3 decimal places
           ))|> 
    mutate(across(where(is.numeric), round, 2)) |> 
    gt()

# save regular table
ter_sm_SA_glmm_res <- summary(sm_SA_glmm_ter)
ter_sm_SA_glmm_res <- data.frame(ter_sm_SA_glmm_res$coefficients$cond)
ter_sm_SA_glmm_res$Param <- rownames(ter_sm_SA_glmm_res)
names(ter_sm_SA_glmm_res) <- c("Estimate", "Std_Error", "z-value", "P-value", "Param")
# use p-value of the 1:1 test
ter_sm_SA_glmm_res$`P-value`[2] <- fit_11_ter$p.value

mar_sm_SA_glmm_res <- summary(sm_SA_glmm_mar)
mar_sm_SA_glmm_res <- data.frame(mar_sm_SA_glmm_res$coefficients$cond)
mar_sm_SA_glmm_res$Param <- rownames(mar_sm_SA_glmm_res)
names(mar_sm_SA_glmm_res) <- c("Estimate", "Std_Error", "z-value", "P-value", "Param")
# use p-value of the 1:1 test
mar_sm_SA_glmm_res$`P-value`[2] <- fit_11_mar$p.value


sm_SA_glmm_res <- rbind(data.frame(Eco = "Ter",ter_sm_SA_glmm_res),
                        data.frame(Eco = "Mar",mar_sm_SA_glmm_res))


if(shiftmodtype=="95"){
    write.csv(sm_SA_glmm_res, 
              here("Data/Output/res_sm_SA_glmm95_all.csv"), 
              row.names = FALSE)
} else {
    write.csv(sm_SA_glmm_res, 
              here("Data/Output/res_sm_SA_glmm_all.csv"), 
              row.names = FALSE)
}

```

### Combined model output

```{r}
tab_model(sm_SA_glmm_ter,sm_SA_glmm_mar, 
          dv.labels = c("Terrestrial","Marine"),
          show.icc = FALSE, 
          show.r2 = FALSE,
          show.ngroups = FALSE,
          show.re.var = FALSE)
```

## Direction model (Shift ~ ShiftMod * dir)

### Terrestrial
```{r fig.width=8, fig.height=4}
var_to_go <- strsplit(sm_formula_dir,"~|[+]|[/]|[|]|[*]")[[1]]
var_to_go <- gsub(" |[(]|[)]","",var_to_go)
var_to_go <- var_to_go[-which(var_to_go=="1")]

subdata_ter <- mod_data |>
    filter(Eco=="Ter") |>
    select(var_to_go) |>
    na.omit()

sm_dir_ter <- glmmTMB(as.formula(sm_formula_dir),
                      data = subdata_ter)

r.squaredGLMM(sm_dir_ter)

# tab_model(sm_dir_ter)

# plot(effects::allEffects(sm_dir_ter), main = "Terrestrial")
```


```{r fig.width=8, fig.height=4}
# test if this model is better that random 
sm_formula_dir_random <- gsub("Shift_alignment","Shuffled_Match_dir",sm_formula_dir)

cl <- makeCluster(detectCores()-2)
clusterExport(cl, c("subdata_ter", "glmmTMB", "sm_formula_dir_random", "r.squaredGLMM"))

null_r_squared <- pbreplicate(100, {
    subdata_ter$Shuffled_Match_dir  <- sample(subdata_ter$Shift_alignment)
    model_null <- glmmTMB(as.formula(sm_formula_dir_random),
                          data = subdata_ter)
    data.frame(r.squaredGLMM(model_null))
}, 
simplify = FALSE,
cl = cl)

stopCluster(cl)

null_r_squared <- rbindlist(null_r_squared)

observed_r_squared <- data.frame(r.squaredGLMM(sm_dir_ter))
mean(null_r_squared$R2m >= observed_r_squared$R2m)
mean(null_r_squared$R2c >= observed_r_squared$R2c)

plot_data <- rbind(data.frame(data = "Null", 
                              R2 = "Marginal R2",
                              value = null_r_squared$R2m),
                   data.frame(data = "Null", 
                              R2 = "Conditional R2",
                              value = null_r_squared$R2c),
                   data.frame(data = "Observed", 
                              R2 = "Marginal R2",
                              value = observed_r_squared$R2m),
                   data.frame(data = "Observed", 
                              R2 = "Conditional R2",
                              value = observed_r_squared$R2c))

ggplot(plot_data |>
           filter(data == "Null"), 
       aes(x = value)) +
    geom_histogram()+
    geom_vline(data = plot_data |>
                   filter(data == "Observed"),
               aes(xintercept = value, color = "red"))+
    theme_bw()+
    facet_wrap(.~R2, scales = "free")
```

Modeled shift velocity is positively related to observed shift velocity only when they shift in the same direction.

### Marine

```{r fig.width=8, fig.height=4}
var_to_go <- strsplit(sm_formula_dir,"~|[+]|[/]|[|]|[*]")[[1]]
var_to_go <- gsub(" |[(]|[)]","",var_to_go)
var_to_go <- var_to_go[-which(var_to_go=="1")]

subdata_mar <- mod_data |>
    filter(Eco=="Mar") |>
    select(var_to_go) |>
    na.omit()

sm_dir_mar <- glmmTMB(as.formula(sm_formula_dir),
                      data = subdata_mar)

r.squaredGLMM(sm_dir_mar)

# tab_model(sm_dir_mar)

# plot(effects::allEffects(sm_dir_mar), main = "Marine")
```


```{r fig.width=8, fig.height=4}
# test if this model is better that random 
sm_formula_dir_random <- gsub("Shift_alignment","Shuffled_Match_dir",sm_formula_dir)

cl <- makeCluster(detectCores()-2)
clusterExport(cl, c("subdata_mar", "glmmTMB", "sm_formula_dir_random", "r.squaredGLMM"))

null_r_squared <- pbreplicate(100, {
    subdata_mar$Shuffled_Match_dir  <- sample(subdata_mar$Shift_alignment)
    model_null <- glmmTMB(as.formula(sm_formula_dir_random),
                          data = subdata_mar)
    data.frame(r.squaredGLMM(model_null))
}, 
simplify = FALSE,
cl = cl)

stopCluster(cl)

null_r_squared <- rbindlist(null_r_squared)

observed_r_squared <- data.frame(r.squaredGLMM(sm_dir_mar))
mean(null_r_squared$R2m >= observed_r_squared$R2m)
mean(null_r_squared$R2c >= observed_r_squared$R2c)

plot_data <- rbind(data.frame(data = "Null", 
                              R2 = "Marginal R2",
                              value = null_r_squared$R2m),
                   data.frame(data = "Null", 
                              R2 = "Conditional R2",
                              value = null_r_squared$R2c),
                   data.frame(data = "Observed", 
                              R2 = "Marginal R2",
                              value = observed_r_squared$R2m),
                   data.frame(data = "Observed", 
                              R2 = "Conditional R2",
                              value = observed_r_squared$R2c))

ggplot(plot_data |>
           filter(data == "Null"), 
       aes(x = value)) +
    geom_histogram()+
    geom_vline(data = plot_data |>
                   filter(data == "Observed"),
               aes(xintercept = value, color = "red"))+
    theme_bw()+
    facet_wrap(.~R2, scales = "free")

```

Modeled shift velocity is positively related to observed shift velocity only when they shift in the same direction.

### Combined model output

```{r fig.height=6,fig.width=7}

sm_dir_ter_eff <- ggeffects::ggeffect(sm_dir_ter, 
                                      terms = c("ShiftMod","Shift_alignment"))
sm_dir_mar_eff <- ggeffects::ggeffect(sm_dir_mar, 
                                      terms = c("ShiftMod","Shift_alignment"))

# unscale vars
params_xy <- bioshifts |>
    select(ShiftRate, bvel_lat) |>
    summarise(ShiftRate_mean = mean(ShiftRate,na.rm = TRUE),
              bvel_lat_mean = mean(bvel_lat,na.rm = TRUE),
              ShiftRate_sd = sd(ShiftRate,na.rm = TRUE),
              bvel_lat_sd = sd(bvel_lat,na.rm = TRUE))


# make plot terrestrial
sm_dir_ter_eff$x <- sm_dir_ter_eff$x * params_xy$bvel_lat_sd + params_xy$bvel_lat_mean

sm_dir_ter_eff$predicted <- sm_dir_ter_eff$predicted * params_xy$ShiftRate_sd + params_xy$ShiftRate_mean
sm_dir_ter_eff$conf.low <-sm_dir_ter_eff$conf.low * params_xy$ShiftRate_sd + params_xy$ShiftRate_mean
sm_dir_ter_eff$conf.high <-sm_dir_ter_eff$conf.high * params_xy$ShiftRate_sd + params_xy$ShiftRate_mean

sm_dir_ter_eff_plot <-
    ggplot(sm_dir_ter_eff, aes(x = x, y = predicted))+
    geom_line(size = 1.5) +
    geom_ribbon(aes(ymin=conf.low, ymax=conf.high), alpha = .3) +
    theme_classic(base_size = 10)+
    labs(x = "Modeled latitudinal shift (km/yr)", 
         y = "Predicted latitudinal shift rate (km/yr)",
         title = "Terrestrial",
         fill = NULL, 
         color = NULL)+
    theme(aspect.ratio=1)+
    facet_wrap(.~group, scales = "free_x")

# make plot marine
sm_dir_mar_eff$x <- sm_dir_mar_eff$x * params_xy$bvel_lat_sd + params_xy$bvel_lat_mean

sm_dir_mar_eff$predicted <- sm_dir_mar_eff$predicted * params_xy$ShiftRate_sd + params_xy$ShiftRate_mean
sm_dir_mar_eff$conf.low <-sm_dir_mar_eff$conf.low * params_xy$ShiftRate_sd + params_xy$ShiftRate_mean
sm_dir_mar_eff$conf.high <-sm_dir_mar_eff$conf.high * params_xy$ShiftRate_sd + params_xy$ShiftRate_mean

sm_dir_mar_eff_plot <-
    ggplot(sm_dir_mar_eff, aes(x = x, y = predicted))+
    geom_line(size = 1.5) +
    geom_ribbon(aes(ymin=conf.low, ymax=conf.high), alpha = .3) +
    theme_classic(base_size = 10)+
    labs(x = "Modeled latitudinal shift (km/yr)", 
         y = "Documented latitudinal shift rate (km/yr)",
         title = "Marine",
         fill = NULL, 
         color = NULL)+
    theme(aspect.ratio=1)+
    facet_wrap(.~group, scales = "free_x")


ggarrange(sm_dir_ter_eff_plot,
          sm_dir_mar_eff_plot,
          labels = "AUTO",
          font.label = list(size = 10),
          align = "hv",
          ncol = 1)

```


```{r}
tab_model(sm_dir_ter,sm_dir_mar, 
          dv.labels = c("Terrestrial","Marine"),  
          show.icc = FALSE, 
          show.r2 = FALSE,
          show.ngroups = FALSE,
          show.re.var = FALSE,
          show.obs = FALSE)
```

### Chi-squared test for direction frequency
```{r}

mod_data  |>
    group_by(Eco, match_direction) |>
    summarise(N = length(match_direction)) |>
    mutate(Percent = round((N / sum(N))*100,1))

# Terrestrial
dir_table <- mod_data |>
    filter(Eco == "Ter") |> 
    group_by(match_direction) |>
    count() |>
    na.omit()

dir_table
chisq.test(dir_table$n)

# Marine
dir_table <- mod_data |>
    filter(Eco == "Mar") |> 
    group_by(match_direction) |>
    count() |>
    na.omit()

dir_table
chisq.test(dir_table$n)

```


## Binomial model

### Terrestrial
```{r}

subdata_ter <- mod_data |>
    filter(Eco=="Ter") |>
    select(mismatch_binary,Param,class,ID) |>
    na.omit()

# fit model
# models dont converge with (1 | Param/class/ID) 
mm01_ter <- glmmTMB(mismatch_binary ~ 1 + (1 | class/ID),
                    family = binomial,
                    data = subdata_ter,
                    control = glmmTMBControl(
                        optimizer = optim,
                        optArgs = list(
                            method = "BFGS")
                    )
)

# tab_model(mm01_ter, transform = NULL, string.est = "Estimate")

# R-squared
r.squaredGLMM(mm01_ter)

performance::r2_nakagawa(mm01_ter)

```

### Marine
```{r}

subdata_mar <- mod_data |>
    filter(Eco=="Mar") |>
    select(mismatch_binary,Param,class,ID) |>
    na.omit()

# fit model
# models dont converge with (1 | Param/class/ID) 
mm01_mar <- glmmTMB(mismatch_binary ~ 1 + (1 | Param/class/ID),
                    family = binomial,
                    data = subdata_mar,
                    control = glmmTMBControl(
                        optimizer = optim,
                        optArgs = list(
                            method = "BFGS")
                    )
)
summary(mm01_mar)
# tab_model(mm01_mar, transform = NULL, string.est = "Estimate")

# R-squared
r.squaredGLMM(mm01_mar)

performance::r2_nakagawa(mm01_mar)

```

<!-- ### Combined model output -->

```{r}
tab_model(mm01_ter, mm01_mar, 
          transform = NULL, string.est = "Estimate", string.p = "P-value",
          dv.labels = c("Terrestrial","Marine"),  
          show.icc = FALSE, 
          show.r2 = FALSE,
          show.ngroups = FALSE,
          show.re.var = FALSE,
          show.obs = FALSE)

# Back-transform the intercept to get the probability
prob_success_ter <- exp(coeffs(mm01_ter)) / (1 + exp(coeffs(mm01_ter)))
prob_success_ter
prob_success_mar <- exp(coeffs(mm01_mar)) / (1 + exp(coeffs(mm01_mar)))
prob_success_mar
```

### Terrestrial - aggregated
```{r}

subdata_ter <- agg_data |>
    filter(Eco=="Terrestrial") |>
    select(match_direction,Param,Class,ID) |>
    mutate(match_direction = case_when(match_direction == "match" ~ 1,
                                       match_direction == "mismatch" ~ 0)) |>
    na.omit()

# fit model
# models dont converge with (1 | Param/class/ID) 
mm01_agg_ter <- glmmTMB(match_direction ~ 1 + (1 | Param),
                        family = binomial,
                        data = subdata_ter,
                        # control = glmmTMBControl(
                        #     optimizer = optim,
                        #     optArgs = list(
                        #         method = "BFGS")
                        # )
)

# tab_model(mm01_agg_ter, transform = NULL, string.est = "Estimate")

# R-squared
r.squaredGLMM(mm01_agg_ter)

performance::r2(mm01_agg_ter)

```

### Marine - aggregated
```{r}

subdata_mar <- agg_data |>
    filter(Eco=="Marine") |>
    select(match_direction,Param,Class,ID) |>
    mutate(match_direction = case_when(match_direction == "match" ~ 1,
                                       match_direction == "mismatch" ~ 0)) |>
    na.omit()

# fit model
# models dont converge with (1 | Param/class/ID) 
mm01_agg_mar <- glmmTMB(match_direction ~ 1 + (1 | Param),
                        family = binomial,
                        data = subdata_mar,
                        control = glmmTMBControl(
                            optimizer = optim,
                            optArgs = list(
                                method = "BFGS")
                        )
)

# tab_model(mm01_agg_mar, transform = NULL, string.est = "Estimate")

# R-squared
r.squaredGLMM(mm01_agg_mar)

performance::r2(mm01_agg_mar)

```

<!-- ### Combined model output -->

```{r}
tab_model(mm01_agg_ter, mm01_agg_mar, 
          transform = NULL, string.est = "Estimate", string.p = "P-value",
          dv.labels = c("Terrestrial","Marine"),  
          show.icc = FALSE, 
          show.r2 = FALSE,
          show.ngroups = FALSE,
          show.re.var = FALSE,
          show.obs = FALSE)

# Back-transform the intercept to get the probability
prob_success_ter <- exp(coeffs(mm01_agg_ter)) / (1 + exp(coeffs(mm01_agg_ter)))
prob_success_ter
prob_success_mar <- exp(coeffs(mm01_agg_mar)) / (1 + exp(coeffs(mm01_agg_mar)))
prob_success_mar
```

## Lag vs overshoots between documented and modeled shifts
### Terrestrial

#### Fit model

```{r}

y <- strsplit(mm_lagover_formula_ter," ")[[1]][1]

var_to_go <- strsplit(mm_lagover_formula_ter,"~|[+]|[/]|[|]|[*]")[[1]]
var_to_go <- gsub(" |[(]|[)]","",var_to_go)
var_to_go <- var_to_go[-which(var_to_go=="1")]
var_to_go <- unique(var_to_go)

subdata_ter <- mod_data |>
    filter(Eco=="Ter") |> 
    select(var_to_go) |>
    na.omit()

mm_lag_over_ter <- glmmTMB(as.formula(mm_lagover_formula_ter),
                           family = Gamma(link = "log"),
                           data = subdata_ter,
                           # control = glmmTMBControl(
                           #     optimizer = optim,
                           #     optArgs = list(
                           #         method = "BFGS",
                           #         rel.tol = 1e-10,   # Reduce relative tolerance
                           #         maxit = 10000      # Increase max iterations
                           #     ))
)

# tab_model(mm_lag_over_ter, transform = NULL, string.est = "Estimate")

# performance::check_model(mm_lag_over_ter)

# R-squared
r.squaredGLMM(mm_lag_over_ter)

performance::r2(mm_lag_over_ter)

# Decompose variance explained by individual predictors
# drop1(mm_lag_over_ter, test = "Chisq")

# RMSE 
# Get model predictions
predicted <- predict(mm_lag_over_ter)

# Get the observed values
observed <- subdata_ter$mismatch

# Compute RMSE
rmse_val <- sqrt(mean((observed - predicted)^2))
rmse_val

# Normalized Metrics (Optional)
# You can compute NRMSE (normalized RMSE) = RMSE / mean(y), which allows comparisons across datasets.
# For proportion data (e.g., binomial with response in 0–1), RMSE below 0.1–0.2 is often considered good, but again it depends on your context.
rmse_val/mean(observed)
rmse_val / sd(observed)

# Compute mean bias
mean_bias <- mean(predicted - observed)
mean_bias

plot(effect(term = c("lag_overshoot","Locomotion_mode"),mod = mm_lag_over_ter))

```

#### Get effects for continuous variables
##### overall effects
```{r }


mm_lag_over_ter_eff <- lapply(cont_vars, function(x){
    if(any(grepl(x,var_to_go))){
        emtrends(object = mm_lag_over_ter, 
                 ~1,
                 var = x)
    }
})
names(mm_lag_over_ter_eff) <- cont_vars_names

test <- sapply(mm_lag_over_ter_eff,is.null)
if(any(test)){
    mm_lag_over_ter_eff <- mm_lag_over_ter_eff[-which(test)]
}
test <- lapply(mm_lag_over_ter_eff, function(x){
    x = data.frame(x)
    names(x)[1] <- "Var"
    names(x)[2] <- "Estimate"
    # Compute Wald z and p-value manually
    z <- x[,2] / x$SE
    p <- 2 * pnorm(-abs(z))
    
    data.frame(x, z = z, p.value = p)
})
test <- rbindlist(test)
test$Var <- cont_vars_names
test <- data.frame(test)
test[,-1] <- round(test[,-1],3)
test$CI <- paste(test$asymp.LCL,test$asymp.UCL,sep = " - ")
test_ter <- test[,c(1,2,3,9,8)]
test_ter <- test_ter %>%
    mutate(p.value = case_when(
        p.value < 0.001 ~ "<0.001",
        p.value < 0.01  ~ "<0.01",
        p.value < 0.05  ~ "<0.05",
        TRUE            ~ sprintf("%.2f", p.value)  # Format to 3 decimal places
    ))

```

##### interactions
```{r }

mm_lag_over_ter_eff <- lapply(cont_vars, function(x){
    if(any(grepl(x,var_to_go))){
        emtrends(object = mm_lag_over_ter, 
                 specs =  "lag_overshoot",
                 var = x)
    }
})
names(mm_lag_over_ter_eff) <- cont_vars_names

test <- sapply(mm_lag_over_ter_eff,is.null)
if(any(test)){
    mm_lag_over_ter_eff <- mm_lag_over_ter_eff[-which(test)]
}
mm_lag_over_ter_eff

```



#### Get emmeans for categorical variables
```{r }

mm_lag_over_ter_emm <- lapply(cat_vars, function(x){
    if(any(grepl(x,var_to_go))){
        emmeans(object = mm_lag_over_ter, 
                specs = c(x,"lag_overshoot"),
                type = "response")
    }
})
names(mm_lag_over_ter_emm) <- cat_vars_names

test <- sapply(mm_lag_over_ter_emm,is.null)
if(any(test)){
    mm_lag_over_ter_emm <- mm_lag_over_ter_emm[-which(test)]
}
mm_lag_over_ter_emm

mm_lag_over_ter_cont <- lapply(mm_lag_over_ter_emm, 
                               contrast, 
                               method = "eff",
                               type = "response")
names(mm_lag_over_ter_cont) <- names(mm_lag_over_ter_cont)
mm_lag_over_ter_cont

mm_lag_over_ter_cont_pairs <- lapply(mm_lag_over_ter_emm,
                                     contrast, 
                                     method = "pairwise",
                                     type = "response",
                                     by = "lag_overshoot",
                                     adjust = "none",
                                     ratios = FALSE
)
names(mm_lag_over_ter_cont_pairs) <- names(mm_lag_over_ter_cont)
mm_lag_over_ter_cont_pairs

```


### Marine

#### Fit model
```{r}

y <- strsplit(mm_lagover_formula_mar," ")[[1]][1]

var_to_go <- strsplit(mm_lagover_formula_mar,"~|[+]|[/]|[|]|[*]")[[1]]
var_to_go <- gsub(" |[(]|[)]","",var_to_go)
var_to_go <- var_to_go[-which(var_to_go=="1")]
var_to_go <- unique(var_to_go)

subdata_mar <- mod_data |>
    filter(Eco=="Mar") |> 
    select(var_to_go) |>
    na.omit()

mm_lag_over_mar <- glmmTMB(as.formula(mm_lagover_formula_mar),
                           family = Gamma(link = "log"),
                           data = subdata_mar)

# tab_model(mm_lag_over_mar, transform = NULL, string.est = "Estimate")

# R-squared
r.squaredGLMM(mm_lag_over_mar)

performance::r2(mm_lag_over_mar)

# RMSE 
# Get model predictions
predicted <- predict(mm_lag_over_mar)

# Get the observed values
observed <- subdata_mar$mismatch

# Compute RMSE
rmse_val <- sqrt(mean((observed - predicted)^2))
rmse_val

# Normalized Metrics (Optional)
# You can compute NRMSE (normalized RMSE) = RMSE / mean(y), which allows comparisons across datasets.
# For proportion data (e.g., binomial with response in 0–1), RMSE below 0.1–0.2 is often considered good, but again it depends on your context.
rmse_val/mean(observed)
rmse_val / sd(observed)

# Compute mean bias
mean_bias <- mean(predicted - observed)
mean_bias
```

#### Get effects for continuous variables
##### overall effects
```{r }

mm_lag_over_mar_eff <- lapply(cont_vars, function(x){
    if(any(grepl(x,var_to_go))){
        emtrends(object = mm_lag_over_mar, 
                 ~1,
                 var = x)
    }
})
names(mm_lag_over_mar_eff) <- cont_vars_names

test <- sapply(mm_lag_over_mar_eff,is.null)
if(any(test)){
    mm_lag_over_mar_eff <- mm_lag_over_mar_eff[-which(test)]
}
test <- lapply(mm_lag_over_mar_eff, function(x){
    x = data.frame(x)
    names(x)[1] <- "Var"
    names(x)[2] <- "Estimate"
    # Compute Wald z and p-value manually
    z <- x[,2] / x$SE
    p <- 2 * pnorm(-abs(z))
    
    data.frame(x, z = z, p.value = p)
})
test <- rbindlist(test)
test$Var <- cont_vars_names
test <- data.frame(test)
test[,-1] <- round(test[,-1],3)
test$CI <- paste(test$asymp.LCL,test$asymp.UCL,sep = " - ")
test_mar<- test[,c(1,2,3,9,8)]
test_mar <- test_mar %>%
    mutate(p.value = case_when(
        p.value < 0.001 ~ "<0.001",
        p.value < 0.01  ~ "<0.01",
        p.value < 0.05  ~ "<0.05",
        TRUE            ~ sprintf("%.2f", p.value)  # Format to 3 decimal places
    ))


```

##### interactions
```{r }

mm_lag_over_mar_eff <- lapply(cont_vars, function(x){
    if(any(grepl(x,var_to_go))){
        emtrends(object = mm_lag_over_mar, 
                 specs =  "lag_overshoot",
                 var = x)
    }
})
names(mm_lag_over_mar_eff) <- cont_vars_names

test <- sapply(mm_lag_over_mar_eff,is.null)
if(any(test)){
    mm_lag_over_mar_eff <- mm_lag_over_mar_eff[-which(test)]
}
mm_lag_over_mar_eff

```

#### Get emmeans for categorical variables
```{r }

mm_lag_over_mar_emm <- lapply(cat_vars, function(x){
    if(any(grepl(x,var_to_go))){
        emmeans(object = mm_lag_over_mar, 
                specs = c(x,"lag_overshoot"),
                type = "response")
    }
})
names(mm_lag_over_mar_emm) <- cat_vars_names

test <- sapply(mm_lag_over_mar_emm,is.null)
if(any(test)){
    mm_lag_over_mar_emm <- mm_lag_over_mar_emm[-which(test)]
}
mm_lag_over_mar_emm

mm_lag_over_mar_cont <- lapply(mm_lag_over_mar_emm, 
                               contrast, 
                               method = "eff",
                               type = "response")
names(mm_lag_over_mar_cont) <- names(mm_lag_over_mar_cont)
mm_lag_over_mar_cont

mm_lag_over_mar_cont_pairs <- lapply(mm_lag_over_mar_emm,
                                     contrast, 
                                     method = "pairwise",
                                     type = "response",
                                     by = "lag_overshoot",
                                     adjust = "none",
                                     ratios = FALSE
)
names(mm_lag_over_mar_cont_pairs) <- names(mm_lag_over_mar_cont)
mm_lag_over_mar_cont_pairs

```

### Combined model output

```{r}
tab_model(mm_lag_over_ter, mm_lag_over_mar, 
          transform = NULL, string.est = "Estimate", string.p = "P-value",
          dv.labels = c("Terrestrial","Marine"),  
          show.icc = FALSE, 
          show.r2 = FALSE,
          show.ngroups = FALSE,
          show.re.var = FALSE,
          show.obs = FALSE)
```


### Combined emmeans output
#### interaction
```{r}

emmeans_cont_ter <- lapply(mm_lag_over_ter_eff, function(x) {
    tmp <- data.frame(x)
    var_name <- gsub(".trend","",names(tmp[2]))
    tmp[,1] <- paste("\u2003",tmp[,1],sep = "")
    tmp$CI <- paste(round(tmp$asymp.LCL,2),round(tmp$asymp.UCL,2),sep = " – ")
    tmp$z = tmp[,2] / tmp$SE
    tmp$`P-value` = 2 * pnorm(-abs(tmp$z))
    tmp <- tmp[,c(1,2,7,9)]
    names(tmp) <- c("Predictors", "Estimate","CI","P-value")
    tmp$Estimate <- round(tmp$Estimate,2)
    tmp <- tmp %>%
        mutate(`P-value` = case_when(
            `P-value` < 0.001 ~ "<0.001",
            `P-value` < 0.01  ~ "<0.01",
            `P-value` < 0.05  ~ "<0.05",
            TRUE            ~ sprintf("%.2f", `P-value`)  # Format to 3 decimal places
        ))
    
    tmp_2 <- tmp[1,]
    tmp_2[1,] <- ""
    tmp_2[,1] <- var_name
    
    return(rbind(tmp_2,tmp))
})
emmeans_cont_ter <- rbindlist(emmeans_cont_ter)


emmeans_cat_ter <- lapply(mm_lag_over_ter_cont_pairs, function(x) {
    tmp <- data.frame(x)
    var_name <- names(x@misc$orig.grid)[1]
    var_name <- gsub("_"," ",var_name)
    ps <- tmp
    tmp <- data.frame(confint(x))
    tmp$`P-value` = ps$p.value
    tmp$CI <- paste(round(tmp$asymp.LCL,2),round(tmp$asymp.UCL,2),sep = " – ")
    
    tmp[,1] <- gsub("(","",tmp[,1],fixed = TRUE)
    tmp[,1] <- gsub(")","",tmp[,1],fixed = TRUE)
    tmp[,1] <- gsub(" - "," – ",tmp[,1],fixed = TRUE)
    tmp[,1] <- paste("\u2003\u2003",tmp[,1],sep = "")
    tmp[,2] <- paste("\u2003",tmp[,2],sep = "")
    
    tmp <- tmp[,c(1,2,3,9,8)]
    
    levels_lag <- unique(tmp$lag_overshoot)
    
    levels_lag_res <- lapply(levels_lag, function(i){
        tmp_tmp <- tmp[which(tmp$lag_overshoot == i),]
        tmp_tmp <- tmp_tmp[,-2]
        names(tmp_tmp) <- c("Predictors", "Estimate","CI","P-value")
        tmp_tmp$Estimate <- round(tmp_tmp$Estimate,2)
        tmp_tmp <- tmp_tmp %>%
            mutate(`P-value` = case_when(
                `P-value` < 0.001 ~ "<0.001",
                `P-value` < 0.01  ~ "<0.01",
                `P-value` < 0.05  ~ "<0.05",
                TRUE            ~ sprintf("%.2f", `P-value`)  # Format to 3 decimal places
            ))
        
        tmp_tmp_2 <- tmp_tmp[1,]
        tmp_tmp_2[1,] <- ""
        tmp_tmp_2[,1] <- i
        
        return(rbind(tmp_tmp_2,tmp_tmp))
    }
    )
    
    levels_lag_res <- rbindlist(levels_lag_res)
    
    tmp_tmp_3 <- levels_lag_res[1,]
    tmp_tmp_3[1,] <- ""
    tmp_tmp_3[,1] <- var_name
    
    levels_lag_res <- rbind(tmp_tmp_3,levels_lag_res)
    
    return(levels_lag_res)
})
emmeans_cat_ter <- rbindlist(emmeans_cat_ter)


```


```{r}
emmeans_cont_mar <- lapply(mm_lag_over_mar_eff, function(x) {
    tmp <- data.frame(x)
    var_name <- gsub(".trend","",names(tmp[2]))
    tmp[,1] <- paste("\u2003",tmp[,1],sep = "")
    tmp$CI <- paste(round(tmp$asymp.LCL,2),round(tmp$asymp.UCL,2),sep = " – ")
    tmp$z = tmp[,2] / tmp$SE
    tmp$`P-value` = 2 * pnorm(-abs(tmp$z))
    tmp <- tmp[,c(1,2,7,9)]
    names(tmp) <- c("Predictors", "Estimate","CI","P-value")
    tmp$Estimate <- round(tmp$Estimate,2)
    tmp <- tmp %>%
        mutate(`P-value` = case_when(
            `P-value` < 0.001 ~ "<0.001",
            `P-value` < 0.01  ~ "<0.01",
            `P-value` < 0.05  ~ "<0.05",
            TRUE            ~ sprintf("%.2f", `P-value`)  # Format to 3 decimal places
        ))
    
    tmp_2 <- tmp[1,]
    tmp_2[1,] <- ""
    tmp_2[,1] <- var_name
    
    return(rbind(tmp_2,tmp))
})
emmeans_cont_mar <- rbindlist(emmeans_cont_mar)


emmeans_cat_mar <- lapply(mm_lag_over_mar_cont_pairs, function(x) {
    tmp <- data.frame(x)
    var_name <- names(x@misc$orig.grid)[1]
    var_name <- gsub("_"," ",var_name)
    ps <- tmp
    tmp <- data.frame(confint(x))
    tmp$`P-value` = ps$p.value
    tmp$CI <- paste(round(tmp$asymp.LCL,2),round(tmp$asymp.UCL,2),sep = " – ")
    
    tmp[,1] <- gsub("(","",tmp[,1],fixed = TRUE)
    tmp[,1] <- gsub(")","",tmp[,1],fixed = TRUE)
    tmp[,1] <- gsub(" - "," – ",tmp[,1],fixed = TRUE)
    tmp[,1] <- paste("\u2003\u2003",tmp[,1],sep = "")
    tmp[,2] <- paste("\u2003",tmp[,2],sep = "")
    
    tmp <- tmp[,c(1,2,3,9,8)]
    
    levels_lag <- unique(tmp$lag_overshoot)
    
    levels_lag_res <- lapply(levels_lag, function(i){
        tmp_tmp <- tmp[which(tmp$lag_overshoot == i),]
        tmp_tmp <- tmp_tmp[,-2]
        names(tmp_tmp) <- c("Predictors", "Estimate","CI","P-value")
        tmp_tmp$Estimate <- round(tmp_tmp$Estimate,2)
        tmp_tmp <- tmp_tmp %>%
            mutate(`P-value` = case_when(
                `P-value` < 0.001 ~ "<0.001",
                `P-value` < 0.01  ~ "<0.01",
                `P-value` < 0.05  ~ "<0.05",
                TRUE            ~ sprintf("%.2f", `P-value`)  # Format to 3 decimal places
            ))
        
        tmp_tmp_2 <- tmp_tmp[1,]
        tmp_tmp_2[1,] <- ""
        tmp_tmp_2[,1] <- i
        
        return(rbind(tmp_tmp_2,tmp_tmp))
    }
    )
    
    levels_lag_res <- rbindlist(levels_lag_res)
    
    tmp_tmp_3 <- levels_lag_res[1,]
    tmp_tmp_3[1,] <- ""
    tmp_tmp_3[,1] <- var_name
    
    levels_lag_res <- rbind(tmp_tmp_3,levels_lag_res)
    
    return(levels_lag_res)
})
emmeans_cat_mar <- rbindlist(emmeans_cat_mar)


```


```{r}
emmeans_res_ter <- rbind(emmeans_cont_ter,emmeans_cat_ter)
names(emmeans_res_ter)[-1] <- paste(names(emmeans_res_ter)[-1],".x", sep = "")

emmeans_res_mar <- rbind(emmeans_cont_mar,emmeans_cat_mar)
names(emmeans_res_mar)[-1] <- paste(names(emmeans_res_mar)[-1],".y", sep = "")

merged_df <- cbind(emmeans_res_ter, emmeans_res_mar[,-1])

```


```{r}

table_lagover <- merged_df %>%
    gt() %>%
    tab_spanner(
        label = "Terrestrial",
        columns = c("Estimate.x", "CI.x", "P-value.x")
    ) %>%
    tab_spanner(
        label = "Marine",
        columns = c("Estimate.y", "CI.y", "P-value.y")
    ) %>%
    cols_label(
        Estimate.x = "Estimate",
        CI.x = "CI",
        `P-value.x` = "P-value",
        Estimate.y = "Estimate",
        CI.y = "CI",
        `P-value.y` = "P-value"
    ) %>%
    fmt_number(
        columns = c(Estimate.x,Estimate.y),
        decimals = 5
    )

table_lagover

gt::gtsave(table_lagover,
           filename = here("Tables/lagovershot_glmmtmb.html"))
```

#### overall effect
```{r}

merged_df2 <- cbind(test_ter,test_mar[,-1])
names(merged_df2) <- make.unique(names(merged_df2))

table_overall <- merged_df2 %>%
    gt() %>%
    tab_spanner(
        label = "Terrestrial",
        columns = c("Estimate", "SE", "CI", "p.value")
    ) %>%
    tab_spanner(
        label = "Marine",
        columns =  c("Estimate.1", "SE.1", "CI.1", "p.value.1")
    ) %>%
    cols_label(
        Estimate.1 = "Estimate",
        SE.1 = "SE",
        CI.1 = "CI",
        p.value = "P-value",
        p.value.1 = "P-value",
    ) 

table_overall

gt::gtsave(table_lagover,
           filename = here("Tables/overall_glmmtmb.html"))
```

### Make legend
```{r}

# Create the scatter plot with quadrants and diagonal splits
lg <- ggplot() +
    # Define Quadrant I (x > 0, y > 0)
    geom_polygon(data = data.frame(x = c(0, Inf, Inf), y = c(0, Inf, 0)),
                 aes(x, y), fill="#228d8d99") +  # x > y
    geom_polygon(data = data.frame(x = c(0, 0, Inf), y = c(0, Inf, Inf)),
                 aes(x, y), fill="#440d5499") +  # x < y
    
    # Define Quadrant II (x < 0, y > 0)
    geom_rect(aes(xmin=-Inf, xmax=0, ymin=0, ymax=Inf), fill="#fde72599") +
    
    # Define Quadrant III (x < 0, y < 0) - Split by x > y and x < y
    geom_polygon(data = data.frame(x = c(-Inf, 0, -Inf), y = c(-Inf, 0, 0)),
                 aes(x, y), fill="#228d8d99") +  # x > y
    geom_polygon(data = data.frame(x = c(0, 0, -Inf), y = c(-Inf, 0, -Inf)),
                 aes(x, y), fill="#440d5499") +  # x < y
    
    # Define Quadrant IV (x > 0, y < 0)
    geom_rect(aes(xmin=0, xmax=Inf, ymin=-Inf, ymax=0), fill="#fde72599") +
    
    # Add axis lines
    geom_hline(yintercept=0, color="black") +
    geom_vline(xintercept=0, color="black") +
    geom_abline(slope = 1, intercept = 0, color="black", linetype = "dashed") +
    
    # change axis
    scale_x_continuous(breaks = 0)+
    scale_y_continuous(breaks = 0)+
    
    labs(x = "Modeled\nrange shift", y = "Documented\nrange shift")+
    theme_classic(base_size = 8)+
    theme(
        # axis.text=element_blank(),
        axis.text.y = element_text(hjust = -5),
        axis.text.x = element_text(vjust = 2),
        axis.ticks=element_blank(),
        aspect.ratio=1,
        axis.line = element_line(arrow = arrow(
            angle = 30,
            length = unit(0.05, "inches"),
            ends = "last", 
            type = "closed")))
# lg
```


### Plot significant results
#### Get terrestrial
```{r }
Eco2go = "Ter"

# choose to set limits to the y-axis. 
# options are: 
# 1) "obs"; for the range of observed mismatch values, or 
# 2) "preds", for the range of predicted mismatch values
my_limits <- c("obs","preds")
# choose if want to draw points (add observation points to plot or not)
# use TRUE or FALSE (here using TRUE for observations and FALSE for predictions)
draw_points_option <- c(TRUE,FALSE)

plots_ter <- list()

for(i in 1:length(my_limits)){
    
    limits2choose <- my_limits[i]
    draw_points <- draw_points_option[i]
    
    #################################
    # get significant variables
    vars_res <- summary(mm_lag_over_ter)
    vars_res <- data.frame(vars_res$coefficients$cond)
    vars_res <- vars_res |>
        filter(Pr...z.. <= .05) |>
        rownames_to_column('variable') 
    
    vars_res <- sapply(all_vars, function(x) any(grepl(x, vars_res$variable)))
    vars_res <- names(vars_res[which(vars_res)])
    vars_res_names <- all_vars_names[which(all_vars %in% vars_res)]
    vars_res_names_plot <- cont_vars_names_with_units[which(all_vars %in% vars_res)]
    
    # get the limits of predictions
    if(limits2choose == "preds"){
        plot_lims_y <- lapply(vars_res, function(x){
            ggeffects::ggeffect(
                mm_lag_over_ter, 
                terms = c(x,"lag_overshoot"))}
        )
        plot_lims_y <- rbindlist(plot_lims_y, fill = TRUE)
        plot_lims_y <- c(0,max(plot_lims_y$predicted) + 1)
    }
    if(limits2choose == "obs"){
        plot_lims_y <- mod_data|>
            filter(Eco==Eco2go) |>
            pull("mismatch") |>
            max()
        plot_lims_y <- c(0,plot_lims_y)
    }
    
    plots_ter[[i]] <- lapply(1:length(vars_res), function(x){
        
        var2go <- vars_res[x]
        var2go_name <- vars_res_names[x]
        var2go_name_plot <- vars_res_names_plot[x]
        
        subdata <- mod_data|>
            filter(Eco==Eco2go) |>
            select(var2go,"mismatch","lag_overshoot",obs_ID)
        
        # is the variable categorical or continuous?
        if(var2go %in% cont_vars){
            
            preds <- ggeffects::ggeffect(
                mm_lag_over_ter, 
                terms = c(var2go,"lag_overshoot"))
            
            # unscale x
            params_x <- bioshifts |>
                filter(obs_ID %in% mod_data$obs_ID) |>
                select(all_vars_data$all_vars_original[which(all_vars_data$all_vars==var2go)]) |>
                pull() 
            
            # apply same transformation as in mod_data
            cont_vars_trans_2go <- cont_vars_trans[which(cont_vars==var2go)]
            
            if(!is.na(cont_vars_trans_2go)){ 
                params_x <- match.fun(cont_vars_trans_2go)(params_x)
            }
            
            # Get attributes used by scale()
            params_x <- scale(params_x)
            center <- attr(params_x, "scaled:center")
            scale_factor <- attr(params_x, "scaled:scale")
            
            # unscale
            preds$x <- preds$x * scale_factor + center
            subdata[,var2go] <- subdata[,var2go] * scale_factor + center
            
            # remove transformation
            if(!is.na(cont_vars_trans_2go)){
                if(cont_vars_trans_2go=="log"){
                    preds$x <- exp(preds$x)
                    subdata[,var2go] <- exp(subdata[,var2go])
                }
                if(cont_vars_trans_2go=="log1p"){
                    preds$x <- exp(preds$x)-1
                    subdata[,var2go] <- exp(subdata[,var2go])-1
                }
            }
            
            sig_tmp <- data.frame(mm_lag_over_ter_eff[[var2go_name]])
            
            preds$sig <- ifelse(same_sign(sig_tmp$asymp.LCL, sig_tmp$asymp.UCL),
                                0,1)
            
            my_plot <- ggplot(preds, aes(x = x, y = predicted))
            
            # add observation points
            if(draw_points){
                my_plot <- my_plot + 
                    geom_point(data = subdata, aes_string(x = var2go, y = "mismatch", color = "lag_overshoot"), alpha = .1) 
            }
            
            if(var2go=="Boyce"){
                breaks_plot <- round(seq(0, 1, length.out = 5),1)
            }
            
            my_plot <- my_plot + 
                # set limits to the entire range of predictions
                coord_cartesian(ylim = plot_lims_y)+
                geom_line(aes(color = group, linetype = as.factor(sig)), size = 1.5) +
                geom_ribbon(aes(ymin=conf.low, ymax=conf.high, fill = group), alpha = .3) +
                theme_classic(base_size = 10)+
                scale_color_viridis_d(direction = -1)+
                scale_fill_viridis_d(direction = -1)+
                labs(x = var2go_name_plot, 
                     y  = "Mismatch",
                     fill = NULL, 
                     color = NULL,
                     linetype = NULL)+
                guides(linetype = "none")+
                theme(aspect.ratio=1)
            
            if(var2go=="Boyce"){
                
                breaks_plot <- seq(min(preds$x), 1, length.out = 4)
                
                my_plot <- my_plot + 
                    scale_x_continuous(breaks = breaks_plot,
                                       labels = round(breaks_plot,1),
                                       limits = c(min(breaks_plot),max(breaks_plot)))
                
                
            }
            
        } else {
            
            preds <- ggeffects::ggemmeans(
                mm_lag_over_ter, 
                terms = c(var2go,"lag_overshoot"))
            
            cat_vars_cont_pairs_res <- mm_lag_over_ter_cont_pairs[[var2go_name]]
            cat_vars_cont_pairs_res <- data.frame(cat_vars_cont_pairs_res)
            
            cat_vars_cont_pairs_res <- cat_vars_cont_pairs_res |>
                dplyr::filter(p.value <= 0.05)
            
            cat_vars_cont_pairs_res$from <- sapply(
                as.character(cat_vars_cont_pairs_res$contrast), 
                function(x){
                    strsplit(x," - ")[[1]][1]
                })
            cat_vars_cont_pairs_res$to <- sapply(
                as.character(cat_vars_cont_pairs_res$contrast), 
                function(x){
                    tmp <- strsplit(x," - ")[[1]][2]
                })
            
            cat_vars_cont_pairs_res$from <- gsub("(","",cat_vars_cont_pairs_res$from, 
                                                 fixed = TRUE)
            cat_vars_cont_pairs_res$from <- gsub(")","",cat_vars_cont_pairs_res$from, 
                                                 fixed = TRUE)
            cat_vars_cont_pairs_res$to <- gsub("(","",cat_vars_cont_pairs_res$to, 
                                               fixed = TRUE)
            cat_vars_cont_pairs_res$to <- gsub(")","",cat_vars_cont_pairs_res$to, 
                                               fixed = TRUE)
            
            y.bar <- sapply(cat_vars_cont_pairs_res$lag_overshoot, function(x){
                preds |>
                    group_by(group) |>
                    # summarise(val=max(conf.high)) |>
                    summarise(val=max(predicted)) |>
                    dplyr::filter(group==x) |>
                    dplyr::select(val) |>
                    pull()
            })
            y.bar.sum <- ave(seq_along(cat_vars_cont_pairs_res$lag_overshoot), cat_vars_cont_pairs_res$lag_overshoot, FUN = seq_along)
            # y.bar <- y.bar + (y.bar.sum - 1) * 1.5
            y.bar <- y.bar + y.bar.sum * 1.5
            
            t.test.df <- data.frame(
                p.value = as.vector(cat_vars_cont_pairs_res$p.value),
                right.tip = cat_vars_cont_pairs_res$to,
                left.tip = cat_vars_cont_pairs_res$from,
                group = cat_vars_cont_pairs_res$lag_overshoot,
                y.bar = y.bar
            )
            
            t.test.df$label_y = as.character(stars.pval(t.test.df$p.value))
            
            
            my_plot <- ggplot(as.data.frame(preds), aes(x = x, y = predicted)) 
            
            if(draw_points){
                my_plot <- my_plot + 
                    geom_jitter(data = subdata, 
                                aes_string(x = var2go, y = "mismatch", color = "lag_overshoot"), 
                                alpha = .05, width = .1) 
            }
            # set limits to the entire range of predictions
            my_plot <- my_plot + 
                coord_cartesian(ylim = plot_lims_y)+
                geom_errorbar(
                    aes(ymin=conf.low,ymax=conf.high, color = group), 
                    width=0, lwd=1, 
                    position = position_dodge(.5)) +
                geom_point(size=3, shape = 21, 
                           aes(fill = group), color = 'black',
                           position = position_dodge2(.5)) +
                theme_classic(base_size = 10)+
                scale_color_viridis_d(direction = -1)+
                scale_fill_viridis_d(direction = -1)+
                labs(x = var2go_name, 
                     y  = "Mismatch",
                     fill = NULL, 
                     color = NULL)+
                theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))+
                geom_text_pairwise(data = t.test.df,
                                   aes(xmin = left.tip,
                                       xmax = right.tip,
                                       y = (y.bar),
                                       label = label_y),
                                   na.rm = TRUE,
                                   size = 5, vjust = 0.3)+
                # expand_limits(y = (max(t.test.df$y.bar)+1)) +
                facet_wrap(.~group)
            
        }
        
        return(my_plot)
    })
    # plots_ter[[1]]
}

```

```{r fig.height=8,fig.width=8}

model_plots_ter <- list()
for(i in 1:length(my_limits)){
    
    plots_go <- plots_ter[[i]][1:5]
    plots_go[[6]] <- ggarrange(
        lg + theme(plot.margin = margin(10, 0, 0, 20)),
        get_legend(plots_ter[[i]][[1]]),
        align = "h",
        ncol = 2)
    
    model_plots_ter[[i]] <- ggarrange(
        ggarrange(plotlist = plots_go, 
                  labels = c(LETTERS[1:5],"Legend"),
                  font.label = list(size = 10),
                  align = "h",
                  legend = "none"),
        ggarrange(plotlist = plots_ter[[i]][6:7],
                  labels = LETTERS[6:7],
                  font.label = list(size = 10),
                  align = "hv",
                  # common.legend = TRUE,
                  legend="none"),
        heights = c(5,3.5), nrow = 2)
    model_plots_ter[[i]]
}

model_plots_ter

```


```{r fig.height=8,fig.width=8}
# 8*(5/(5+4)) + 8*(4/(5+4))
# 
# (8*(5/(5+4)) / 2) + 8*(4/(5+4))

file_names <- c(
    here("Figures/EffectPlots_lagovershoot_Ter_pairwise_all_obs"),
    here("Figures/EffectPlots_lagovershoot_Ter_pairwise_all")
)

for(i in 1:length(my_limits)){
    
    ggsave(plot = model_plots_ter[[i]], 
           filename = paste0(file_names[i],".png"), 
           device = "png",
           dpi = 320,
           bg = "white",
           width = 8, height = 8, 
           units = "in")
    
    ggsave(plot = model_plots_ter[[i]], 
           filename = paste0(file_names[i],".pdf"), 
           device = "pdf",
           bg = "white",
           width = 8, height = 8, 
           units = "in")
    
}

```

#### Get marine
```{r }

Eco2go = "Mar"

# choose to set limits to the y-axis. 
# options are: 
# 1) "obs"; for the range of observed mismatch values, or 
# 2) "preds", for the range of predicted mismatch values
my_limits <- c("obs","preds")
# choose if want to draw points (add observation points to plot or not)
# use TRUE or FALSE (here using TRUE for observations and FALSE for predictions)
draw_points_option <- c(TRUE,FALSE)

plots_mar <- list()

for(i in 1:length(my_limits)){
    
    limits2choose <- my_limits[i]
    draw_points <- draw_points_option[i]
    
    #################################
    # get significant variables
    vars_res <- summary(mm_lag_over_mar)
    vars_res <- data.frame(vars_res$coefficients$cond)
    vars_res <- vars_res |>
        filter(Pr...z.. <= .05) |>
        rownames_to_column('variable') 
    
    vars_res <- sapply(all_vars, function(x) any(grepl(x, vars_res$variable)))
    vars_res <- names(vars_res[which(vars_res)])
    vars_res_names <- all_vars_names[which(all_vars %in% vars_res)]
    vars_res_names_plot <- cont_vars_names_with_units[which(all_vars %in% vars_res)]
    
    # get the limits of predictions
    if(limits2choose == "preds"){
        plot_lims_y <- lapply(vars_res, function(x){
            ggeffects::ggeffect(
                mm_lag_over_mar, 
                terms = c(x,"lag_overshoot"))}
        )
        plot_lims_y <- rbindlist(plot_lims_y, fill = TRUE)
        plot_lims_y <- c(0,max(plot_lims_y$predicted) + 1)
    }
    if(limits2choose == "obs"){
        plot_lims_y <- mod_data|>
            filter(Eco==Eco2go) |>
            pull("mismatch") |>
            max()
        plot_lims_y <- c(0,plot_lims_y)
    }
    
    plots_mar[[i]] <- lapply(1:length(vars_res), function(x){
        
        var2go <- vars_res[x]
        var2go_name <- vars_res_names[x]
        var2go_name_plot <- vars_res_names_plot[x]
        
        subdata <- mod_data|>
            filter(Eco==Eco2go) |>
            select(var2go,"mismatch","lag_overshoot",obs_ID)
        
        # is the variable categorical or continuous?
        if(var2go %in% cont_vars){
            
            preds <- ggeffects::ggeffect(
                mm_lag_over_mar, 
                terms = c(var2go,"lag_overshoot"))
            
            # unscale x
            params_x <- bioshifts |>
                filter(obs_ID %in% mod_data$obs_ID) |>
                select(all_vars_data$all_vars_original[which(all_vars_data$all_vars==var2go)]) |>
                pull() 
            
            # apply same transformation as in mod_data
            cont_vars_trans_2go <- cont_vars_trans[which(cont_vars==var2go)]
            
            if(!is.na(cont_vars_trans_2go)){ 
                params_x <- match.fun(cont_vars_trans_2go)(params_x)
            }
            
            # Get attributes used by scale()
            params_x <- scale(params_x)
            center <- attr(params_x, "scaled:center")
            scale_factor <- attr(params_x, "scaled:scale")
            
            # unscale
            preds$x <- preds$x * scale_factor + center
            subdata[,var2go] <- subdata[,var2go] * scale_factor + center
            
            # remove transformation
            if(!is.na(cont_vars_trans_2go)){
                if(cont_vars_trans_2go=="log"){
                    preds$x <- exp(preds$x)
                    subdata[,var2go] <- exp(subdata[,var2go])
                }
                if(cont_vars_trans_2go=="log1p"){
                    preds$x <- exp(preds$x)-1
                    subdata[,var2go] <- exp(subdata[,var2go])-1
                }
            }
            
            sig_tmp <- data.frame(mm_lag_over_mar_eff[[var2go_name]])
            
            preds$sig <- ifelse(same_sign(sig_tmp$asymp.LCL, sig_tmp$asymp.UCL),
                                0,1)
            
            my_plot <- ggplot(preds, aes(x = x, y = predicted))
            
            # add observation points
            if(draw_points){
                my_plot <- my_plot + 
                    geom_point(data = subdata, aes_string(x = var2go, y = "mismatch", color = "lag_overshoot"), alpha = .1) 
            }
            
            if(var2go=="Boyce"){
                breaks_plot <- round(seq(0, 1, length.out = 5),1)
            }
            
            my_plot <- my_plot + 
                # set limits to the entire range of predictions
                coord_cartesian(ylim = plot_lims_y)+
                geom_line(aes(color = group, linetype = as.factor(sig)), size = 1.5) +
                geom_ribbon(aes(ymin=conf.low, ymax=conf.high, fill = group), alpha = .3) +
                theme_classic(base_size = 10)+
                scale_color_viridis_d(direction = -1)+
                scale_fill_viridis_d(direction = -1)+
                labs(x = var2go_name_plot, 
                     y  = "Mismatch",
                     fill = NULL, 
                     color = NULL,
                     linetype = NULL)+
                guides(linetype = "none")+
                theme(aspect.ratio=1)
            
            if(var2go=="Boyce"){
                
                breaks_plot <- seq(min(preds$x), 1, length.out = 4)
                
                my_plot <- my_plot + 
                    scale_x_continuous(breaks = breaks_plot,
                                       labels = round(breaks_plot,1),
                                       limits = c(min(breaks_plot),max(breaks_plot)))
                
                
            }
            
        } else {
            
            preds <- ggeffects::ggemmeans(
                mm_lag_over_mar, 
                terms = c(var2go,"lag_overshoot"))
            
            cat_vars_cont_pairs_res <- mm_lag_over_mar_cont_pairs[[var2go_name]]
            cat_vars_cont_pairs_res <- data.frame(cat_vars_cont_pairs_res)
            
            cat_vars_cont_pairs_res <- cat_vars_cont_pairs_res |>
                dplyr::filter(p.value <= 0.05)
            
            cat_vars_cont_pairs_res$from <- sapply(
                as.character(cat_vars_cont_pairs_res$contrast), 
                function(x){
                    strsplit(x," - ")[[1]][1]
                })
            cat_vars_cont_pairs_res$to <- sapply(
                as.character(cat_vars_cont_pairs_res$contrast), 
                function(x){
                    tmp <- strsplit(x," - ")[[1]][2]
                })
            
            cat_vars_cont_pairs_res$from <- gsub("(","",cat_vars_cont_pairs_res$from, 
                                                 fixed = TRUE)
            cat_vars_cont_pairs_res$from <- gsub(")","",cat_vars_cont_pairs_res$from, 
                                                 fixed = TRUE)
            cat_vars_cont_pairs_res$to <- gsub("(","",cat_vars_cont_pairs_res$to, 
                                               fixed = TRUE)
            cat_vars_cont_pairs_res$to <- gsub(")","",cat_vars_cont_pairs_res$to, 
                                               fixed = TRUE)
            
            y.bar <- sapply(cat_vars_cont_pairs_res$lag_overshoot, function(x){
                preds |>
                    group_by(group) |>
                    # summarise(val=max(conf.high)) |>
                    summarise(val=max(predicted)) |>
                    dplyr::filter(group==x) |>
                    dplyr::select(val) |>
                    pull()
            })
            y.bar.sum <- ave(seq_along(cat_vars_cont_pairs_res$lag_overshoot), cat_vars_cont_pairs_res$lag_overshoot, FUN = seq_along)
            # y.bar <- y.bar + (y.bar.sum - 1) * 1.5
            y.bar <- y.bar + y.bar.sum * 1.5
            
            t.test.df <- data.frame(
                p.value = as.vector(cat_vars_cont_pairs_res$p.value),
                right.tip = cat_vars_cont_pairs_res$to,
                left.tip = cat_vars_cont_pairs_res$from,
                group = cat_vars_cont_pairs_res$lag_overshoot,
                y.bar = y.bar
            )
            
            t.test.df$label_y = as.character(stars.pval(t.test.df$p.value))
            
            
            my_plot <- ggplot(as.data.frame(preds), aes(x = x, y = predicted)) 
            
            if(draw_points){
                my_plot <- my_plot + 
                    geom_jitter(data = subdata, 
                                aes_string(x = var2go, y = "mismatch", color = "lag_overshoot"), 
                                alpha = .05, width = .1) 
            }
            # set limits to the entire range of predictions
            my_plot <- my_plot + 
                coord_cartesian(ylim = plot_lims_y)+
                geom_errorbar(
                    aes(ymin=conf.low,ymax=conf.high, color = group), 
                    width=0, lwd=1, 
                    position = position_dodge(.5)) +
                geom_point(size=3, shape = 21, 
                           aes(fill = group), color = 'black',
                           position = position_dodge2(.5)) +
                theme_classic(base_size = 10)+
                scale_color_viridis_d(direction = -1)+
                scale_fill_viridis_d(direction = -1)+
                labs(x = var2go_name, 
                     y  = "Mismatch",
                     fill = NULL, 
                     color = NULL)+
                theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))+
                geom_text_pairwise(data = t.test.df,
                                   aes(xmin = left.tip,
                                       xmax = right.tip,
                                       y = (y.bar),
                                       label = label_y),
                                   na.rm = TRUE,
                                   size = 5, vjust = 0.3)+
                # expand_limits(y = (max(t.test.df$y.bar)+1)) +
                facet_wrap(.~group)
            
        }
        
        return(my_plot)
    })
    # plots_mar[[1]]
}

```


```{r fig.height=5.77,fig.width=8}

model_plots_mar <- list()
for(i in 1:length(my_limits)){
    
    plots_go <- plots_mar[[i]][1:2]
    plots_go[[3]] <- ggarrange(
        lg + theme(plot.margin = margin(10, 0, 0, 20)),
        get_legend(plots_mar[[i]][[1]]),
        align = "h",
        ncol = 2)
    
    model_plots_mar[[i]] <- ggarrange(
        ggarrange(plotlist = plots_go, 
                  labels = c(LETTERS[1:2],"Legend"),
                  font.label = list(size = 10),
                  align = "h",
                  legend = "none",
                  nrow = 1),
        ggarrange(plotlist = plots_mar[[i]][3:4],
                  labels = LETTERS[3:4],
                  font.label = list(size = 10),
                  align = "hv",
                  legend="none"),
        heights = c(2.5,3.5), nrow = 2)
    model_plots_mar[[i]]
}
model_plots_mar

```


```{r fig.height=5.77,fig.width=8}

file_names <- c(
    here("Figures/EffectPlots_lagovershoot_Mar_pairwise_all_obs"),
    here("Figures/EffectPlots_lagovershoot_Mar_pairwise_all")
)

for(i in 1:length(my_limits)){
    
    ggsave(plot = model_plots_mar[[i]], 
           filename = paste0(file_names[i],".png"), 
           device = "png",
           dpi = 320,
           bg = "white",
           width = 8, height = 5.77, 
           units = "in")
    
    ggsave(plot = model_plots_mar[[i]], 
           filename = paste0(file_names[i],".pdf"), 
           device = "pdf",
           bg = "white",
           width = 8, height = 5.77, 
           units = "in")
    
}


```


## Mismatch between documented and modeled shifts

### Terrestrial

#### Fit model
```{r fig.width=8, fig.height=8}

y <- strsplit(mm_formula_ter," ")[[1]][1]

var_to_go <- strsplit(paste(mm_formula_ter, "+", RE_nest),"~|[+]|[/]|[|]")[[1]]
var_to_go <- gsub(" |[(]|[)]","",var_to_go)
var_to_go <- var_to_go[-which(var_to_go=="1")]
subdata_ter <- mod_data |>
    filter(Eco=="Ter") |> 
    select(var_to_go,lag_overshoot) |>
    na.omit()

# fit model
mm_ter <- glmmTMB(as.formula(paste(mm_formula_ter, "+", RE_nest)),
                  family = Gamma(link = "log"),
                  data = subdata_ter)

# tab_model(mm_ter, transform = NULL, string.est = "Estimate")

# R-squared
r.squaredGLMM(mm_ter)
```


#### Get emmeans for categorical variables
```{r fig.width=8, fig.height=8}
# get emmeans for categorical variables
cat_vars_names <- gsub("_"," ",cat_vars)

mm_ter_emm <- lapply(cat_vars, function(x){
    if(any(grepl(x,var_to_go))){
        emmeans(object = mm_ter, 
                specs = x,
                type = "response")
    }
})
names(mm_ter_emm) <- cat_vars_names

test <- sapply(mm_ter_emm,is.null)
if(any(test)){
    mm_ter_emm <- mm_ter_emm[-which(test)]
}
mm_ter_emm

mm_ter_cont <- lapply(mm_ter_emm, 
                      contrast, 
                      method = "eff",
                      type = "response")
names(mm_ter_cont) <- names(mm_ter_cont)
mm_ter_cont

mm_ter_cont_pairs <- lapply(mm_ter_emm, 
                            contrast, 
                            method = "pairwise",
                            type = "response",
                            adjust = "none")
names(mm_ter_cont_pairs) <- names(mm_ter_cont)
mm_ter_cont_pairs

```




### Marine
#### Fit model
```{r fig.width=8, fig.height=8}
y <- strsplit(mm_formula_mar," ")[[1]][1]

var_to_go <- strsplit(paste(mm_formula_mar, "+", RE_nest),"~|[+]|[/]|[|]")[[1]]
var_to_go <- gsub(" |[(]|[)]","",var_to_go)
var_to_go <- var_to_go[-which(var_to_go=="1")]

subdata_mar <- mod_data |>
    filter(Eco=="Mar") |> 
    select(var_to_go) |>
    na.omit()

# fit model
mm_mar <- glmmTMB(as.formula(paste(mm_formula_mar, "+", RE_nest)),
                  family = Gamma(link = "log"),
                  data = subdata_mar)

# tab_model(mm_mar, transform = NULL, string.est = "Estimate")

# R-squared
r.squaredGLMM(mm_mar)
```

#### Get emmeans for categorical variables
```{r fig.width=8, fig.height=8}
cat_vars_names <- gsub("_"," ",cat_vars)

mm_mar_emm <- lapply(cat_vars, function(x){
    if(any(grepl(x,var_to_go))){
        emmeans(object = mm_mar, 
                specs = x,
                type = "response")
    }
})
names(mm_mar_emm) <- cat_vars_names

test <- sapply(mm_mar_emm,is.null)
if(any(test)){
    mm_mar_emm <- mm_mar_emm[-which(test)]
}
mm_mar_emm

mm_mar_cont <- lapply(mm_mar_emm, 
                      contrast, 
                      method = "eff",
                      type = "response")
names(mm_mar_cont) <- names(mm_mar_cont)
mm_mar_cont

mm_mar_cont_pairs <- lapply(mm_mar_emm, 
                            contrast, 
                            method = "pairwise",
                            type = "response",
                            adjust = "none")
names(mm_mar_cont_pairs) <- names(mm_mar_cont)
mm_mar_cont_pairs

```


### Combined model output

```{r}
tab_model(mm_ter, mm_mar, 
          transform = NULL, string.est = "Estimate", string.p = "P-value",
          dv.labels = c("Terrestrial","Marine"),  
          show.icc = FALSE, 
          show.r2 = FALSE,
          show.ngroups = FALSE,
          show.re.var = FALSE,
          show.obs = FALSE)
```

### Plot significant results
#### Get terrestrial
```{r }

# get significant variables
vars_res <- summary(mm_ter)
vars_res <- data.frame(vars_res$coefficients$cond)
vars_res <- vars_res |>
    filter(Pr...z.. <= .05) |>
    rownames_to_column('variable') 


vars_res <- sapply(all_vars, function(x) any(grepl(x, vars_res$variable)))
vars_res <- names(vars_res[which(vars_res)])
vars_res_names <- all_vars_names[which(all_vars %in% vars_res)]

plots_ter <- lapply(1:length(vars_res), function(x){
    
    var2go <- vars_res[x]
    var2go_name <- vars_res_names[x]
    
    # is the variable categorical or continuous?
    if(var2go %in% cont_vars){
        
        preds <- ggeffects::ggeffect(
            mm_ter, 
            terms = c(var2go))
        
        # unscale x
        params_x <- bioshifts |>
            select(all_vars_original[x]) |>
            pull()
        
        if(!is.na(cont_vars_trans[x])){
            params_x <- match.fun(cont_vars_trans[x])(params_x)
        }
        
        sd_x <- sd(params_x,na.rm = TRUE)
        mean_x <- mean(params_x,na.rm = TRUE)
        
        preds$x <- preds$x * sd_x + mean_x
        
        if(!is.na(cont_vars_trans[x])){
            if(cont_vars_trans[x]=="log"){
                preds$x <- exp(preds$x)
            }
            if(cont_vars_trans[x]=="log1p"){
                preds$x <- exp(preds$x)-1
            }
        }
        
        my_plot <- ggplot(preds, aes(x = x, y = predicted))+
            geom_line(size = 1.5) +
            geom_ribbon(aes(ymin=conf.low, ymax=conf.high), alpha = .3) +
            theme_classic(base_size = 10)+
            scale_color_viridis_d()+
            scale_fill_viridis_d()+
            labs(x = var2go_name, 
                 y  = "Mismatch",
                 fill = NULL, 
                 color = NULL)+
            theme(aspect.ratio=1)
        
    } else {
        
        preds <- ggeffects::ggemmeans(
            mm_ter, 
            terms = c(var2go))
        
        cat_vars_cont_pairs_res <- mm_ter_cont_pairs[[var2go_name]]
        cat_vars_cont_pairs_res <- data.frame(cat_vars_cont_pairs_res)
        
        cat_vars_cont_pairs_res <- cat_vars_cont_pairs_res |>
            dplyr::filter(p.value <= 0.05)
        
        cat_vars_cont_pairs_res$from <- sapply(
            as.character(cat_vars_cont_pairs_res$contrast), 
            function(x){
                strsplit(x," / ")[[1]][1]
            })
        cat_vars_cont_pairs_res$to <- sapply(
            as.character(cat_vars_cont_pairs_res$contrast), 
            function(x){
                tmp <- strsplit(x," / ")[[1]][2]
            })
        
        cat_vars_cont_pairs_res$from <- gsub("(","",cat_vars_cont_pairs_res$from, 
                                             fixed = TRUE)
        cat_vars_cont_pairs_res$from <- gsub(")","",cat_vars_cont_pairs_res$from, 
                                             fixed = TRUE)
        cat_vars_cont_pairs_res$to <- gsub("(","",cat_vars_cont_pairs_res$to, 
                                           fixed = TRUE)
        cat_vars_cont_pairs_res$to <- gsub(")","",cat_vars_cont_pairs_res$to, 
                                           fixed = TRUE)
        
        t.test.df <- data.frame(
            p.value = as.vector(cat_vars_cont_pairs_res$p.value),
            right.tip = cat_vars_cont_pairs_res$to,
            left.tip = cat_vars_cont_pairs_res$from,
            y.bar = max(preds$conf.high) + seq(1,nrow(cat_vars_cont_pairs_res))
        )
        
        t.test.df$label_y = as.character(stars.pval(t.test.df$p.value))
        
        
        my_plot <- ggplot(as.data.frame(preds), aes(x = x, y = predicted)) +
            geom_errorbar(
                aes(ymin=conf.low,ymax=conf.high), 
                width=0, lwd=1, 
                position = position_dodge(.5)) +
            geom_point(size=3, shape = 21, 
                       color = 'black',
                       fill = "white",
                       position = position_dodge2(.5)) +
            theme_classic(base_size = 10)+
            labs(x = var2go_name, 
                 y  = "Mismatch")+
            theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))+
            geom_text_pairwise(data = t.test.df,
                               aes(xmin = left.tip,
                                   xmax = right.tip,
                                   y = (y.bar),
                                   label = label_y),
                               na.rm = TRUE,
                               size = 5, vjust = 0.1)+
            expand_limits(y = (max(t.test.df$y.bar)+1))
        
    }
    
    return(my_plot)
})
# plots_ter
```

```{r fig.height=8,fig.width=4}
model_plots_ter <- ggarrange(
    ggarrange(plotlist = plots_ter[1:4], 
              labels = LETTERS[1:4],
              font.label = list(size = 10),
              align = "hv",
              legend = "none"),
    ggarrange(plotlist = plots_ter[5],
              labels = LETTERS[5],
              font.label = list(size = 10),
              align = "hv",
              common.legend = TRUE,
              legend="bottom"),
    heights = c(5,4), nrow = 2)
model_plots_ter

# 8*(5/(5+4)) + 8*(4/(5+4))
# 
# (8*(5/(5+4)) / 2) + 8*(4/(5+4))

ggsave(plot = model_plots_ter, 
       filename = here("Figures/EffectPlots_mismatch_Ter_pairwise_all.svg"), 
       device = "svg", 
       dpi = 300,
       width = 8, height = 8, 
       units = "in")


```

#### Get marine
```{r }

# get significant variables
vars_res <- summary(mm_mar)
vars_res <- data.frame(vars_res$coefficients$cond)
vars_res <- vars_res |>
    filter(Pr...z.. <= .05) |>
    rownames_to_column('variable') 


vars_res <- sapply(all_vars, function(x) any(grepl(x, vars_res$variable)))
vars_res <- names(vars_res[which(vars_res)])
vars_res_names <- all_vars_names[which(all_vars %in% vars_res)]

plots_mar <- lapply(1:length(vars_res), function(x){
    
    var2go <- vars_res[x]
    var2go_name <- vars_res_names[x]
    
    # is the variable categorical or continuous?
    if(var2go %in% cont_vars){
        
        preds <- ggeffects::ggeffect(
            mm_mar, 
            terms = c(var2go))
        
        # unscale x
        params_x <- bioshifts |>
            select(all_vars_original[x]) |>
            pull()
        
        if(!is.na(cont_vars_trans[x])){
            params_x <- match.fun(cont_vars_trans[x])(params_x)
        }
        
        sd_x <- sd(params_x,na.rm = TRUE)
        mean_x <- mean(params_x,na.rm = TRUE)
        
        preds$x <- preds$x * sd_x + mean_x
        
        if(!is.na(cont_vars_trans[x])){
            if(cont_vars_trans[x]=="log"){
                preds$x <- exp(preds$x)
            }
            if(cont_vars_trans[x]=="log1p"){
                preds$x <- exp(preds$x)-1
            }
        }
        
        my_plot <- ggplot(preds, aes(x = x, y = predicted))+
            geom_line(size = 1.5) +
            geom_ribbon(aes(ymin=conf.low, ymax=conf.high), alpha = .3) +
            theme_classic(base_size = 10)+
            scale_color_viridis_d()+
            scale_fill_viridis_d()+
            labs(x = var2go_name, 
                 y  = "Mismatch",
                 fill = NULL, 
                 color = NULL)+
            theme(aspect.ratio=1)
        
    } else {
        
        preds <- ggeffects::ggemmeans(
            mm_mar, 
            terms = c(var2go))
        
        cat_vars_cont_pairs_res <- mm_mar_cont_pairs[[var2go_name]]
        cat_vars_cont_pairs_res <- data.frame(cat_vars_cont_pairs_res)
        
        cat_vars_cont_pairs_res <- cat_vars_cont_pairs_res |>
            dplyr::filter(p.value <= 0.05)
        
        cat_vars_cont_pairs_res$from <- sapply(
            as.character(cat_vars_cont_pairs_res$contrast), 
            function(x){
                strsplit(x," / ")[[1]][1]
            })
        cat_vars_cont_pairs_res$to <- sapply(
            as.character(cat_vars_cont_pairs_res$contrast), 
            function(x){
                tmp <- strsplit(x," / ")[[1]][2]
            })
        
        cat_vars_cont_pairs_res$from <- gsub("(","",cat_vars_cont_pairs_res$from, 
                                             fixed = TRUE)
        cat_vars_cont_pairs_res$from <- gsub(")","",cat_vars_cont_pairs_res$from, 
                                             fixed = TRUE)
        cat_vars_cont_pairs_res$to <- gsub("(","",cat_vars_cont_pairs_res$to, 
                                           fixed = TRUE)
        cat_vars_cont_pairs_res$to <- gsub(")","",cat_vars_cont_pairs_res$to, 
                                           fixed = TRUE)
        
        t.test.df <- data.frame(
            p.value = as.vector(cat_vars_cont_pairs_res$p.value),
            right.tip = cat_vars_cont_pairs_res$to,
            left.tip = cat_vars_cont_pairs_res$from,
            y.bar = max(preds$conf.high) + seq(1,nrow(cat_vars_cont_pairs_res))
        )
        
        t.test.df$label_y = as.character(stars.pval(t.test.df$p.value))
        
        
        my_plot <- ggplot(as.data.frame(preds), aes(x = x, y = predicted)) +
            geom_errorbar(
                aes(ymin=conf.low,ymax=conf.high), 
                width=0, lwd=1, 
                position = position_dodge(.5)) +
            geom_point(size=3, shape = 21, 
                       color = 'black',
                       fill = "white",
                       position = position_dodge2(.5)) +
            theme_classic(base_size = 10)+
            labs(x = var2go_name, 
                 y  = "Mismatch")+
            theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))+
            geom_text_pairwise(data = t.test.df,
                               aes(xmin = left.tip,
                                   xmax = right.tip,
                                   y = (y.bar),
                                   label = label_y),
                               na.rm = TRUE,
                               size = 5, vjust = 0.1)+
            expand_limits(y = (max(t.test.df$y.bar)+1)) 
        
    }
    
    return(my_plot)
})
# plots_mar
```




```{r fig.height=8,fig.width=4}
model_plots_mar <- ggarrange(
    ggarrange(plotlist = plots_mar[1:3], 
              labels = LETTERS[1:2],
              font.label = list(size = 10),
              align = "hv",
              legend = "none"),
    ggarrange(plotlist = plots_mar[4],
              labels = LETTERS[4],
              font.label = list(size = 10),
              align = "hv",
              common.legend = TRUE,
              legend="bottom"),
    heights = c(5,4), nrow = 2)
model_plots_mar

ggsave(plot = model_plots_mar, 
       filename = here("Figures/EffectPlots_mismatch_Mar_pairwise_all.png"), 
       device = "png",
       dpi = 300,
       width = 8, height = 5.77, 
       units = "in")

```


## Test if elevation shifts explain the mismatch
```{r}
# 1) How many cases are there of reported both latitude and elevation shifts for the same species
subdata_ter <- mod_data |> 
    filter(Eco=="Ter") |>
    select(ID, sp_name_std)

bioshifts_full <- read.csv(here(Bioshifts_dir,Bioshifts_DB_v3))

bioshifts_type <- bioshifts_full |>
    mutate(sp_name_std = sp_name_checked) |>
    select(ID, Type, Param, sp_name_std, Rate,  LatCentDeg) |>
    group_by(ID, Type, Param,  LatCentDeg, sp_name_std) |>
    summarise(Rate = mean(Rate)) |>
    spread(Type, Rate) |>
    na.omit()

bioshifts_type <- bioshifts_type[which(bioshifts_type$sp_name_std %in% subdata_ter$sp_name_std),]

```

### Model
```{r}
# ------------------------
# fit a model including elevation shift velocity
# we expect this could help explain the mismatches
mm_lagover_formula_ter_ele <- "mismatch ~ lag_overshoot + Boyce + Study_duration + Interannual_var + Lat_extent + Connectivity + Sampling_method + Locomotion_mode + ELE + (1 | Param/class/ID)"

mm_lagover_formula_ter <- "mismatch ~ lag_overshoot + Boyce + Study_duration + Interannual_var + Lat_extent + Connectivity + Sampling_method + Locomotion_mode + (1 | Param/class/ID)"


y <- strsplit(mm_lagover_formula_ter_ele," ")[[1]][1]

var_to_go <- strsplit(mm_lagover_formula_ter_ele,"~|[+]|[/]|[|]|[*]")[[1]]
var_to_go <- gsub(" |[(]|[)]","",var_to_go)
var_to_go <- var_to_go[-which(var_to_go=="1")]
var_to_go <- unique(var_to_go)

# add documented velocities to the modeling dataset
bioshifts_type$ELE <- abs(bioshifts_type$ELE)
subdata_ter <- mod_data |>
    filter(Eco=="Ter") |>
    left_join(bioshifts_type[,c("ID", "Param", "sp_name_std","ELE")], 
              by = c("ID", "Param", "sp_name_std"))|>
    select(var_to_go,sp_name_std) |>
    na.omit()

# N studies 
length(unique(subdata_ter$ID)) # 10
# N species
length(unique(subdata_ter$sp_name_std)) # 299
# total dataset size
nrow(subdata_ter) # 512

mm_lag_over_ter <- glmmTMB(as.formula(mm_lagover_formula_ter),
                           family = Gamma(link = "log"),
                           data = subdata_ter)

mm_lag_over_ter_vel <- glmmTMB(as.formula(mm_lagover_formula_ter_ele),
                               family = Gamma(link = "log"),
                               data = subdata_ter)

# R-squared of the model without velEle
performance::r2(mm_lag_over_ter)

# R-squared of the model with velEle
performance::r2(mm_lag_over_ter_vel)

comp_mod <- anova(mm_lag_over_ter,mm_lag_over_ter_vel)
comp_mod

tab_model(mm_lag_over_ter, mm_lag_over_ter_vel, 
          transform = NULL, string.est = "Estimate", string.p = "P-value",
          dv.labels = c("Latitude","Latitude + Elevation"),  
          show.icc = FALSE, 
          show.r2 = FALSE,
          show.ngroups = FALSE,
          show.re.var = FALSE,
          show.obs = FALSE)

```

### Plot
```{r fig.width=4, fig.height=4}

pred_ELE <- ggeffects::ggeffect(
    mm_lag_over_ter_vel, 
    terms = c("ELE"))

# unscale vars
params_xy <- bioshifts |>
    select(ShiftRate, bvel_lat) |>
    summarise(ShiftRate_mean = mean(ShiftRate,na.rm = TRUE),
              ShiftRate_sd = sd(ShiftRate,na.rm = TRUE))


# make plot terrestrial
pred_ELE$predicted <- pred_ELE$predicted * params_xy$ShiftRate_sd + params_xy$ShiftRate_mean
pred_ELE$conf.low <-pred_ELE$conf.low * params_xy$ShiftRate_sd + params_xy$ShiftRate_mean
pred_ELE$conf.high <-pred_ELE$conf.high * params_xy$ShiftRate_sd + params_xy$ShiftRate_mean

subdata_ter_plot <- subdata_ter
subdata_ter_plot$mismatch <- subdata_ter$mismatch * params_xy$ShiftRate_sd + params_xy$ShiftRate_mean

pred_ELE_plot <-
    ggplot(pred_ELE, aes(x = x, y = predicted))+
    geom_point(data = subdata_ter_plot, 
               aes(x = ELE, y = mismatch), alpha = .1) +
    geom_line(size = 1.5) +
    geom_ribbon(aes(ymin=conf.low, ymax=conf.high), alpha = .3) +
    theme_classic(base_size = 10)+
    labs(x = "Elevational range shift (km/yr)", 
         y = "Mismatch",
         fill = NULL, 
         color = NULL)+
    theme(aspect.ratio=1)

pred_ELE_plot 

ggsave(plot = pred_ELE_plot, 
       filename = here("Figures/EffectPlots_mismatch_Elevation.png"), 
       device = "png", 
       dpi = 300,
       width = 4, height = 4, 
       units = "in")

```

## Test if latitude explains the mismatch

### Add latitude to the dataset
```{r}
mod_data_lat <- merge(mod_data,
                      bioshifts_full %>%
                          mutate(LatCentDeg = abs(LatCentDeg)) %>%
                          select(ID, LatCentDeg) %>%
                          unique(),
                      by = "ID")
```

### Correlations
```{r}

cor(mod_data_lat |>
        filter(Eco=="Ter") |>
        select(LatCentDeg, ShiftRate_raw, ShiftMod_raw))

cor(mod_data_lat |>
        filter(Eco=="Mar") |>
        select(LatCentDeg, ShiftRate_raw, ShiftMod_raw))

```

### Models
```{r}
y <- "mismatch"

mm_lagover_formula_lat <- "mismatch ~ Boyce*lag_overshoot + Study_duration*lag_overshoot + Interannual_var*lag_overshoot + Lat_extent*lag_overshoot + Connectivity*lag_overshoot + Sampling_method*lag_overshoot + Locomotion_mode*lag_overshoot + LatCentDeg + (1 | Param/class/ID)"

var_to_go <- strsplit(mm_lagover_formula_lat,"~|[+]|[/]|[|]|[*]")[[1]]
var_to_go <- gsub(" |[(]|[)]","",var_to_go)
var_to_go <- var_to_go[-which(var_to_go=="1")]
var_to_go <- unique(var_to_go)

subdata_ter <- mod_data_lat |>
    filter(Eco=="Ter") |> 
    select(var_to_go) |>
    na.omit()

mm_lag_over_ter_lat <- glmmTMB(as.formula(mm_lagover_formula_lat),
                               family = Gamma(link = "log"),
                               data = subdata_ter)

summary(mm_lag_over_ter_lat)

subdata_mar <- mod_data_lat |>
    filter(Eco=="Mar") |> 
    select(var_to_go) |>
    na.omit()

mm_lag_over_mar_lat <- glmmTMB(as.formula(mm_lagover_formula_lat),
                               family = Gamma(link = "log"),
                               data = subdata_mar)

summary(mm_lag_over_mar_lat)

```

# Map study areas
```{r }

mundi <- rnaturalearth::ne_coastline(scale = 110, returnclass = "sf")
mundi <- crop(vect(mundi), ext(c(-180,180,-60,90)))

# get raster bioshifts shp files study areas
get_raster_bioshifts = "YES"
if(get_raster_bioshifts=="YES"){
    my_ext = terra::ext(mundi)
    my_crs = crs(mundi)
    rast_bioshifts_studies <- terra::rast(my_ext, crs = my_crs, res = 5)
    values(rast_bioshifts_studies) <- 0
    
    rast_bioshifts_species <- rast_bioshifts_studies
    
    fgdb <- geopackage(
        here(Bioshifts_dir,"ShapefilesBioShiftsv3_studyareas.gpkg"), 
        connect = TRUE)
    
    fgdb <- gpkg_tables(fgdb, "ShapefilesBioShiftsv3_studyareas", collect = TRUE)
    fgdb <- fgdb$ShapefilesBioShiftsv3_studyareas
    
    fc_list <- unique(mod_data$ID)
    
    for(i in 1:length(fc_list)){ cat("\r",i,"from",length(fc_list))
        tmp = fgdb |> filter(Name == fc_list[i])
        tmp = terra::cells(rast_bioshifts_studies,tmp)
        tmp_cell = tmp[,2]
        tmp_vals = rast_bioshifts_studies[tmp_cell][,1]
        rast_bioshifts_studies[tmp_cell] = tmp_vals+1
        
        tmp_vals = rast_bioshifts_species[tmp_cell][,1]
        n_sps <- mod_data |> 
            filter(ID == fc_list[i]) |> 
            summarise(N = length(unique(sp_name_std))) |>
            pull(N)
        rast_bioshifts_species[tmp_cell] = tmp_vals+n_sps
    }
    names(rast_bioshifts_studies) <- "SA"
    rast_bioshifts_studies[rast_bioshifts_studies==0] <- NA
    
    names(rast_bioshifts_species) <- "sps"
    rast_bioshifts_species[rast_bioshifts_species==0] <- NA
    
    rast_bioshifts_studies <- c(rast_bioshifts_studies,
                                rast_bioshifts_species)
    
    writeRaster(rast_bioshifts_studies, here::here("Data/raster_bioshifts_SA_SDMs.tif"), overwrite = TRUE)
    
} else {
    rast_bioshifts_studies <- terra::rast(here::here("Data/raster_bioshifts_SA_SDMs.tif"))
    rast_bioshifts_studies <- crop(rast_bioshifts_studies, ext(mundi))
}

mundi_eq <- project(mundi, "+proj=moll")
rast_bioshifts_studies <- project(rast_bioshifts_studies, "+proj=moll")
rast_bioshifts_studies <- rast_bioshifts_studies

bio_sp_fig_breaks <- round(exp(seq(
    log(as.numeric(global(rast_bioshifts_studies$sps,min,na.rm=TRUE))),
    log(as.numeric(global(rast_bioshifts_studies$sps,max,na.rm=TRUE))),
    length.out = 5)),0)

bio_sp_fig <- ggplot(data=mundi_eq) +
    stat_summary_hex(
        data=as.data.frame(rast_bioshifts_studies$sps, xy = TRUE),
        aes(x = x, y = y, z = sps),
        color = "white",
        fun = function(x) mean(x,na.rm = TRUE),
        bins = c(80,80),
        alpha = .5) +
    geom_spatvector() +
    scale_fill_viridis_c(
        name = "Number of species",
        alpha=.7,
        na.value = "white",
        trans="log",
        labels=bio_sp_fig_breaks,
        breaks=bio_sp_fig_breaks,
        limits = range(bio_sp_fig_breaks)
    )+
    theme_void()+
    guides(fill = guide_colorbar(title.position = "top",
                                 position = "bottom",
                                 theme = theme(
                                     legend.key.width  = unit(20, "lines"),
                                     legend.key.height = unit(.8, "lines"),
                                     legend.title = element_text(hjust = 0.5))))

bio_SA_fig_breaks <- round(exp(seq(
    log(as.numeric(global(rast_bioshifts_studies$SA,min,na.rm=TRUE))),
    log(as.numeric(global(rast_bioshifts_studies$SA,max,na.rm=TRUE))),
    length.out = 5)),0)

bio_SA_fig <-
    ggplot(data=mundi_eq) +
    stat_summary_hex(
        data=as.data.frame(rast_bioshifts_studies$SA, xy = TRUE),
        aes(x = x, y = y, z = SA),
        color = "white",
        fun = function(x) mean(x,na.rm = TRUE),
        bins = c(80,80),
        alpha = .5) +
    geom_spatvector() +
    scale_fill_viridis_c(
        name = "Number of study areas",
        alpha=.7,
        na.value = "white",
        trans="log",
        labels=bio_SA_fig_breaks,
        breaks=bio_SA_fig_breaks,
        limits = range(bio_SA_fig_breaks)
    )+
    theme_void()+
    guides(fill = guide_colorbar(title.position = "top",
                                 position = "bottom",
                                 theme = theme(
                                     legend.key.width  = unit(20, "lines"),
                                     legend.key.height = unit(.8, "lines"),
                                     legend.title = element_text(hjust = 0.5)))) 


```


```{r fig.height=8,fig.width=8}

my_map <- ggarrange(bio_SA_fig,
                    bio_sp_fig, 
                    nrow = 2,
                    labels = "AUTO")
my_map

ggsave(plot = my_map,
       filename = here("Figures/Map_studies.png"), 
       device = "png",
       width = 8, height = 8, 
       units = "in", bg = "white")

```





***




