---
title: "Compare velocity values from Bioshifts v1 and v3"
author: "Brunno Oliveira"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
    code_folding: show
---

# Setup
```{r message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(terra)
library(RColorBrewer)
library(VoCC)

# set computer
# computer = "muse"
# 
# if(computer == "muse"){
#     setwd("/storage/simple/projects/t_cesab/brunno/Exposure-SDM")
# 
#     work_dir <- getwd()
#     here::i_am("R/7_velocity/2_compare_velocity_v3_v1.R")
# } else {
#   work_dir <- here::here(work_dir)
# }

# source settings
source(here::here("R/settings.R"))

```

# Load in v3 velocities
```{r}
SAs <- list.files(here::here("Data/Velocity_SA"))
SAs <- gsub(".csv","",SAs)
SA_got <- list.files(here::here("Data/Velocity_SA"), full.names = TRUE)
SA_got <- lapply(1:length(SA_got), function(i) {
    tmp <- read.csv(SA_got[i])
    tmp$ID = SAs[i]
    tmp
})
SA_got <- rbindlist(SA_got, fill = TRUE)

SA_got <- SA_got %>% 
    dplyr::select(c(v.lat.median.mat,baseline.mat,trend.mean.mat,v.ele.median.mat,baseline.sst,trend.mean.sst,v.lat.median.sst,ID))

SA_got$ECO <- ifelse(is.na(SA_got$v.lat.median.sst), "T", "M")

SA_got <- SA_got %>%
    mutate(vel_v3 = ifelse(ECO == "T", v.lat.median.mat, v.lat.median.sst),
           baseline.temp_v3 = ifelse(ECO == "T", baseline.mat, baseline.sst),
           Trend = ifelse(ECO == "T", trend.mean.mat, trend.mean.sst)) %>%
    dplyr::select(ECO, vel_v3, Trend, ID, baseline.temp_v3)

```

# Load in Bioshifts v1
```{r}
Bioshifts_DB <- read.csv(here::here("Data/Bioshifts/biov1_fixednames.csv"))

# Filter Polygons in Study areas v3
Bioshifts_DB <- Bioshifts_DB[Bioshifts_DB$ID %in% unique(SA_got$ID),]

# convert everything to km/year
Bioshifts_DB <- Bioshifts_DB %>%
    mutate(
        Type = case_when(
            Type=="HOR" ~ "LAT",
            TRUE ~ as.character(Type)),
        # Standardize shift measures to km/year
        SHIFT = case_when(
            UNIT == "m/year" ~ SHIFT/1000, 
            UNIT == "km/year" ~ SHIFT, 
            TRUE ~ NA_real_))

```

# Merge v1 to v3
```{r}
Bioshifts_vel <- merge(Bioshifts_DB %>%
                           filter(Type %in% c("LAT","ELE")) %>%
                           mutate(vel_v1 = ifelse(Type == "LAT", v.lat.mean, v.ele.mean),
                                  baseline.temp_v1 = baseline.temp) %>% 
                           dplyr::select(ID, vel_v1, Type, ECO, SHIFT, baseline.temp_v1), 
                       SA_got  %>%
                           dplyr::select(ID, vel_v3, ECO, baseline.temp_v3), 
                       by = c("ID","ECO"))
head(Bioshifts_vel)


```

# Compare velocity
## Density plots
```{r}
v1 <- Bioshifts_vel[,c("ID","Type","ECO","vel_v1","baseline.temp_v1","SHIFT")]
v3 <- Bioshifts_vel[,c("ID","Type","ECO","vel_v3","baseline.temp_v3","SHIFT")]

v1$baseline.temp <- v1$baseline.temp_v1
v1$vel <- v1$vel_v1
v1$version = "v1"

v3$baseline.temp <- v3$baseline.temp_v3
v3$vel <- v3$vel_v3
v3$version = "v3"

tmp <- rbind(v1[,c("ID","Type","ECO","vel","baseline.temp","SHIFT","version")],
             v3[,c("ID","Type","ECO","vel","baseline.temp","SHIFT","version")])

# All values
ggplot(tmp, aes(x = vel, fill = version))+
    geom_histogram(alpha = .5)+
    theme_classic()+
    labs(x = "Velocity")

## Terrestrials
### Latitude
ggplot(tmp %>% filter(ECO=="T",
                      Type == "LAT"), 
       aes(x = vel, fill = version))+
    ggtitle("Terrestrials\nLatitude")+
    geom_histogram(alpha = .5)+
    theme_classic()+
    labs(x = "Velocity")

### Elevation
ggplot(tmp %>% filter(ECO=="T",
                      Type == "ELE"), 
       aes(x = vel, fill = version))+
    ggtitle("Terrestrials\nElevation")+
    geom_histogram(alpha = .5)+
    theme_classic()+
    labs(x = "Velocity")

## Marine
ggplot(tmp %>% filter(ECO=="M"), 
       aes(x = vel, fill = version))+
    ggtitle("Marine")+
    geom_histogram(alpha = .5)+
    theme_classic()+
    labs(x = "Velocity")

```

V3 velocity values are mainly negative (towards the tropics).

## Velocity scatter plots
```{r}

# All data
ggplot(Bioshifts_vel, aes(x = vel_v3, y = vel_v1)) +
    geom_point()+
    theme_classic()+
    labs(x = "Velocity v3", y = "Velocity v1")

# Terrestrial
ggplot(Bioshifts_vel %>% filter(ECO == "T"), 
       aes(x = vel_v3, y = vel_v1)) +
    ggtitle("Terrestrials") + 
    geom_point()+
    theme_classic()+
    facet_wrap(.~Type, scales = "free") +
    labs(x = "Velocity v3", y = "Velocity v1")

# Marine
ggplot(Bioshifts_vel %>% filter(ECO == "M"), 
       aes(x = vel_v3, y = vel_v1)) +
    ggtitle("Marine") + 
    geom_point()+
    theme_classic()+
    facet_wrap(.~Type, scales = "free") +
    labs(x = "Velocity v3", y = "Velocity v1")
```

Velocity values from v3 and v1 are highly decoupled, and tend to be negatively correlated to each other.

## Trends
```{r}
# all data
ggplot(Bioshifts_vel, aes(x = baseline.temp_v3, y = baseline.temp_v1)) +
    geom_point()+
    theme_classic()+
    facet_wrap(.~Type, scales = "free") +
    labs(x = "Trend v3", y = "Trend v1")

# Terrestrial
ggplot(Bioshifts_vel %>% filter(ECO == "T"), 
       aes(x = baseline.temp_v3, y = baseline.temp_v1)) +
    ggtitle("Terrestrials") + 
    geom_point()+
    theme_classic()+
    facet_wrap(.~Type, scales = "free") +
    labs(x = "Trend v3", y = "Trend v1")

# Marine
ggplot(Bioshifts_vel %>% filter(ECO == "M"), 
       aes(x = baseline.temp_v3, y = baseline.temp_v1)) +
    ggtitle("Marine") + 
    geom_point()+
    theme_classic()+
    facet_wrap(.~Type, scales = "free") +
    labs(x = "Trend v3", y = "Trend v1")
```

When comparing trend values between v1 and v3, they seem to be correlated. This means that the decoupling in velocity values between v1 and v3 comes from differences in the way the spatial gradient is calculated.

## Velocity vs shift
```{r}

# Velocity v1 vs Shift
ggplot(Bioshifts_vel, 
       aes(x = vel_v1, y = SHIFT)) +
    ggtitle("Terrestrial") + 
    geom_point()+
    theme_classic()+
    facet_wrap(ECO~Type, scales = "free") +
    labs(x = "Velocity v1", y = "SHIFT v1")

# Velocity v3 vs Shift
ggplot(Bioshifts_vel, 
       aes(x = vel_v3, y = SHIFT)) +
    ggtitle("Terrestrial") + 
    geom_point()+
    theme_classic()+
    facet_wrap(ECO~Type, scales = "free") +
    labs(x = "Velocity v3", y = "SHIFT v1")

```


As expected based on the decoupling between velocity values in v1 and v3, the effect of velocity on observed range shift can be quite different depending on version of bioshifts used to extract velocity values.

## OLS results: Velocity vs Shift
```{r fig.width=10, fig.height=3}

# Terrestrial + Lat
shift_vel_v1_ter_lat <- summary(lm(SHIFT~vel_v1,
                                   Bioshifts_vel %>%
                                       filter(ECO=="T" & Type == "LAT")))

shift_vel_v3_ter_lat <- summary(lm(SHIFT~vel_v3,
                                   Bioshifts_vel %>%
                                       filter(ECO=="T" & Type == "LAT")))

# Terrestrial + Ele
shift_vel_v1_ter_ele <- summary(lm(SHIFT~vel_v1,
                                   Bioshifts_vel %>%
                                       filter(ECO=="T" & Type == "ELE")))

shift_vel_v3_ter_ele <- summary(lm(SHIFT~vel_v3,
                                   Bioshifts_vel %>%
                                       filter(ECO=="T" & Type == "ELE" & !is.infinite(vel_v3))))

# Marine
shift_vel_v1_mar <- summary(lm(SHIFT~vel_v1,
                               Bioshifts_vel %>%
                                   filter(ECO=="M")))

shift_vel_v3_mar <- summary(lm(SHIFT~vel_v3,
                               Bioshifts_vel %>%
                                   filter(ECO=="M")))

models <- list(
    shift_vel_v1_ter_lat,
    shift_vel_v3_ter_lat,
    shift_vel_v1_ter_ele,
    shift_vel_v3_ter_ele,
    shift_vel_v1_mar,
    shift_vel_v3_mar
)
names(models) <- c("shift_vel_v1_ter_lat",
                   "shift_vel_v3_ter_lat",
                   "shift_vel_v1_ter_ele",
                   "shift_vel_v3_ter_ele",
                   "shift_vel_v1_mar",
                   "shift_vel_v3_mar")

Coeffs <- lapply(models, function(x){
    data.frame(x$coefficients)[2,]
})
Coeffs <- rbindlist(Coeffs)

R2 <- sapply(models, function(x){
    x$r.squared
})

# group all
ols <- data.frame(Coeffs, 
                  R2,
                  model = names(models)
)
ols$version = ifelse(grepl("v1",ols$model),"v1","v3")
ols$Type = ifelse(grepl("ele",ols$model),"ELE","LAT")
ols$ECO = ifelse(grepl("ter",ols$model),"TER","MAR")


ggplot(ols, aes(y = model, x = Estimate, fill = version))+
    geom_col()+
    theme_classic() +
    facet_wrap(ECO~Type, scales = "free")

ggplot(ols, aes(y = model, x = R2, fill = version))+
    geom_col()+
    theme_classic() +
    facet_wrap(ECO~Type, scales = "free")

```

Coefficients tend to be stepper when using velocities v3 relative to v1. However, the coefficient in negative at the marine realm.

In general, I believe that velocities at the landscape scale are not helpful to explain observed range shifts. We might need to work better metrics, perhaps using sdms to apply weights on landscape scale velocity based on suitability. This way, we have a species level, rather than landscape level velocity.

A next step we made was to  test if the metric of velocity calculated in Bioshifts v3 matches with velocity patterns shown in previous studies. This is crutial for us to test if we can trust the metric. Therefore, we calculated velocity at global scale at the marine realm and compare with a published global map.

# Global velocity maps
```{r}
# Undirectional 
gVel <- terra::rast(here::here("Data/Mar_gVel.tif"))

gVelrange <- range(gVel[],na.rm = TRUE)
the_palette_fc <- leaflet::colorNumeric(
    palette = "RdBu", 
    domain = c(-max(abs(gVelrange)),max(abs(gVelrange))),
    reverse = TRUE)
the_colors <- the_palette_fc(seq(min(gVelrange), max(gVelrange), length.out = 50))

plot(gVel, col = the_colors)
```


The climate gradient that we calculate using our function is very similar to what has been presented in previous studies. For example, see Fig 2c from https://www.nature.com/articles/s41558-020-0889-7.

```{r}
ttrend <- terra::rast(here::here("Data/Mar_mean_trend.tif"))

ttrendrange <- range(ttrend[],na.rm = TRUE)
the_palette_fc <- leaflet::colorNumeric(
    palette = "RdBu", 
    domain = c(-max(abs(ttrendrange)),max(abs(ttrendrange))),
    reverse = TRUE)
the_colors <- the_palette_fc(seq(min(ttrendrange), max(ttrendrange), length.out = 50))

plot(ttrend, col = the_colors)
```

The temporal trend is that we calculate using our function is also very similar to what has been presented in previous studies. For example, see Fig 2a from https://www.nature.com/articles/s41558-020-0889-7.


When we convert angles to calculate the velocity across the latitudinal gradient,  velocity values become quite different from the multidirectional velocity gradient showed above.

```{r}
# Latitude 
gVelLat <- terra::rast(here::here("Data/Mar_gVelLat.tif"))

gVelLatrange <- range(gVelLat[],na.rm = TRUE)
the_palette_fc <- leaflet::colorNumeric(
    palette = "RdBu", 
    domain = c(-max(abs(gVelLatrange)),max(abs(gVelLatrange))),
    reverse = TRUE)
the_colors <- the_palette_fc(seq(min(gVelLatrange), max(gVelLatrange), length.out = 50))

plot(gVelLat, col = the_colors)
```


The difference is due to the angle of the gradient. Take a closer look to the direction of climate change.
```{r}
# Latitude angle
gVelAngle <- terra::rast(here::here("Data/Mar_gVelAngle.tif"))

{layout(matrix(c(1,1,1,2), nrow = 1, ncol = 4, byrow=T))
    plot(gVelAngle, col = colorRampPalette(colors = c("blue", "red","green","yellow","blue"))(360))
    plotrix::polar.plot(lengths = c(rnorm(360,mean = 1, sd = 0.001)),
                        polar.pos = seq(0,360,by=1),
                        cex=.01,
                        show.grid.labels=0,
                        line.col=colorRampPalette(colors = c("blue", "red","green","yellow","blue"))(360))
}
```

Observed that many trajectories are West (blue) and East (green) and just a few are North (red), or South (yellow). The West and East trajectories ultimately will results in slower South-North (at the North hemisphere) velocity (clear "white" colors in the latitudinal velocity map above). 

Plotting trajetory lines
```{r}
# Trajectories

# coarser resolution raster 
gVelAngle2 <- terra::aggregate(gVelAngle, fact = 20)
    
## get the set of starting cells for the trajectories
lonlat <- na.omit(as.data.frame(gVelAngle2, xy=TRUE))[,1:2]

    
## calculate trajectory:
traj = voccTraj(lonlat, # starting lat lon coords
                raster::aggregate(raster::raster(gVel),20), # velocity 
                raster::aggregate(raster::raster(gVelAngle),20), # angle
                raster::aggregate(raster::raster("Data/Mar_mean_avg_climate_layers.tif"),20), # mean temp 
                tyr = 40 # length of period of interest, years
                )

## create a spatial line data frame from traj
lns <- trajLine(x = traj)
plot(lns)

```

