---
title: "Get species data for SDMs"
author: "Brunno F Oliveira & Bioshifts group"
date: "Last compiled on `r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
---

# *Setup*

```{r setup, message = FALSE, warning = FALSE}

rm(list=ls())
gc()

list.of.packages <- c("stringr", "rgbif", "parallel", "pbapply", "raster", "dplyr", "here", "tidyr","ggrepel",
                      "data.table", "ggplot2", "readxl", "knitr",
                      "sqldf","RSQLite", "scales")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

if(length(new.packages)) install.packages(new.packages)

sapply(list.of.packages, require, character.only = TRUE)

```

# *Load parameters*

```{r}

source(here::here("R/settings.R"))

range_env_data <- sapply(temporal_range_env_data, function(x){
    c(min(x),max(x))
})
range_env_data <- c(min(range_env_data),max(range_env_data))

```

# *Load functions*

```{r }

source(here::here("R/my_functions.R"))

```

# *Load species list*

```{r}

splist <- read.csv(here::here("Data/splist.csv"), header = T)
# remove duplicated sp_names
splist <- splist %>%
    dplyr::select(scientificName,kingdom,phylum,class,order,family,db) %>%
    filter(!duplicated(scientificName))

length(unique(splist$scientificName))

```

# *Load bioshifts*

*While we dont have a v3, just add papers from v2 to v1*

```{r }

# Load v3
biov3 <- read.csv(here::here("Data/bioshifts_v3_raw.csv"))
biov3 <- biov3 %>%
    filter(!is.na(sp_name)) %>%
    mutate(sp_name=gsub("_"," ",sp_name))
# add taxonomy
tojoin <- splist %>%
    dplyr::select(scientificName, kingdom, phylum, class, order, family) %>%
    mutate(sp_name = scientificName) %>%
    filter(!duplicated(sp_name))

any(!biov3$sp_name %in% tojoin$sp_name)

biov3 <- merge(biov3, tojoin, by = "sp_name", all.x = TRUE)

length(unique(biov3$scientificName))

# Load v1
biov1 <- read.csv(here::here("Data/biov1_fixednames.csv"), header = T)
# Fix references in biov1
biov1$Article_ID <- gsub("[^0-9.-]", "", biov1$Article_ID)
biov1$Article_ID <- as.numeric(biov1$Article_ID)
biov1$sp_name_std_v1 <- gsub("_"," ",biov1$sp_name_std_v1)
biov1 <- biov1 %>%
    dplyr::select(ID,Study_ID,Type,Param,Trend,SHIFT,UNIT,DUR,
                  START,END,group,ECO,
                  Phylum,Class,Order,Family,Genus,sp_name_std_v1) %>% # select columns
    mutate(
        Type = case_when(
            Type=="HOR" ~ "LAT",
            TRUE ~ as.character(Type)),
        Species = sp_name_std_v1) %>% 
    mutate( # from m/year to km/year
        kmyr = case_when(
            UNIT == "m/year" ~ SHIFT/1000, 
            UNIT == "km/year" ~ SHIFT, 
            TRUE ~ NA_real_))

biov1 <- biov1 %>%
    filter(!is.na(sp_name_std_v1))

all(biov1$sp_name_std_v1 %in% splist$scientificName[which(splist$v1==1)])

# # "T" "M"
# unique(biov1$group)
# # "vertebrate"         "vascular.plant"     "other.animal"  
# # "fungi"              "algae"              "virus" 
# # "bacteria"           "non.vascular.plant"

# Fishes are those from Class Actinopterygii
# Class Actinopterygii + ECO T = freshwater fish
# Class Actinopterygii + ECO M = marine fish
# Sharks are those from Class Chondrichthyes

# Load v2 
biov2 <- read.csv(here::here("Data/biov2_fixednames.csv"), header = T)
biov2$Scientific.Name <- gsub(" ","_",biov2$Scientific.Name)
biov2$sp_name_std_v2 <- gsub("_"," ",biov2$sp_name_std_v2)
biov2 <- biov2  %>%
    dplyr::select(Paper.ID,Study.Period,Dimension,Parameter,Numeric.Change,Unit,Total.Years.Studied,
                  Start.Year,End.Year,Taxonomic.Group,Ecosystem.Type,
                  phylum,class,order,family,genus,sp_name_std_v2) %>% # select columns
    filter(!Parameter == "west" | !Parameter == "east" | !is.na(Numeric.Change)) %>%
    mutate(
        Type = case_when(
            Dimension=="latitude" ~ "LAT",
            Dimension=="elevation" ~ "ELE",
            TRUE ~ as.character(Dimension)),
        ID=Paper.ID,
        Study_ID=Study.Period,
        Type=Type,
        Param=Parameter,
        Trend=Numeric.Change,
        UNIT=Unit,
        DUR=Total.Years.Studied,
        START=Start.Year,
        END=End.Year,
        group=Taxonomic.Group,
        ECO=case_when(
            Ecosystem.Type=="terrestrial" ~ "T",
            Ecosystem.Type=="marine" ~ "M",
            Ecosystem.Type=="aquatic" ~ "A",
            TRUE ~ as.character(Ecosystem.Type)),
        Phylum=phylum,
        Class=class,
        Order=order,
        Family=family,
        Genus=genus,
        Species=sp_name_std_v2) %>%
    mutate(
        Type = case_when(
            Dimension=="latitude" ~ "LAT",
            Dimension=="elevation" ~ "ELE",
            TRUE ~ as.character(Dimension)),
        Param = case_when(
            Parameter=="leading edge" ~ "LE",
            Parameter=="maximum/optimum" ~ "O",
            Parameter=="trailing edge" ~ "TE",
            Parameter=="mean" ~ "O",
            TRUE ~ as.character(Parameter))) %>%
    mutate(Total.Years.Studied=as.numeric(Total.Years.Studied)) %>% 
    mutate(
        kmyr = case_when(
            UNIT == "ft" ~ Trend/3280.8 /Total.Years.Studied,
            UNIT == 'degrees lat' ~ Trend*111/Total.Years.Studied,
            UNIT =='degrees lat/year' ~ Trend*111,
            UNIT=='degree latitude per decade'~ Trend*111/10,
            UNIT=='km' ~ Trend/Total.Years.Studied,
            UNIT=='km/decade' ~ Trend/10,
            UNIT=='km/year' ~ Trend,
            UNIT=='m' ~ Trend/1000/Total.Years.Studied,
            UNIT=='m/year' ~ Trend/1000,
            UNIT=='m/decade'~ Trend/1000/10,
            TRUE ~ NA_real_
        )
    )

biov2 <- biov2 %>%
    filter(!is.na(sp_name_std_v2))

all(biov2$biov2 %in% splist$scientificName[which(splist$v2==1)])

# unique(biov2$ECO)
# # "terrestrial" "marine"      "aquatic"
# unique(biov2$group)
# # "Bird"                                "Insect"                             
# # "Amphibian"                           "Mammal"                             
# # "Spider"                              "Plant"                              
# # "Sea anemones and corals"             "Polychaetes"                        
# # "Molluscs"                            "NA"                                 
# # "Crustacean"                          "Starfish"                           
# # "Ascidians tunicates and sea squirts" "Fish"                               
# # "Sea urchin"                          "Crinoid"                            
# # "Sea cucumber"                        "Reptile"                            
# # "Brittle stars"                       "Centipedes"                         
# # "Millipedes"                          "Hydrozoa" 

```

## *Remove species identified to the genus level or cf.*

```{r }

if(any(grep("sp[.]",biov1$sp_reported_name_v1))){
    biov1 <- biov1 %>% filter(!grepl("sp[.]",sp_reported_name_v1))
}
if(any(grep("sp[.]",biov1$sp_name_std_v1))){
    biov1 <- biov1 %>% filter(!grepl("sp[.]",sp_name_std_v1))
}
if(any(grep("cf[.]",biov1$sp_reported_name_v1))){
    biov1 <- biov1 %>% filter(!grepl("cf[.]",sp_reported_name_v1))
}
if(any(grep("cf[.]",biov1$sp_name_std_v1))){
    biov1 <- biov1 %>% filter(!grepl("cf[.]",sp_name_std_v1))
}

if(any(grep("sp[.]",biov2$sp_reported_name_v2))){
    biov2 <- biov2 %>% filter(!grepl("sp[.]",sp_reported_name_v2))
}
if(any(grep("sp[.]",biov2$sp_name_std_v2))){
    biov2 <- biov2 %>% filter(!grepl("sp[.]",sp_name_std_v2))
}
if(any(grep("cf[.]",biov2$sp_reported_name_v2))){
    biov2 <- biov2 %>% filter(!grepl("cf[.]",sp_reported_name_v2))
}
if(any(grep("cf[.]",biov2$sp_name_std_v2))){
    biov2 <- biov2 %>% filter(!grepl("cf[.]",sp_name_std_v2))
}


if(any(grep("sp[.]",biov3$sp_reported_name_v3))){
    biov3 <- biov3 %>% filter(!grepl("sp[.]",sp_reported_name_v3))
}
if(any(grep("sp[.]",biov3$sp_name_std_v3))){
    biov3 <- biov3 %>% filter(!grepl("sp[.]",sp_name_std_v3))
}
if(any(grep("cf[.]",biov3$sp_reported_name_v3))){
    biov3 <- biov3 %>% filter(!grepl("cf[.]",sp_reported_name_v3))
}
if(any(grep("cf[.]",biov3$sp_name_std_v3))){
    biov3 <- biov3 %>% filter(!grepl("cf[.]",sp_name_std_v3))
}


if(any(grep("sp[.]",splist$scientificName))){
    splist <- splist %>% filter(!grepl("sp[.]",scientificName))
}
if(any(grep("sp[.]",splist$scientificName))){
    splist <- splist %>% filter(!grepl("sp[.]",scientificName))
}
if(any(grep("cf[.]",splist$scientificName))){
    splist <- splist %>% filter(!grepl("cf[.]",scientificName))
}
if(any(grep("cf[.]",splist$scientificName))){
    splist <- splist %>% filter(!grepl("cf[.]",scientificName))
}
```

## *Classify species*

```{r }

# Class species as Terrestrial, Marine or Freshwater
Terv1 <- unique(biov1$sp_name_std_v1[which(biov1$ECO == "T")])
Terv2 <- unique(biov2$sp_name_std_v2[which(biov2$ECO == "T")])
Terrestrials = unique(c(Terv1,Terv2))

Mar1 <- unique(biov1$sp_name_std_v1[which(biov1$ECO == "M")])
Mar2 <- unique(biov2$sp_name_std_v2[which(biov2$ECO == "M")])
Marine = unique(c(Mar1,Mar2))

Aquatic = unique(biov2$sp_name_std_v2[which(biov2$ECO == "A")])

# Freshwater fish
FFishv1 <- unique(biov1$sp_name_std_v1[(biov1$Class == "Actinopterygii" | biov1$Class == "Cephalaspidomorphi") & biov1$ECO == "T"])
FFishv2 <- unique(biov2$sp_name_std_v2[biov2$group == "Fish" & 
                                           (biov2$ECO == "T" | biov2$ECO == "A")])
FFish = unique(c(FFishv1,FFishv2))
# Marine fish
MFishv1 <- unique(biov1$sp_name_std_v1[biov1$Class == "Actinopterygii" & biov1$ECO == "M"])
MFishv2 <- unique(biov2$sp_name_std_v2[biov2$group == "Fish" & biov2$ECO == "marine"])
MFish = unique(c(MFishv1,MFishv2))

splist$ECO = NA
splist$ECO[which(splist$scientificName %in% Terrestrials)] <- "T"
splist$ECO[which(splist$scientificName %in% Marine)] <- "M"
splist$ECO[which(splist$scientificName %in% Aquatic)] <- "A"
splist$ECO[which(splist$scientificName %in% MFish)] <- "M"
splist$ECO[which(splist$scientificName %in% FFish)] <- "A"

biov1$ECO = NA
biov1$ECO[which(biov1$sp_name_std_v1 %in% Terrestrials)] <- "T"
biov1$ECO[which(biov1$sp_name_std_v1 %in% Marine)] <- "M"
biov1$ECO[which(biov1$sp_name_std_v1 %in% Aquatic)] <- "A"
biov1$ECO[which(biov1$sp_name_std_v1 %in% MFish)] <- "M"
biov1$ECO[which(biov1$sp_name_std_v1 %in% FFish)] <- "A"

biov2$ECO = NA
biov2$ECO[which(biov2$sp_name_std_v2 %in% Terrestrials)] <- "T"
biov2$ECO[which(biov2$sp_name_std_v2 %in% Marine)] <- "M"
biov2$ECO[which(biov2$sp_name_std_v2 %in% Aquatic)] <- "A"
biov2$ECO[which(biov2$sp_name_std_v2 %in% MFish)] <- "M"
biov2$ECO[which(biov2$sp_name_std_v2 %in% FFish)] <- "A"

biov3$ECO = NA
biov3$ECO[which(biov3$sp_name_std_v3 %in% Terrestrials)] <- "T"
biov3$ECO[which(biov3$sp_name_std_v3 %in% Marine)] <- "M"
biov3$ECO[which(biov3$sp_name_std_v3 %in% Aquatic)] <- "A"
biov3$ECO[which(biov3$sp_name_std_v3 %in% MFish)] <- "M"
biov3$ECO[which(biov3$sp_name_std_v3 %in% FFish)] <- "A"

splist$Group = NA
splist$Group[which(splist$class == "Phaeophyceae")] <- "Chromista"
splist$kingdom[which(splist$class == "Phaeophyceae")] <- "Chromista"
splist$Group[which(splist$phylum == "Rhodophyta")] <- "Seaweed"
splist$kingdom[which(splist$phylum == "Rhodophyta")] <- "Plantae"
splist$Group[which(splist$family == "Elminiidae")] <- "Barnacles"
splist$Group[which(splist$kingdom == "Bacteria")] <- "Bacteria"
splist$Group[which(splist$class == "Holothuroidea")] <- "Sea cucumber"
splist$Group[which(splist$class == "Aves")] <- "Bird"
splist$Group[which(splist$class == "Insecta")] <- "Insect"
splist$Group[which(splist$class == "Mammalia")] <- "Mammal"
splist$Group[which(splist$class == "Arachnida")] <- "Spider"
splist$kingdom[which(splist$kingdom == "Viridiplantae")] <- "Plantae"
splist$kingdom[which(splist$phylum == "Tracheophyta")] <- "Plantae"
splist$Group[which(splist$kingdom == "Plantae")] <- "Plant"
splist$Group[which(splist$class == "Hydrozoa")] <- "Hydrozoa"
splist$Group[which(splist$class == "Anthozoa")] <- "Sea anemones and corals"
splist$Group[which(splist$class == "Polychaeta")] <- "Polychaetes"
splist$Group[which(splist$phylum == "Mollusca")] <- "Molluscs"
splist$Group[which(splist$class == "Malacostraca")] <- "Crustacean"
splist$Group[which(splist$class == "Hexanauplia")] <- "Crustacean"
splist$Group[which(splist$class == "Maxillopoda")] <- "Crustacean"
splist$Group[which(splist$class == "Ostracoda")] <- "Crustacean"
splist$Group[which(splist$class == "Branchiopoda")] <- "Crustacean"
splist$Group[which(splist$class == "Asteroidea")] <- "Starfish"
splist$Group[which(splist$class == "Ascidiacea")] <- "Ascidians tunicates and sea squirts"
splist$class[which(splist$class == "Actinopteri")] <- "Actinopterygii"
splist$Group[which(splist$class == "Actinopterygii")] <- "Fish"
splist$Group[which(splist$class == "Elasmobranchii")] <- "Fish"
splist$Group[which(splist$order == "Perciformes")] <- "Fish"
splist$Group[which(splist$class == "Chondrichthyes")] <- "Fish"
splist$Group[which(splist$class == "Holocephali")] <- "Fish"
splist$Group[which(splist$class == "Cephalaspidomorphi")] <- "Fish"
splist$Group[which(splist$class == "Echinoidea")] <- "Sea urchin"
splist$Group[which(splist$class == "Crinoidea")] <- "Crinoid"
splist$Group[which(splist$class == "Holothuroidea")] <- "Sea cucumber"
splist$Group[which(splist$class == "Reptilia")] <- "Reptile"
splist$Group[which(splist$order == "Squamata")] <- "Reptile"
splist$Group[which(splist$class == "Ophiuroidea")] <- "Brittle stars"
splist$Group[which(splist$class == "Chilopoda")] <- "Centipedes"
splist$Group[which(splist$class == "Diplopoda")] <- "Millipedes"
splist$Group[which(splist$class == "Amphibia")] <- "Amphibian"
splist$Group[which(splist$kingdom == "Fungi")] <- "Fungi"
splist$Group[which(splist$order == "Balanomorpha")] <- "Barnacles"
splist$Group[which(splist$phylum == "Nematoda")] <- "Nematodes"
splist$Group[which(splist$class == "Myxini")] <- "Hagfish"
splist$Group[which(splist$kingdom == "Chromista")] <- "Chromista"

######################################
biov1 <- merge(biov1[,-which(names(biov1) %in% c("Group","ECO"))], 
               splist[,c("Group","ECO","scientificName")], 
               by.x = "sp_name_std_v1", by.y = "scientificName", 
               all.x = T)

biov2 <- merge(biov2[,-which(names(biov2) %in% c("Group","ECO"))], splist[,c("Group","ECO","scientificName")], 
               by.x = "sp_name_std_v2", by.y = "scientificName", 
               all.x = T)

biov3 <- merge(biov3[,-which(names(biov3) %in% c("Group","ECO"))], splist[,c("Group","ECO","scientificName")], 
               by.x = "sp_name", by.y = "scientificName", 
               all.x = T)

```

# *Populate v3 columns*

```{r}

# Trend
biov3$Trend <- as.numeric(biov3$Trend_v1)
pos <- which(is.na(biov3$Trend))
biov3$Trend[pos] <- biov3$Trend_v2[pos]

# Unit trend
biov3$Unit.trend <- biov3$Unit.trend_v1
pos <- which(is.na(biov3$Unit.trend))
biov3$Unit.trend[pos] <- biov3$Unit.trend_v2[pos]

# Start
biov3$START <- biov3$START_v1
pos <- which(is.na(biov3$START))
biov3$START[pos] <- biov3$START_v2[pos]

# End
biov3$END <- biov3$END_v1
pos <- which(is.na(biov3$END))
biov3$END[pos] <- biov3$END_v2[pos]

# Duration
biov3$DUR <- biov3$END-biov3$START+1

# Trend km (standardized)
unique(biov3$Unit.trend)
biov3 <- biov3 %>%
    mutate(
        Trend_km = as.numeric(Trend),
        DUR = as.numeric(DUR),
        Trend_km = case_when(
            Unit.trend == "degree" ~ Trend_km*111,
            Unit.trend == "degrees lat" ~ Trend_km*111,
            Unit.trend =='degree/year' ~ (Trend_km*111)*DUR,
            Unit.trend=='degrees lat/year'~ (Trend_km*111)*DUR,
            Unit.trend=='degree/decade'~ (Trend_km*111/10)*DUR,
            Unit.trend=='degree latitude per decade'~ (Trend_km*111/10)*DUR,
            Unit.trend=='km' ~ Trend_km,
            Unit.trend=='km/decade' ~ (Trend_km/10)*DUR,
            Unit.trend=='km/year' ~ Trend_km*DUR,
            Unit.trend=='m' ~ Trend_km/1000,
            Unit.trend=='m/year' ~ (Trend_km/1000)*DUR,
            Unit.trend=='m/decade'~ (Trend_km/1000/10)*DUR
        )
    )

# SHIFT (shift values come standardized)
biov3$SHIFT <- biov3$SHIFT_v1
pos <- which(is.na(biov3$SHIFT))
biov3$SHIFT[pos] <- biov3$SHIFT_v2[pos]

# remove categorical shifts
biov3 <- biov3 %>%
    mutate(SHIFT=as.numeric(SHIFT)) %>%
    filter(!is.na(SHIFT))

```

## *Filter LAT / ELE shifts*

```{r}

# v1
biov1 <- biov1 %>%
    filter(Type %in% c("ELE","LAT"))  # Use LAT ELE shifts
# splist
sps <- unique(biov1$sp_name)
splist <- splist %>% filter(scientificName %in% sps)

# v2
biov2 <- biov2 %>%
    filter(Type %in% c("ELE","LAT"))  # Use LAT ELE shifts
# splist
sps <- unique(biov2$sp_name)
splist <- splist %>% filter(scientificName %in% sps)

# v3
biov3 <- biov3 %>%
    filter(Type %in% c("ELE","LAT"))  # Use LAT ELE shifts
# splist
sps <- unique(biov3$sp_name)
splist <- splist %>% filter(scientificName %in% sps)

```

# *Stats before subsetting*

## *How many species in Bioshifts ?*

```{r}

# N species
tmp <- data.frame(sps = biov3$sp_name,
                  ECO = biov3$ECO,
                  Group = biov3$Group)
tmp$v1 <- NA
tmp$v1 <- ifelse(is.na(biov3$SHIFT_v1),0,1)

tmp$v2 <- NA
tmp$v2 <- ifelse(is.na(biov3$SHIFT_v2),0,1)

tmp$version <- NA
tmp$version[which(tmp$v1==1 & tmp$v2==1)] <- "both"
tmp$version[which(tmp$v1==1 & tmp$v2==0)] <- "v1"
tmp$version[which(tmp$v1==0 & tmp$v2==1)] <- "v2"

tmp %>% summarise(N = length(unique(sps))) 

# N shifts
nrow(tmp)

nrow(biov1)

# N species
length(unique(tmp$sps))


# N species by group
toplot <- 
    tmp %>%
    group_by(Group) %>%
    summarise(N = length(unique(sps)))  %>%
    arrange(desc(N))

nice_table(toplot)

g_ord <- unique(toplot$Group[order(toplot$N)])
toplot$Group <- factor(toplot$Group, levels = g_ord)

ggplot(toplot, aes(x = Group, y = N)) +
    geom_bar(stat="identity")+
    geom_text(aes(y=N, label=N, 
                  hjust = ifelse(N < max(toplot$N), -0.1, 1.1)), vjust=0.2, size=3, 
              position = position_dodge(0.9))+
    theme_classic()+
    coord_flip()

toplot <- 
    tmp %>%
    group_by(Group,version) %>%
    summarise(N = length(unique(sps)))  %>%
    arrange(desc(N))

toplot$Group <- factor(toplot$Group, levels = g_ord)

ggplot(toplot, aes(x = Group, y = N, fill=version)) +
    geom_bar(position="stack", stat="identity")+
    theme_classic()+
    coord_flip()


# N species by ECO
tmp %>%
    group_by(ECO) %>%
    summarise(N = length(unique(sps)))  

# N species by group
toplot <- 
    tmp %>%
    group_by(Group, ECO) %>%
    summarise(N = length(unique(sps)))  %>%
    arrange(desc(N))

nice_table(toplot)

toplot$Group <- factor(toplot$Group, levels = unique(toplot$Group[order(toplot$N)]))
ggplot(toplot, aes(x = Group, y = N, )) +
    geom_bar(stat="identity")+
    geom_text(aes(y=N, label=N, 
                  hjust = ifelse(N < max(toplot$N), -0.1, 1.1)), vjust=0.2, size=3, 
              position = position_dodge(0.9))+
    theme_classic()+
    coord_flip()+
    facet_wrap(.~ECO)

# N species
tmp <- data.frame(sps = biov3$sp_name,
                  Type = biov3$Type,
                  Param = biov3$Param,
                  ECO = biov3$ECO,
                  Group = biov3$Group)

# N species by group
toplot <- 
    tmp %>%
    group_by(ECO, Group) %>%
    summarise(N = length(unique(sps))) %>%
    mutate(Group = factor(Group, levels = Group[order(N)]))

ggplot(toplot, aes(x = ECO, y = N, fill = Group, label = N)) +
    geom_bar(stat = "identity") +
    geom_text(size = 3, position = position_stack(vjust = 0.5))+
    facet_wrap(.~ECO, scales = "free")
```

## *How many shifts in Bioshifts ?*

```{r}

# N species
tmp <- data.frame(sps = biov3$sp_name,
                  ECO = biov3$ECO,
                  Group = biov3$Group)
tmp$v1 <- NA
tmp$v1 <- ifelse(is.na(biov3$SHIFT_v1),0,1)

tmp$v2 <- NA
tmp$v2 <- ifelse(is.na(biov3$SHIFT_v2),0,1)

tmp$version <- NA
tmp$version[which(tmp$v1==1 & tmp$v2==1)] <- "both"
tmp$version[which(tmp$v1==1 & tmp$v2==0)] <- "v1"
tmp$version[which(tmp$v1==0 & tmp$v2==1)] <- "v2"

tmp %>% summarise(N = length(unique(sps))) 

# N shifts
nrow(tmp)

# N species by group
toplot <- 
    tmp %>%
    group_by(Group) %>%
    summarise(N = length((sps)))  %>%
    arrange(desc(N))

nice_table(toplot)

g_ord <- unique(toplot$Group[order(toplot$N)])
toplot$Group <- factor(toplot$Group, levels = g_ord)

ggplot(toplot, aes(x = Group, y = N)) +
    geom_bar(stat="identity")+
    geom_text(aes(y=N, label=N, 
                  hjust = ifelse(N < max(toplot$N), -0.1, 1.1)), vjust=0.2, size=3, 
              position = position_dodge(0.9))+
    theme_classic()+
    coord_flip()

# N species by ECO
table(tmp$ECO)

tmp %>%
    filter(ECO=="T") %>%
    group_by(Group) %>%
    summarise(N = length(sps),
              Perc = round(length(sps)/length(which(tmp$ECO=="T")),2))  %>%
    arrange(desc(N)) %>%
    nice_table()

tmp %>%
    filter(ECO=="M") %>%
    group_by(Group) %>%
    summarise(N = length(sps),
              Perc = round(length(sps)/length(which(tmp$ECO=="M")),2))  %>%
    arrange(desc(N)) %>%
    nice_table()

tmp %>%
    filter(ECO=="T") %>%
    group_by(Group) %>%
    summarise(N = length(sps),
              Perc = round(length(sps)/length(which(tmp$ECO=="A")),2))  %>%
    arrange(desc(N)) %>%
    nice_table()

# by version
toplot <- 
    tmp %>%
    group_by(Group,version) %>%
    summarise(N = length(unique(sps)))  %>%
    arrange(desc(N))

toplot$Group <- factor(toplot$Group, levels = g_ord)

ggplot(toplot, aes(x = Group, y = N, fill=version)) +
    geom_bar(position="stack", stat="identity")+
    theme_classic()+
    coord_flip()


# N species by ECO
tmp %>%
    group_by(ECO) %>%
    summarise(N = length((sps)))  

# N species by group
toplot <- 
    tmp %>%
    group_by(Group, ECO) %>%
    summarise(N = length((sps)))  %>%
    arrange(desc(N))

nice_table(toplot)

toplot$Group <- factor(toplot$Group, levels = unique(toplot$Group[order(toplot$N)]))
ggplot(toplot, aes(x = Group, y = N, )) +
    geom_bar(stat="identity")+
    geom_text(aes(y=N, label=N, 
                  hjust = ifelse(N < max(toplot$N), -0.1, 1.1)), vjust=0.2, size=3, 
              position = position_dodge(0.9))+
    theme_classic()+
    coord_flip()+
    facet_wrap(.~ECO)

# N species
tmp <- data.frame(sps = biov3$sp_name,
                  Type = biov3$Type,
                  Param = biov3$Param,
                  ECO = biov3$ECO,
                  Group = biov3$Group)

# N species by group
toplot <- 
    tmp %>%
    group_by(ECO, Group) %>%
    summarise(N = length(unique(sps))) %>%
    mutate(Group = factor(Group, levels = Group[order(N)]))

ggplot(toplot, aes(x = ECO, y = N, fill = Group, label = N)) +
    geom_bar(stat = "identity") +
    geom_text(size = 3, position = position_stack(vjust = 0.5))+
    facet_wrap(.~ECO, scales = "free")
```

## *Shift types*

```{r}

# v1
toplot <- biov1 %>%
    group_by(Type) %>%
    summarise(N = length(Type),
              percent = N/nrow(biov1))

df2 <- toplot %>% 
    mutate(csum = rev(cumsum(rev(percent))), 
           pos = percent/2 + lead(csum, 1),
           pos = if_else(is.na(pos), percent/2, pos))

ggplot(df2, aes(x = "", y = percent, fill = Type)) +
    geom_col(width = 1, color = 1) +
    coord_polar(theta = "y") +
    scale_fill_brewer(palette = "Pastel1") +
    geom_label_repel(data = df2,
                     aes(y = pos, label = N),
                     size = 4.5, nudge_x = 1, show.legend = FALSE) +
    guides(fill = guide_legend(title = "Group")) +
    theme_void()

# v2
toplot <- biov2 %>%
    group_by(Type) %>%
    summarise(N = length(Type),
              percent = N/nrow(biov2))

df2 <- toplot %>% 
    mutate(csum = rev(cumsum(rev(percent))), 
           pos = percent/2 + lead(csum, 1),
           pos = if_else(is.na(pos), percent/2, pos))

ggplot(df2, aes(x = "", y = percent, fill = Type)) +
    geom_col(width = 1, color = 1) +
    coord_polar(theta = "y") +
    scale_fill_brewer(palette = "Pastel1") +
    geom_label_repel(data = df2,
                     aes(y = pos, label = N),
                     size = 4.5, nudge_x = 1, show.legend = FALSE) +
    guides(fill = guide_legend(title = "Group")) +
    theme_void()

# v3
toplot <- biov3 %>%
    group_by(Type) %>%
    summarise(N = length(Type),
              percent = N/nrow(biov3))

df2 <- toplot %>% 
    mutate(csum = rev(cumsum(rev(percent))), 
           pos = percent/2 + lead(csum, 1),
           pos = if_else(is.na(pos), percent/2, pos))

ggplot(df2, aes(x = "", y = percent, fill = Type)) +
    geom_col(width = 1, color = 1) +
    coord_polar(theta = "y") +
    scale_fill_brewer(palette = "Pastel1") +
    geom_label_repel(data = df2,
                     aes(y = pos, label = paste0(round(percent,2), "%")),
                     size = 4.5, nudge_x = 1, show.legend = FALSE) +
    guides(fill = guide_legend(title = "Group")) +
    theme_void()

# N species
tmp <- data.frame(sps = biov3$sp_name,
                  Type = biov3$Type,
                  Param = biov3$Param,
                  ECO = biov3$ECO,
                  Group = biov3$Group)

# N species by group
toplot <- 
    tmp %>%
    group_by(Type, ECO, Group) %>%
    summarise(N = length(unique(sps))) %>%
    mutate(Group = factor(Group, levels = Group[order(N)]))

ggplot(toplot, aes(x = Type, y = N, fill = Group, label = N)) +
    geom_bar(stat = "identity") +
    geom_text(size = 3, position = position_stack(vjust = 0.5))+
    facet_wrap(.~ECO, scales = "free")




```

## *Shift parameters*

```{r}

# v1
toplot <- biov1 %>%
    group_by(Type,Param) %>%
    summarise(N = length(Param))

ggplot(toplot, aes(y=N,x=Type))+
    geom_bar(stat = "identity")+
    facet_wrap(.~Param)

# v2
toplot <- biov2 %>%
    group_by(Type,Param) %>%
    summarise(N = length(Param))

ggplot(toplot, aes(y=N,x=Type))+
    geom_bar(stat = "identity")+
    facet_wrap(.~Param)

# v3
toplot <- biov3 %>%
    group_by(Param) %>%
    summarise(N = length(Param),
              percent = N/nrow(biov3))

df2 <- toplot %>% 
    mutate(csum = rev(cumsum(rev(percent))), 
           pos = percent/2 + lead(csum, 1),
           pos = if_else(is.na(pos), percent/2, pos))

ggplot(df2, aes(x = "", y = percent, fill = Param)) +
    geom_col(width = 1, color = 1) +
    coord_polar(theta = "y") +
    scale_fill_brewer(palette = "Pastel1") +
    geom_label_repel(data = df2,
                     aes(y = pos, label = paste0(round(percent,2), "%")),
                     size = 4.5, nudge_x = 1, show.legend = FALSE) +
    guides(fill = guide_legend(title = "Group")) +
    theme_void()

# N species
tmp <- data.frame(sps = biov3$sp_name,
                  Param = biov3$Param,
                  Type = biov3$Type,
                  ECO = biov3$ECO,
                  Group = biov3$Group)

# N species by group
toplot <- 
    tmp %>%
    group_by(Param, ECO, Group) %>%
    summarise(N = length(unique(sps))) %>%
    mutate(Group = factor(Group, levels = Group[order(N)]))

ggplot(toplot, aes(x = Param, y = N, fill = Group, label = N)) +
    geom_bar(stat = "identity") +
    geom_text(size = 3, position = position_stack(vjust = 0.5))+
    facet_wrap(.~ECO, scales = "free")




```

## *Shift values*

```{r fig.width=4, fig.height=5}

ggplot(biov3, aes(x=SHIFT))+
    geom_histogram()+
    facet_wrap(Param~Type, scales = "free", ncol = 2)+
    xlab("Shift (km/yr)")

ggplot(biov3, aes(x=Trend_km))+
    geom_histogram()+
    facet_wrap(Param~Type, scales = "free", ncol = 2)+
    xlab("Trend (km)")


```

## *Shift periods*

```{r fig.width=4, fig.height=5}

ggplot(biov3, aes(x=SHIFT))+
    geom_histogram()+
    facet_wrap(Param~Type, scales = "free", ncol = 2)+
    xlab("Shift (km/yr)")

ggplot(biov3, aes(x=Trend_km))+
    geom_histogram()+
    facet_wrap(Param~Type, scales = "free", ncol = 2)+
    xlab("Trend (km)")


```
### *Shifts across taxa*

```{r fig.width=5, fig.height=5}
ggplot(biov3, aes(x=SHIFT))+
    geom_histogram()+
    facet_wrap(.~Group, scales = "free")+
    xlab("Shift (km/yr)")
```

### *Across studies*
Within the same studies, how many shifts we have?

```{r fig.width=5, fig.height=5}

test <- biov3 %>%
    group_by(paste(sp_name, ID_v1, ID_v2, Type, Param, START, END)) %>%
    summarise(N=length(sp_name))

length(which(test$N>5))
length(which(test$N>6))
length(which(test$N>7))
length(which(test$N>8))

```


## *Shift duration*

```{r}

ggplot(data = biov3 %>%
           mutate(Time_periods = paste0(START,END,sep="-")))+
    geom_segment(aes(x=Time_periods, xend=Time_periods, y=START, yend=END)) +
    geom_point(aes(x=Time_periods, y=START), 
               color=rgb(0.2,0.7,0.1,0.5), size=1 ) + #green
    geom_point(aes(x=Time_periods, y=END), 
               color=rgb(0.7,0.2,0.1,0.5), size=1 ) + # red
    coord_flip()+
    theme(
        axis.text=element_text(size=6),
        legend.position = "none") +
    theme_classic()+
    ylab("Shift period (years)")

ggplot(data = biov3 %>%
           mutate(Time_periods = paste0(START,END,sep="-")))+
    geom_segment(aes(x=Time_periods, xend=Time_periods, y=START, yend=END)) +
    geom_point(aes(x=Time_periods, y=START), 
               color=rgb(0.2,0.7,0.1,0.5), size=1 ) + #green
    geom_point(aes(x=Time_periods, y=END), 
               color=rgb(0.7,0.2,0.1,0.5), size=1 ) + # red
    coord_flip()+
    theme(
        axis.text=element_text(size=6),
        legend.position = "none") +
    theme_classic()+
    ylab("Shift period (years)")+
    facet_wrap(Type~Param)

ggplot(data = biov3 %>%
           mutate(Time_periods = paste0(START,END,sep="-")))+
    geom_segment(aes(x=Time_periods, xend=Time_periods, y=START, yend=END)) +
    geom_point(aes(x=Time_periods, y=START), 
               color=rgb(0.2,0.7,0.1,0.5), size=1 ) + #green
    geom_point(aes(x=Time_periods, y=END), 
               color=rgb(0.7,0.2,0.1,0.5), size=1 ) + # red
    coord_flip()+
    theme(
        axis.text=element_text(size=6),
        legend.position = "none") +
    theme_classic()+
    ylab("Shift period (years)")+
    facet_wrap(.~Group)



```

# *Compare shift values between v1 and v2*

```{r}

# combine full and subseted data
comp <- rbind(data.frame(dataset = "v1", biov1[,c("Type","Param","kmyr","group")]),
              data.frame(dataset = "v2", biov2[,c("Type","Param","kmyr","group")]))

# compare the distribution of kmyr values
ggplot(comp, aes(y=kmyr, x=dataset, color=dataset))+
    geom_violin(alpha=.5)+
    facet_wrap(Type~Param, scales = "free")

# calculate the difference in kmyr for all combinations of Type and Param
combinations <- expand.grid(unique(comp$Type),unique(comp$Param))
names(combinations) <- c("Type", "Parameter")
combinations$mean_v1 <- NA
combinations$mean_v2 <- NA
combinations$p_value = NA

for (i in 1:nrow(combinations)) {
    tmp_data <- comp %>% 
        filter(Type == combinations[i,1] &
                   Param == combinations[i,2])
    tmp <- aov(kmyr~dataset, tmp_data)
    
    combinations$mean_v1[i] <- round(mean(tmp$model$kmyr[which(tmp$model$dataset=="v1")]),3)
    combinations$mean_v2[i] <- round(mean(tmp$model$kmyr[which(tmp$model$dataset=="v2")]),3)
    combinations$p_value[i] <- round(summary(tmp)[[1]]$`Pr(>F)`[1],3)
}

combinations

```

# *Subset*

## *Filter shifts within the temporal range of the environmental data*

```{r}

biov3_sdm <- biov3 %>%
    dplyr::filter(START >= range_env_data[1]) # shifts within the range of environmental data

# splist
splist_sdm <- splist %>% filter(scientificName %in% biov3_sdm$sp_name)

all(splist_sdm$scientificName %in% biov3_sdm$sp_name)

length(unique(splist_sdm$scientificName))

```

*Just by filtering species with shift starting after `r range_env_data[1]`, we reduce the number of species from `r length(unique(splist$scientificName))` to `r length(unique(splist_sdm$scientificName))`.*

## *Filter species with GBIF data*

```{r}

# From those species, how many we have GBIF data?

# Get N occurrences in GBIF
my_db_file <- here::here("Data/BioshiftsExposure.sqlite")
# driver
drv <- dbDriver("SQLite") 
# connect
db <- dbConnect(drv, my_db_file) 
N_OCC <- dbReadTable(conn=db, name="SpInfo")
# disconnect
DBI::dbDisconnect(db)

# remove species absent from bioshifts
N_OCC <- N_OCC %>% filter(scientificName %in% splist_sdm$scientificName)
N_OCC <- N_OCC %>% filter(!duplicated(db_code))

# check if all sps in splist are in N_OCC? 
all(splist_sdm$scientificName %in% N_OCC$scientificName)
all(N_OCC$scientificName %in% splist_sdm$scientificName)

# % bioshift with occ
bio <- data.frame(table(splist_sdm$class))
names(bio) <- c("Class","Bioshifts_Freq")

gbifocc <- data.frame(table(N_OCC$class))
names(gbifocc) <- c("Class","GBIF_Freq")

bio <- merge(bio, gbifocc)
bio$percent = paste(round((bio$GBIF_Freq/bio$Bioshifts_Freq)*100,2),"%")

melted <- reshape::melt(bio[,c(1:3)], id="Class")
melted <- merge(melted, bio[,c(1,4)])
melted$percent[which(melted$variable == "GBIF_Freq")] = NA

keep <- unique(melted$Class[which(melted$value > 20)])
toplot = melted[which(melted$Class %in% keep),]
toplot$Class <- factor(toplot$Class, levels = unique(toplot$Class[order(toplot$value)]))

ggplot(toplot, 
       aes(x = Class, y = value, fill = variable))+
    geom_bar(stat="identity",position = "identity", alpha = .5) +
    geom_text(aes(label=percent),
              stat='identity', size = 3, hjust = "inward")+
    coord_flip()+
    ylab("N species")+ylab("N species")+
    theme_classic()+
    theme(legend.position="bottom",
          legend.title = element_blank()) 

############ Filter

sel <- N_OCC %>% dplyr::filter(n_occ > 30) %>% dplyr::select(scientificName)

biov3_sdm <- biov3_sdm %>% filter(sp_name %in% unique(sel$scientificName))
splist_sdm <- splist_sdm %>% filter(scientificName %in% unique(sel$scientificName))

```

# *Stats after subsetting*

## *How many species in Bioshifts ?*

```{r}

# N species
tmp <- data.frame(sps = biov3_sdm$sp_name,
                  ECO = biov3_sdm$ECO,
                  Type = biov3_sdm$Type,
                  Group = biov3_sdm$Group)

tmp %>% summarise(N = length(unique(sps))) 

# N shifts
nrow(tmp)

# Cumulative n species by starting shift
ggplot(biov3, aes(START)) +
    stat_ecdf(geom = "step") +
    geom_vline(xintercept = range_env_data[1], color = "red")+
    theme(plot.title = element_text(hjust = 0.5))+
    ylab("Cumulative N observations")

ggplot(biov3, aes(START, colour = Type)) +
    stat_ecdf(geom = "step") +
    geom_vline(xintercept = range_env_data[1], color = "red")+
    theme(plot.title = element_text(hjust = 0.5))+
    ylab("Cumulative N observations")

# N species by group
toplot <- 
    tmp %>%
    group_by(Group) %>%
    summarise(N = length(unique(sps)))  %>%
    arrange(desc(N))

nice_table(toplot)

toplot$Group <- factor(toplot$Group, levels = unique(toplot$Group[order(toplot$N)]))
ggplot(toplot, aes(x = Group, y = N)) +
    geom_bar(stat="identity")+
    geom_text(aes(y=N, label=N, 
                  hjust = ifelse(N < max(toplot$N), -0.1, 1.1)), vjust=0.2, size=3, 
              position = position_dodge(0.9))+
    theme_classic()+
    coord_flip()

# N species by group
toplot <- 
    tmp %>%
    group_by(Group, ECO) %>%
    summarise(N = length(unique(sps)))  %>%
    arrange(desc(N))

nice_table(toplot)

toplot$Group <- factor(toplot$Group, levels = unique(toplot$Group[order(toplot$N)]))
ggplot(toplot, aes(x = Group, y = N)) +
    geom_bar(stat="identity")+
    geom_text(aes(y=N, label=N, 
                  hjust = ifelse(N < max(toplot$N), -0.1, 1.1)), vjust=0.2, size=3, 
              position = position_dodge(0.9))+
    theme_classic()+
    coord_flip()+
    facet_wrap(.~ECO)

# N species
tmp <- data.frame(sps = biov3_sdm$sp_name,
                  Type = biov3_sdm$Type,
                  Param = biov3_sdm$Param,
                  ECO = biov3_sdm$ECO,
                  Group = biov3_sdm$Group)

# N species by group
toplot <- 
    tmp %>%
    group_by(ECO, Group) %>%
    summarise(N = length(unique(sps))) %>%
    mutate(Group = factor(Group, levels = Group[order(N)]))

ggplot(toplot, aes(x = ECO, y = N, fill = Group, label = N)) +
    geom_bar(stat = "identity") +
    geom_text(size = 3, position = position_stack(vjust = 0.5))+
    facet_wrap(.~ECO, scales = "free")



```


# *Compare shift values between the subset and the complete data*

```{r}

# combine full and subseted data
comp <- rbind(data.frame(dataset = "Full", biov3),
              data.frame(dataset = "SDM", biov3_sdm))

# compare the distribution of shift values
ggplot(comp, aes(y=SHIFT, x=dataset, color=dataset))+
    geom_violin(alpha=.5)+
    facet_wrap(Type~Param, scales = "free")

# calculate the difference in shift for all combinations of Type and Param
combinations <- expand.grid(unique(biov3$Type),unique(biov3$Param))
names(combinations) <- c("Type", "Parameter")
combinations$mean_full <- NA
combinations$mean_sdm <- NA
combinations$p_value = NA

for (i in 1:nrow(combinations)) {
    tmp_data <- comp %>% 
        filter(Type == combinations[i,1] &
                   Param == combinations[i,2])
    tmp <- aov(SHIFT~dataset, tmp_data)
    
    combinations$mean_full[i] <- round(mean(tmp$model$SHIFT[which(tmp$model$dataset=="Full")]),3)
    combinations$mean_sdm[i] <- round(mean(tmp$model$SHIFT[which(tmp$model$dataset=="SDM")]),3)
    combinations$p_value[i] <- round(summary(tmp)[[1]]$`Pr(>F)`[1],3)
}


```


# *Shift values < than threshold*
How many shift values are < spatial resolution environmental data
```{r}

biov3_sdm %>%
    filter(ECO=="T") %>%
    group_by(Type, Param) %>%
    summarise('> 1km' = length(which(SHIFT>=1)),
              '< 1km' = length(which(SHIFT<1)))

biov3_sdm %>%
    filter(ECO=="A") %>%
    group_by(Type, Param) %>%
    summarise('> 1km' = length(which(SHIFT>=1)),
              '< 1km' = length(which(SHIFT<1)))

biov3_sdm %>%
    filter(ECO=="M") %>%
    group_by(Type, Param) %>%
    summarise('> 5km' = length(which(SHIFT>=5)),
              '< 5km' = length(which(SHIFT<5)))

```

