---
title: "Get species data for SDMs"
author: "Brunno F Oliveira & Bioshifts group"
date: "Last compiled on `r Sys.Date()`"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
---

# *Setup*

```{r setup, message = FALSE, warning = FALSE}

rm(list=ls())
gc()

list.of.packages <- c("dplyr", "here", "tidyr","data.table", "ggplot2")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

if(length(new.packages)) install.packages(new.packages)

sapply(list.of.packages, require, character.only = TRUE)

```

# *Load species list*

This is the list of species from Bioshifts v1 considering only:
1 - latitudinal shifts
2 - marine (except marine birds) or terrestrial shifts (removing species shifts were coded as aquatic or fresh water)
3 - species with > 30 GBIF occurrence records. For the GBIF records we used only the basis of record "HUMAN_OBSERVATION",  "OBSERVATION" or "OCCURRENCE", and occurrences recorded after 1901, occurrences flagged as has no issue.

```{r}

N_OCC <- read.csv(here("Data/n_occ.csv"))

```


# *N species*
```{r}

splist <- read.csv(here("Data/Bioshifts/splist.csv"), header = T)

# remove duplicated sp_names
splist <- splist %>%
    filter(v1==1) %>%
    dplyr::select(scientificName,kingdom,phylum,class,order,family,db,db_code) %>%
    filter(!duplicated(scientificName))

# original N species v1
nrow(splist)

# N species with occurrences in GBIF after removing the species mentioned above
N_OCC <- N_OCC %>%
    filter(n_occ > 30)
nrow(N_OCC)

all(N_OCC$scientificName %in% splist$scientificName)

# By ECO and group
## Terrestrials
N_OCC %>%
    filter(ECO == "T") %>%
    group_by(Group) %>%
    tally() %>%
    arrange(by = desc(n))

## Marine
N_OCC %>%
    filter(ECO == "M") %>%
    group_by(Group) %>%
    tally() %>%
    arrange(by = desc(n))

table(N_OCC$ECO)

####################
#### Plot
ggplot(N_OCC, aes(x = Group))+
    geom_bar()+
    theme_classic()+
    # coord_flip()+
    ylab("N species")+xlab("")+
    theme(axis.text.x = element_text(angle = 45, hjust=1))+
    facet_wrap(.~ECO, scales = "free")

```


# *N shifts*
```{r}
# Load v1
biov1 <- read.csv(here("Data/Bioshifts/biov1_fixednames.csv"), header = T)
biov1$sp_name_std_v1 <- gsub("_"," ",biov1$sp_name_std_v1)

biov1 <- biov1 %>%
    mutate(
        Type = case_when(
            Type=="HOR" ~ "LAT",
            TRUE ~ as.character(Type)),
        Species = sp_name_std_v1) %>%
    filter(Type == "LAT", # Use only latitudinal shifts
           (ECO == "T" | ECO == "M")) # Shifts Marine or Terrestrial

biov1 <- biov1 %>%
    filter(!is.na(sp_name_std_v1))

all(biov1$sp_name_std_v1 %in% splist$scientificName)


# filter species with GBIF data
sdms_sps <- list.files(here("Data/SHIFT"),recursive = TRUE)
sdms_sps
biov1 <- biov1 %>%
    filter(sp_name_std_v1 %in% N_OCC$scientificName) 
biov1 <- merge(biov1[,-which(names(biov1) %in% c("Group","ECO"))], 
               N_OCC[,c("Group","ECO","scientificName")], 
               by.x = "sp_name_std_v1", by.y = "scientificName", 
               all.x = T)

table(biov1$Type)
table(biov1$ECO)

biov1 %>%
    group_by(Group,ECO,Type,Param) %>%
    tally()

####################
#### Plot
ggplot(biov1, aes(x = Group))+
    geom_bar()+
    theme_classic()+
    # coord_flip()+
    ylab("N shifts")+xlab("")+
    theme(axis.text.x = element_text(angle = 45, hjust=1))+
    facet_wrap(.~ECO, scales = "free")
```

Select only groups with shift for all the three edges
```{r}

ggplot(biov1 %>%
           filter(ECO == "T"), 
       aes(x = Group))+
    ggtitle("Terrestrial")+
    geom_bar()+
    theme_classic()+
    # coord_flip()+
    ylab("N shifts")+xlab("")+
    theme(axis.text.x = element_text(angle = 45, hjust=1))+
    facet_grid(.~Param, scales = "free")

ggplot(biov1 %>%
           filter(ECO == "M"), 
       aes(x = Param))+
    ggtitle("Marine")+
    geom_bar()+
    theme_classic()+
    coord_flip()+
    ylab("N shifts")+xlab("")+
    theme(axis.text.x = element_text(angle = 45, hjust=1))+
    facet_wrap(.~Group, scales = "free")

```

